---
title: "02_P-matrix"
author: "Brandie QC / Julin Maloof"
date: "`r Sys.Date()`"
output: html_document
---
P-matrix from a simple simulated population

This is Brandie's script

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# IPM from Population Simulation

\*Simulations done in "Population_Sim_Trial.Rmd" file (currently using AlphaSimR)

**Goal:** Compare IPM created from just phenotypes from the simulation to an IPM created from phenotypes + genotype scores.

## Libraries

```{r}
library(tidyverse)
library(broom)
library(broom.mixed)
library(magrittr)
library(lmerTest)
conflicted::conflicts_prefer(lmerTest::lmer)
conflicted::conflicts_prefer(dplyr::filter)
conflicted::conflicts_prefer(dplyr::mutate)
```

## Load the data
```{r}
simpop <- read_csv("../output/PopSim_Phenos-Genos_v3.csv")
names(simpop)
#logit_pheno = scaled probability/phenotype
#_geno = genotype score for that trait 
#size = height calculated from simulated Weibull parameters (alpha, beta, k, and delta)
####Note: we will use the genotype score for k for this 
#elapsed_weeks = interval between weeks - should all be 1 in this data
```

### Add size next and weekly surv
```{r}
unique(simpop$established) #NA = did not germinate, 1 = established (survived first 3 weeks in field); 0 = did not establish
unique(simpop$y1surv) #NA = did not establish, 1 = survived to week 12; 0 = did not survive 
#simpop %>% select(Indiv_ID, germinated, established, y1surv, k_pheno, week, size) %>% filter(Indiv_ID<5)

simpop_timeprep <- simpop %>%
  mutate(weeks = as.numeric(week - 1), #Weeks since week 1 (equivalent to pre-transplant or initial size) 
         size = if_else(germinated==0, NA, size) #no size if no germ
         ) %>%   
  group_by(Indiv_ID) %>% 
  mutate(surv=if_else(germinated==0, NA, #surv = NA if you did not germ 
                        if_else(week==1, 1, #if you did germinate, assume you survived to week 1 (equivalent to pre-transplant size time point)
                                if_else(established==0, 0, #surv = 0 after week 1 if you did not establish 
                                        if_else(week<5, 1, #if you did establish, your surv is equal to 1 for the first 3 weeks post-transplant 
                                                y1surv) #your survival for weeks 5-12 is then equal to your y1surv
                                        )))) %>% 
  mutate(death.week=if_else(y1surv==1, NA, #if survive year 1 no death week 
                            first(week, order_by = surv)) #first week where surv = 0 
           ) %>% 
  #note we only have establishment and y1surv survival from the pop sim --> little weekly variation 
  mutate(size_next = lead(size, order_by = week), #next time point's size) 
         size = if_else(is.na(death.week), size,
                        if_else(week>death.week, NA, size))) %>%  
  ungroup() %>% 
  mutate(size_next=if_else(surv==0, NA, size_next))

simpop_timeprep %>% #look at data for a few indivs to see if above worked
  drop_na(surv) %>% 
  select(Indiv_ID, week, weeks, germinated, established, y1surv, size, size_next, surv, death.week) %>% 
  filter(Indiv_ID<25)

simpop_timeprep_surv <- simpop_timeprep %>% drop_na(surv, size) #for surv models 
simpop_timeprep_size <- simpop_timeprep %>% drop_na(size, size_next) #for size models 
```

### Scaling and transformations
```{r}
simpop_timeprep_surv %>%  #slightly skewed
  ggplot(aes(x=size)) +
  geom_histogram()

simpop_timeprep_surv_scaled <- simpop_timeprep_surv %>% 
  mutate(logSize=log(size),
         sqSize=sqrt(size)) 

simpop_timeprep_surv_scaled %>% #not better
  ggplot(aes(x=logSize)) +
  geom_histogram()

simpop_timeprep_surv_scaled %>% #not better
  ggplot(aes(x=sqSize)) +
  geom_histogram()
```

## Survival Models

### Phenos Only
```{r}
surv.models_pheno <- tribble(
  ~name,          ~f,
  "1_linear_size",                 "surv ~ size", 
  "2_linear_weeks",                "surv ~ size + weeks",
  "3a_quadratic_size",             "surv ~ size + I(size^2)", 
  "3b_quadratic_size_weeks",       "surv ~ size + I(size^2) + weeks", 
  "4a_cubic_size",                 "surv ~ size + I(size^2) + I(size^3)",
  "4b_cubic_size_weeks",           "surv ~ size + I(size^2) + I(size^3) + weeks"
)

#run the models 
surv.models_pheno <- surv.models_pheno %>%
  mutate(glm = map(f, ~ glm(as.formula(.), 
                            data = simpop_timeprep_surv, 
                            family = "binomial")), #, #run the models 
         predict = map(glm, predict, type = "response"), 
         glance = map(glm, tidy),
         AIC = map(glm, AIC),
         BIC = map(glm, BIC)) #glance at the model results

surv.models_pheno %>% select(-f, -glm, -predict) %>% unnest(BIC) %>% arrange(BIC) #look at the model fitting info 
surv.models_pheno %>% select(-f, -glm, -predict) %>% unnest(AIC) %>% arrange(AIC) #look at the model fitting info 
#3a_quadratic size best by AIC and BIC
```

```{r}
survival.model.final_pheno <- surv.models_pheno %>% filter(name == "3a_quadratic_size") %>% pull(glm) %>% magrittr::extract2(1)
summary(survival.model.final_pheno)
```

#### Predicted vs. Observed Survival
```{r}
simpop_timeprep_surv %>% 
  mutate(pred_fixef = predict(survival.model.final_pheno, newdata = ., type="response", re.form = NA)) %>% 
  ggplot(aes(x = size, y = surv)) +
  geom_point(alpha=.2) +
  geom_line(aes(y = pred_fixef), 
            color = "blue", 
            size = 1.2, alpha=0.3) 
```

### Phenos + Genos
```{r}
surv.models_genos <- tribble(
  ~name,          ~f,
  "1a_linear_size",                 "surv ~ size", 
  "1b_linear_size_genos",           "surv ~ size + establishment.logit_geno + y1surv.logit_geno", 
  "2a_linear_weeks",                "surv ~ size + weeks",
  "2b_linear_weeks_geno",           "surv ~ size + weeks + establishment.logit_geno + y1surv.logit_geno",
  "3a_quadratic_size",              "surv ~ size + I(size^2)", 
  "3b_quadratic_size_weeks",        "surv ~ size + I(size^2) + weeks", 
  "3c_quadratic_size_weeks_geno",   "surv ~ size + I(size^2) + weeks + establishment.logit_geno + y1surv.logit_geno",
  "4a_cubic_size",                  "surv ~ size + I(size^2) + I(size^3)",
  "4b_cubic_size_weeks",            "surv ~ size + I(size^2) + I(size^3) + weeks",
  "4c_cubic_size_weeks_geno",       "surv ~ size + I(size^2) + I(size^3) + weeks + establishment.logit_geno + y1surv.logit_geno"
)

#run the models 
surv.models_genos <- surv.models_genos %>%
  mutate(glm = map(f, ~ glm(as.formula(.), 
                            data = simpop_timeprep_surv, 
                            family = "binomial")), #, #run the models 
         predict = map(glm, predict, type = "response"), 
         glance = map(glm, tidy),
         AIC = map(glm, AIC),
         BIC = map(glm, BIC)) #glance at the model results

surv.models_genos %>% select(-f, -glm, -predict) %>% unnest(BIC) %>% arrange(BIC) #look at the model fitting info 
surv.models_genos %>% select(-f, -glm, -predict) %>% unnest(AIC) %>% arrange(AIC) #look at the model fitting info 
#3a_quadratic_size best by BIC
#3c_quadratic_size_weeks_geno best by AIC 
```

```{r}
survival.model.final_genos <- surv.models_genos %>% filter(name == "3c_quadratic_size_weeks_geno") %>% pull(glm) %>% magrittr::extract2(1)
summary(survival.model.final_genos)
```

#### Predicted vs. Observed Survival
```{r}
simpop_timeprep_surv %>% 
  mutate(pred_fixef = predict(survival.model.final_genos, newdata = ., type="response", re.form = NA)) %>% 
  ggplot(aes(x = size, y = surv)) +
  geom_point(alpha=.2) +
  geom_line(aes(y = pred_fixef), 
            color = "blue", 
            size = 1.2, alpha=0.3) 
```

## Growth Models

### Observed Patterns
```{r}
simpop_timeprep_size %>% 
  ggplot(aes(x=size, y=size_next)) +
  geom_point()  +
  geom_abline() +
  geom_smooth(method = "lm")
```

### Phenos Only
```{r}
size.models_pheno <- tribble(
  ~name,          ~f,
  "1_linear_size",                 "size_next ~ size", 
  "2_linear_weeks",                "size_next ~ size + weeks",
  "3a_quadratic_size",             "size_next ~ size + I(size^2)", 
  "3b_quadratic_size_weeks",       "size_next ~ size + I(size^2) + weeks", 
  "4a_cubic_size",                 "size_next ~ size + I(size^2) + I(size^3)",
  "4b_cubic_size_weeks",           "size_next ~ size + I(size^2) + I(size^3) + weeks"
)

#run the models 
size.models_pheno <- size.models_pheno %>%
  mutate(lm = map(f, ~ lm(as.formula(.), 
                            data = simpop_timeprep_size)), #, #run the models 
         predict = map(lm, predict), 
         glance = map(lm, glance)) #glance at the model results

size.models_pheno %>% select(-f, -lm, -predict) %>% unnest(glance) %>% arrange(BIC) #look at the model fitting info 
size.models_pheno %>% select(-f, -lm, -predict) %>% unnest(glance) %>% arrange(AIC) #look at the model fitting info 
#3b_quadratic_size_weeks best by AIC and BIC
```

```{r}
size.model.final_pheno <- size.models_pheno %>% filter(name == "3b_quadratic_size_weeks") %>% pull(lm) %>% magrittr::extract2(1)
summary(size.model.final_pheno)
```

#### Predicted vs. Observed Growth
```{r}
simpop_timeprep_size %>% 
  cbind(predicted={size.models_pheno %>% filter(name=="3b_quadratic_size_weeks") %>% pull(predict) %>% unlist()}) %>%
  ggplot(aes(x=size_next, y = predicted)) +
  geom_point() +
  geom_abline(color="skyblue2")
```

### Phenos + Genos
```{r}
size.models_genos <- tribble(
  ~name,          ~f,
  "1a_linear_size",                 "size_next ~ size", 
  "1b_linear_size_geno",            "size_next ~ size + k_geno", 
  "2a_linear_weeks",                "size_next ~ size + weeks",
  "2b_linear_weeks_geno",           "size_next ~ size + weeks + k_geno",
  "3a_quadratic_size",              "size_next ~ size + I(size^2)", 
  "3b_quadratic_size_weeks",        "size_next ~ size + I(size^2) + weeks", 
  "3c_quadratic_size_weeks_geno",   "size_next ~ size + I(size^2) + weeks + k_geno",
  "4a_cubic_size",                  "size_next ~ size + I(size^2) + I(size^3)",
  "4b_cubic_size_weeks",            "size_next ~ size + I(size^2) + I(size^3) + weeks",
  "4c_cubic_size_weeks_geno",       "size_next ~ size + I(size^2) + I(size^3) + weeks + k_geno"
)

#run the models 
size.models_genos <- size.models_genos %>%
  mutate(lm = map(f, ~ lm(as.formula(.), 
                            data = simpop_timeprep_size)), #, #run the models 
         predict = map(lm, predict), 
         glance = map(lm, glance)) #glance at the model results

size.models_genos %>% select(-f, -lm, -predict) %>% unnest(glance) %>% arrange(BIC) #look at the model fitting info 
size.models_genos %>% select(-f, -lm, -predict) %>% unnest(glance) %>% arrange(AIC) #look at the model fitting info 
#3c_quadratic_size_weeks_geno best by AIC and BIC
```

```{r}
size.model.final_genos <- size.models_genos %>% filter(name == "3c_quadratic_size_weeks_geno") %>% pull(lm) %>% magrittr::extract2(1)
summary(size.model.final_genos)
```

#### Predicted vs. Observed Growth
```{r}
simpop_timeprep_size %>% 
  cbind(predicted={size.models_genos %>% filter(name=="3c_quadratic_size_weeks_geno") %>% pull(predict) %>% unlist()}) %>%
  ggplot(aes(x=size_next, y = predicted)) +
  geom_point() +
  geom_abline(color="skyblue2")
```

## P Matrix

### Make a dataframe to store the parameters
```{r}
params=data.frame(
  surv.int=NA, # Intercept from logistic regression of survival
  surv.slope1=NA, # Slope from logistic regression of survival
  surv.slope2=NA, # Quadratic slope from logistic regression of survival
  growth.int=NA, # Intercept from linear regression of growth
  growth.slope1=NA, # Slope from linear regression of growth
  growth.slope2=NA, # Quadratic slope from linear regression of growth
  growth.sd=NA # Residual sd from the linear regression of growth
)

params_phenos <- params
params_genos <- params
```

### Use broom:tidy to create a df with the coef from each model
```{r}
surv.coefs_pheno <- surv.models_pheno %>% 
  filter(name == "3a_quadratic_size") %>%
  mutate(coefs = map(glm, tidy)) %>% 
  select(name, f, coefs) %>% 
  unnest(coefs) 

params_phenos$surv.int <- surv.coefs_pheno %>% filter(term == "(Intercept)") %>% pull(estimate)
params_phenos$surv.slope1 <- surv.coefs_pheno %>% filter(term == "size") %>% pull(estimate)
params_phenos$surv.slope2 <- surv.coefs_pheno %>% filter(term == "I(size^2)") %>% pull(estimate)
```

```{r}
surv.coefs_genos <- surv.models_genos %>% 
  filter(name == "3c_quadratic_size_weeks_geno") %>%
  mutate(coefs = map(glm, tidy)) %>% 
  select(name, f, coefs) %>% 
  unnest(coefs) 

params_genos$surv.int <- surv.coefs_genos %>% filter(term == "(Intercept)") %>% pull(estimate)
params_genos$surv.slope1 <- surv.coefs_genos %>% filter(term == "size") %>% pull(estimate)
params_genos$surv.slope2 <- surv.coefs_genos %>% filter(term == "I(size^2)") %>% pull(estimate)
```

```{r}
growth.coefs_pheno <- size.models_pheno %>% 
  filter(name == "3b_quadratic_size_weeks") %>% 
  mutate(coefs = map(lm, tidy)) %>% 
  select(name, f, coefs) %>% 
  unnest(coefs) 

params_phenos$growth.int <- growth.coefs_pheno %>% filter(term == "(Intercept)") %>% pull(estimate)
params_phenos$growth.slope1 <- growth.coefs_pheno %>% filter(term == "size") %>% pull(estimate)
params_phenos$growth.slope2 <- growth.coefs_pheno %>% filter(term == "I(size^2)") %>% pull(estimate)

#Pull sigma which is the _modeled_ standard deviation of the residuals.  Merow uses observed sd of residuals.  
params_phenos$growth.sd <- size.models_pheno %>% filter(name == "3b_quadratic_size_weeks") %>% unnest(glance) %>% pull(sigma)
```

```{r}
growth.coefs_genos <- size.models_genos %>% 
  filter(name == "3c_quadratic_size_weeks_geno") %>% 
  mutate(coefs = map(lm, tidy)) %>% 
  select(name, f, coefs) %>% 
  unnest(coefs) 

params_genos$growth.int <- growth.coefs_genos %>% filter(term == "(Intercept)") %>% pull(estimate)
params_genos$growth.slope1 <- growth.coefs_genos %>% filter(term == "size") %>% pull(estimate)
params_genos$growth.slope2 <- growth.coefs_genos %>% filter(term == "I(size^2)") %>% pull(estimate)

#Pull sigma which is the _modeled_ standard deviation of the residuals.  Merow uses observed sd of residuals.  
params_genos$growth.sd <- size.models_genos %>% filter(name == "3c_quadratic_size_weeks_geno") %>% unnest(glance) %>% pull(sigma)
```

### Define the functions
```{r}
# 1. survival probability function
##This is inverse logit.  Provides survival probability based on size.
s.x_phenos=function(x,params_phenos) {
  u=exp(params_phenos$surv.int + params_phenos$surv.slope1*x + params_phenos$surv.slope2*x^2)
  return(u/(1+u)) 
}

# 2. growth function
## Return a probability distribution of new sizes at t+1 (xp) at a given size x.  
g.yx_phenos=function(xp,x,params_phenos) {
  dnorm(xp,mean=params_phenos$growth.int + params_phenos$growth.slope1*x + params_phenos$growth.slope2*x^2, sd=params_phenos$growth.sd)
}
```

```{r}
# 1. survival probability function
##This is inverse logit.  Provides survival probability based on size.
s.x_genos=function(x,params_genos) {
  u=exp(params_genos$surv.int + params_genos$surv.slope1*x + params_genos$surv.slope2*x^2)
  return(u/(1+u)) 
}

# 2. growth function
## Return a probability distribution of new sizes at t+1 (xp) at a given size x.  
g.yx_genos=function(xp,x,params_genos) {
  dnorm(xp,mean=params_genos$growth.int + params_genos$growth.slope1*x + params_genos$growth.slope2*x^2, sd=params_genos$growth.sd)
}
```

### Define the structure of the IPM
```{r}
# the sizes we are integrating over
minSize<-min(simpop_timeprep$size,na.rm=T) 
maxSize<-max(simpop_timeprep$size,na.rm=T) 

n=100 # dimensions of the matrix 

b=minSize+c(0:n)*(maxSize-minSize)/n # boundary points
y=0.5*(b[1:n]+b[2:(n+1)]) # mesh points
h=y[2]-y[1] # step size
```

### Make the matrices (G, S, and P)
```{r}
G_phenos=h*outer(y,y,g.yx_phenos,params=params_phenos) # growth matrix

S_phenos=s.x_phenos(y,params=params_phenos) # survival at each size midpoint

P_phenos=G_phenos # placeholder; redefine P on the next line
for(i in 1:n) P_phenos[,i]=G_phenos[,i]*S_phenos[i] # growth/survival matrix
```

```{r}
G_genos=h*outer(y,y,g.yx_genos,params=params_genos) # growth matrix

S_genos=s.x_genos(y,params=params_genos) # survival at each size midpoint

P_genos=G_genos # placeholder; redefine P on the next line
for(i in 1:n) P_genos[,i]=G_genos[,i]*S_genos[i] # growth/survival matrix
```

### Plot the matrix
```{r}
P_phenos %>% as_tibble() %>%
  set_colnames(y) %>% #column names = each size mesh point 
  mutate(size.t1=y) %>%
  pivot_longer(-size.t1, names_to = "size.t", names_transform = as.numeric) %>%
  ggplot(aes(x=size.t, y = size.t1)) +
  geom_raster(aes(fill = value)) + #basic contour with fill determined by the growth*surv value 
  geom_contour(aes(z = value),lwd=.25) + #adds contour lines 
  geom_abline(intercept=0, slope = 1, color="gray90", lty=5) + #add 1:1 line 
  scale_fill_viridis_c(option = "plasma") + #change contour colors 
  labs(x = "Size (t)", y = "Size (t + 1)", title = "P Matrix: Phenos Only") +
  coord_equal() + #make it a square plot 
  theme_bw()
```
```{r}
P_genos %>% as_tibble() %>%
  set_colnames(y) %>% #column names = each size mesh point 
  mutate(size.t1=y) %>%
  pivot_longer(-size.t1, names_to = "size.t", names_transform = as.numeric) %>%
  ggplot(aes(x=size.t, y = size.t1)) +
  geom_raster(aes(fill = value)) + #basic contour with fill determined by the growth*surv value 
  geom_contour(aes(z = value),lwd=.25) + #adds contour lines 
  geom_abline(intercept=0, slope = 1, color="gray90", lty=5) + #add 1:1 line 
  scale_fill_viridis_c(option = "plasma") + #change contour colors 
  labs(x = "Size (t)", y = "Size (t + 1)", title = "P Matrix: Phenos + Genos") +
  coord_equal() + #make it a square plot 
  theme_bw()
```
### Check for eviction
```{r}
plot(y,s.x_phenos(y,params_phenos), #fitted survival model 
     xlab="Size",type="l",
        ylab="Survival Probability",lwd=12)
     points(y,apply(P_phenos,2,sum),col="red",lwd=3,cex=.1,pch=19) # column sums 
#some eviction at larger sizes
```
```{r}
plot(y,s.x_genos(y,params_genos), #fitted survival model 
     xlab="Size",type="l",
        ylab="Survival Probability",lwd=12)
     points(y,apply(P_genos,2,sum),col="red",lwd=3,cex=.1,pch=19) # column sums 
#some eviction at larger sizes
```
#### Eviction correction using a constant 
```{r}
Pc_phenos=G_phenos # placeholder; redefine P on the next line
for(i in 1:n) Pc_phenos[,i]=G_phenos[,i]*S_phenos[i] # growth/survival matrix
nvals <- colSums(Pc_phenos, na.rm = TRUE)
Pc_phenos <- t((t(Pc_phenos)/nvals) * s.x_phenos(y, params=params_phenos))

plot(y,s.x_phenos(y,params_phenos),xlab="Size",type="l",
        ylab="Survival Probability",lwd=12)
     points(y,apply(Pc_phenos,2,sum),col="red",lwd=3,cex=.1,pch=19) # solution worked
```

```{r}
Pc_genos=G_genos # placeholder; redefine P on the next line
for(i in 1:n) Pc_genos[,i]=G_genos[,i]*S_genos[i] # growth/survival matrix
nvals <- colSums(Pc_genos, na.rm = TRUE)
Pc_genos <- t((t(Pc_genos)/nvals) * s.x_genos(y, params=params_genos))

plot(y,s.x_genos(y,params_genos),xlab="Size",type="l",
        ylab="Survival Probability",lwd=12)
     points(y,apply(Pc_genos,2,sum),col="red",lwd=3,cex=.1,pch=19) # solution worked
```

#### Plot corrected P matrices 
```{r}
Pc_phenos %>% as_tibble() %>%
  set_colnames(y) %>% #column names = each size mesh point 
  mutate(size.t1=y) %>%
  pivot_longer(-size.t1, names_to = "size.t", names_transform = as.numeric) %>%
  ggplot(aes(x=size.t, y = size.t1)) +
  geom_raster(aes(fill = value)) + #basic contour with fill determined by the growth*surv value 
  geom_contour(aes(z = value),lwd=.25) + #adds contour lines 
  geom_abline(intercept=0, slope = 1, color="gray90", lty=5) + #add 1:1 line 
  scale_fill_viridis_c(option = "plasma") + #change contour colors 
  labs(x = "Size (t)", y = "Size (t + 1)", title = "P Matrix: Phenos Only") +
  coord_equal() + #make it a square plot 
  theme_bw()
```

```{r}
Pc_genos %>% as_tibble() %>%
  set_colnames(y) %>% #column names = each size mesh point 
  mutate(size.t1=y) %>%
  pivot_longer(-size.t1, names_to = "size.t", names_transform = as.numeric) %>%
  ggplot(aes(x=size.t, y = size.t1)) +
  geom_raster(aes(fill = value)) + #basic contour with fill determined by the growth*surv value 
  geom_contour(aes(z = value),lwd=.25) + #adds contour lines 
  geom_abline(intercept=0, slope = 1, color="gray90", lty=5) + #add 1:1 line 
  scale_fill_viridis_c(option = "plasma") + #change contour colors 
  labs(x = "Size (t)", y = "Size (t + 1)", title = "P Matrix: Phenos + Genos") +
  coord_equal() + #make it a square plot 
  theme_bw()
```