---
title: "Merow Appendix C"
author: "Julin Maloof"
date: "2024-12-04"
output:
  html_document:
    keep_markdown: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(broom)
#library(patchwork)
#library(ggbeeswarm)
#library(magrittr)
#library(nlme)
library(IPMpack)
#conflicted::conflict_prefer("select", "dplyr")
```

```{r}
data("dataIPMpackHypericumCov")

d <- dataIPMpackHypericumCov
head(d)
```

```{r}
skimr::skim(d)
```
```{r}
d1 <- subset(d,is.na(d$size)==FALSE | d$ontogenyNext==1)
d1 <- subset(d1,d1$year==1994 | d1$year==1996)
```

some additional data:
```{r}
fec2 <- 13.78 # number of seed per fruit
fec3 <- 0.001336 # probability of germination the year that the seed was prodiced
fec4 <- 0.14 # probability of seedling survival from germination to time of next census
goSB <- 0.08234528 # probability of a seed entering the seed bank
staySB <- 0.671 # probability of a seed staying in the seed bank
```

keep columns that we are going to use
```{r}
d1 <- d1 |>
  dplyr::select(year, size, surv, sizeNext, fec0, fec1)
```

Add a variable to indicate the number of individuals changing size.  This will be be used later.  Set up stage and stage next

```{r}
d1$number <- 1
d1$stageNext <- d1$stage <- "continuous"
d1$stage[is.na(d1$size)] <- NA
d1$stageNext[d1$surv==0] <- "dead"
```

Set up a seedbank data frame
```{r}
seedbank <- data.frame(year="All",size=NA,surv=1,sizeNext=NA,fec0=NA,fec1=NA,
                       stage=c("seedbank","seedbank","continuous"),
                       stageNext=c("seedbank","continuous","seedbank"),
                       number=c(staySB,(1-staySB)*fec3*fec4,1))
seedbank
```
combine data frames
```{r}
d1 <- rbind(d1, seedbank)
```

## Analysis 

### 94 data

```{r}
d94 <- subset(d1, d1$year == "1994" | d1$year == "All")
minSize<-min(d1$size,na.rm=T)
maxSize<-max(d1$size,na.rm=T)
x<-seq(from=minSize,to=maxSize,length=1001)
x0<-data.frame(size=x,size2=x^2,size3=x^3) # for later use
```

#### Survival and Growth Kernel

** Merow Version **

Survival

```{r}
survModelComp(dataf = d94[!is.na(d$size),],
              expVars = c(surv~1, surv~size, surv~size + size2,
                          surv~size + size2 + size3),
              makePlot = TRUE, legendPos = "bottomleft", mainTitle = "Survival")
```
choose quadratic
```{r}
so94 <- makeSurvObj(d94, surv~size+I(size^2))
```


** JM Version **

```{r}
newdata = tibble(size=seq(min(d94$size, na.rm = TRUE), max(d94$size, na.rm = TRUE), length.out=101))

surv.models <- tibble(name=c("0_intercept", "1_linear", "2_quadratic", "3_cubic"),
                      f = c("surv ~ 1", 
                            "surv ~ size", 
                            "surv ~ size + I(size^2)", 
                            "surv ~ size + I(size^2) + I(size^3)"))

surv.models <- surv.models %>%
  mutate(glm = map(f, ~ glm(as.formula(.), data = d94, family = "binomial") ),
         predict = map(glm, ~ cbind(newdata, surv=predict(., newdata, type = "response"))),
         glance = map(glm, glance))

surv.models %>% unnest(glance)
```
AIC is **VERY** similar.  Merow chose quadratic.

plot the fits
```{r}
surv.predictions <- surv.models %>% dplyr::select(name, predict) %>% unnest(predict)

d94 %>% filter(!is.na(surv)) %>%
  ggplot(aes(x=size, y =surv)) +
  geom_point() +
  geom_line(aes(color=name), data = surv.predictions) +
  scale_color_viridis_d(end = 0.9)
```


Growth

** Merrow **
```{r}
growthModelComp(dataf = d94, expVars = c(sizeNext~1, sizeNext~size,
                                         sizeNext~size + size2, sizeNext~size + size2 + size3), makePlot = TRUE,
                legendPos = "bottomright", mainTitle = "Growth")
abline(a = 0, b = 1, lty= 2, col = "gray", lwd=2)
```

Choose linear
```{r}
go94 <- makeGrowthObj(d94, sizeNext~size)
```

** Julin **

** JM Version **

```{r}
growth.models <- tibble(name=c("0_intercept", "1_linear", "2_quadratic", "3_cubic"),
                      f = c("sizeNext ~ 1", 
                            "sizeNext ~ size", 
                            "sizeNext ~ size + I(size^2)", 
                            "sizeNext ~ size + I(size^2) + I(size^3)"))

growth.models <- growth.models %>%
  mutate(glm = map(f, ~ glm(as.formula(.), data = d94) ),
         predict = map(glm, ~ cbind(newdata, sizeNext=predict(., newdata, type = "response"))),
         glance = map(glm, glance))

growth.models %>% unnest(glance)
```
AIC is AIC is lowest for linear.

plot the fits
```{r}
growth.predictions <- growth.models %>% dplyr::select(name, predict) %>% unnest(predict)

d94 %>% filter(!is.na(sizeNext) & !is.na(size)) %>%
  ggplot(aes(x=size, y =sizeNext)) +
  geom_point() +
  geom_abline(slope = 1, intercept = 0, lty = 5, color = "grey") + 
  geom_line(aes(color=name), data = growth.predictions) +
  scale_color_viridis_d(end = 0.9)
```


Make the P matrix and plot it

** Merow **
```{r}
Pmatrix94 <- makeIPMPmatrix(survObj = so94, growObj = go94,
                            minSize = minSize, maxSize = maxSize,
                            nBigMatrix = 80, correction = "constant")
require(fields)
image.plot(Pmatrix94@meshpoints,
           Pmatrix94@meshpoints,
           t(Pmatrix94),
           main = "Pmatrix: survival and growth",
           xlab = "Size at t",
           ylab = "Size at t+1")
abline(a = 0, b = 1, lty= 2, col = "white", lwd=2)
```
evaluate P matrix
```{r, eval=FALSE}
diagnosticsPmatrix(Pmatrix94, growObj = go94, survObj = so94,
                   correction = "constant")
```

** Julin **

paramater data frame

```{r}
params=data.frame(
  surv.int=NA, # Intercept from logistic regression of survival
  surv.slope1=NA, # Slope from logistic regression of survival
  surv.slope2=NA, # Quadratic slope from logistic regression of survival
  growth.int=NA, # Intercept from linear regression of growth
  growth.slope=NA, # Slope from linear regression of growth
  growth.sd=NA # Residual sd from the linear regression of growth
)
```

```{r}
# would be better to do this with broom::tidy
surv.final <- surv.models %>% filter(name=="2_quadratic") %>% pull(glm) %>% magrittr::extract2(1)
coef(surv.final)
```


evaluate P matrix
```{r, eval=FALSE}
diagnosticsPmatrix(Pmatrix94, growObj = go94, survObj = so94,
                   correction = "constant")
```


#### Fecundity Kernel

Compare different models for flowering probability:
getting an error
```{r, eval=FALSE}
fec0.0_94 <- makeFecObj(d94, Formula = fec0~1, Family = "binomial")
fec0.1_94 <- makeFecObj(d94, Formula = fec0~size, Family = "binomial")
fec0.2_94 <- makeFecObj(d94, Formula = fec0~size+size2, Family = "binomial")
fec0.3_94 <- makeFecObj(d94, Formula = fec0~size+size2+size3, Family = "binomial")
```

```{r}
newdata = tibble(size=seq(min(d94$size, na.rm = TRUE), max(d94$size, na.rm = TRUE), length.out=101))

fec0.models <- tibble(name=c("0_intercept", "1_linear", "2_quadratic", "3_cubic"),
                      f = c("fec0 ~ 1", 
                            "fec0 ~ size", 
                            "fec0 ~ size + I(size^2)", 
                            "fec0 ~ size + I(size^2) + I(size^3)"))

fec0.models <- fec0.models %>%
  mutate(glm = map(f, ~ glm(as.formula(.), data = d94, family = "binomial") ),
         predict = map(glm, ~ cbind(newdata, fec0=predict(., newdata, type = "response"))),
         glance = map(glm, glance))

fec0.models %>% unnest(glance)

# AIC lowest in cubic, although not by a ton.
```

plot predictions:
```{r}
fec0.predictions <- fec0.models %>% dplyr::select(name, predict) %>% unnest(predict)

d94 %>% filter(!is.na(fec0)) %>%
  ggplot(aes(x=size, y =fec0)) +
  geom_point() +
  geom_line(aes(color=name), data = fec0.predictions) +
  scale_color_viridis_d(end = 0.9)
```


Compare different models for number of fruits
getting an error
```{r, eval=FALSE}
fec1.0_94 <- makeFecObj(d94, Formula = fec1~1, Family = "poisson")
fec1.1_94 <- makeFecObj(d94, Formula = fec1~size, Family = "poisson")
fec1.2_94 <- makeFecObj(d94, Formula = fec1~size+size2, Family = "poisson")
fec1.3_94 <- makeFecObj(d94, Formula = fec1~size+size2+size3, Family = "poisson")
```

```{r}
fec1.models <- tibble(name=c("0_intercept", "1_linear", "2_quadratic", "3_cubic"),
                      f = c("fec1 ~ 1", 
                            "fec1 ~ size", 
                            "fec1 ~ size + I(size^2)", 
                            "fec1 ~ size + I(size^2) + I(size^3)"))

fec1.models <- fec1.models %>%
  mutate(glm = map(f, ~ glm(as.formula(.), data = d94, family = "poisson") ),
         predict = map(glm, ~ cbind(newdata, fec1=predict(., newdata, type = "response"))),
         glance = map(glm, glance))

fec1.models %>% unnest(glance)

# AIC clearly lowest in cubic
```

plot predictions:
```{r}
fec1.predictions <- fec1.models %>% dplyr::select(name, predict) %>% unnest(predict)

d94 %>% filter(!is.na(fec1)) %>%
  ggplot(aes(x=size, y =fec1)) +
  geom_point() +
  geom_line(aes(color=name), data = fec1.predictions) +
  scale_color_viridis_d(end=.9)
```

put it together:
```{r}
fo94 <- makeFecObj(d94, Formula=c(fec0~size+size2, fec1~size+size2+size3),
                   Family=c("binomial", "poisson"),
                   Transform=c("none", -1),
                   meanOffspringSize=mean(d94[is.na(d1$size)==TRUE &
                                                is.na(d94$sizeNext)==FALSE,"sizeNext"]),
                   sdOffspringSize=sd(d94[is.na(d1$size)==TRUE &
                                            is.na(d94$sizeNext)==FALSE,"sizeNext"]),
                   fecConstants=data.frame(fec2=fec2,fec3=fec3,fec4=fec4),
                   offspringSplitter=data.frame(seedbank=goSB,
                                                continuous=(1-goSB)),
                   vitalRatesPerOffspringType=data.frame(
                     seedbank=c(1,1,1,0,0),
                     continuous=c(1,1,1,1,1),
                     row.names=c("fec0","fec1",
                                 "fec2","fec3","fec4")))
```

