---
title: "Merow Appendix C"
author: "Julin Maloof"
date: "2024-12-04"
output:
  html_document:
    keep_md: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(broom)
#library(patchwork)
#library(ggbeeswarm)
library(magrittr)
#library(nlme)
library(IPMpack)
source("makeDiscreteTrans.R", local=as.environment("package:IPMpack"))
#conflicted::conflict_prefer("select", "dplyr")
```

```{r}
data("dataIPMpackHypericumCov")

d <- dataIPMpackHypericumCov
head(d)
```

```{r}
skimr::skim(d)
```
```{r}
d1 <- subset(d,is.na(d$size)==FALSE | d$ontogenyNext==1)
d1 <- subset(d1,d1$year==1994 | d1$year==1996)
```

some additional data:
```{r}
fec2 <- 13.78 # number of seed per fruit
fec3 <- 0.001336 # probability of germination the year that the seed was prodiced
fec4 <- 0.14 # probability of seedling survival from germination to time of next census
goSB <- 0.08234528 # probability of a seed entering the seed bank
staySB <- 0.671 # probability of a seed staying in the seed bank
```

keep columns that we are going to use
```{r}
d1 <- d1 |>
  dplyr::select(year, size, surv, sizeNext, fec0, fec1)
```

Add a variable to indicate the number of individuals changing size.  This will be be used later.  Set up stage and stage next

```{r}
d1$number <- 1
d1$stageNext <- d1$stage <- "continuous"
d1$stage[is.na(d1$size)] <- NA
d1$stageNext[d1$surv==0] <- "dead"
```

Set up a seedbank data frame
```{r}
seedbank <- data.frame(year="All",size=NA,surv=1,sizeNext=NA,fec0=NA,fec1=NA,
                       stage=c("seedbank","seedbank","continuous"),
                       stageNext=c("seedbank","continuous","seedbank"),
                       number=c(staySB,(1-staySB)*fec3*fec4,1))
seedbank
```
combine data frames
```{r}
d1 <- rbind(d1, seedbank)
```

## Analysis 

### 94 data

```{r}
d94 <- subset(d1, d1$year == "1994" | d1$year == "All")
minSize<-min(d1$size,na.rm=T)
maxSize<-max(d1$size,na.rm=T)
x<-seq(from=minSize,to=maxSize,length=1001)
x0<-data.frame(size=x,size2=x^2,size3=x^3) # for later use
```

#### Survival and Growth Kernel

** Merow Version **

Survival

```{r}
survModelComp(dataf = d94[!is.na(d$size),],
              expVars = c(surv~1, surv~size, surv~size + size2,
                          surv~size + size2 + size3),
              makePlot = TRUE, legendPos = "bottomleft", mainTitle = "Survival")
```
choose quadratic
```{r}
so94 <- makeSurvObj(d94, surv~size+I(size^2))
```


** JM Version **

```{r}
newdata = tibble(size=seq(min(d94$size, na.rm = TRUE), max(d94$size, na.rm = TRUE), length.out=101))

surv.models <- tibble(name=c("0_intercept", "1_linear", "2_quadratic", "3_cubic"),
                      f = c("surv ~ 1", 
                            "surv ~ size", 
                            "surv ~ size + I(size^2)", 
                            "surv ~ size + I(size^2) + I(size^3)"))

surv.models <- surv.models %>%
  mutate(glm = map(f, ~ glm(as.formula(.), data = d94, family = "binomial") ),
         predict = map(glm, ~ cbind(newdata, surv=predict(., newdata, type = "response"))),
         glance = map(glm, glance))

surv.models %>% unnest(glance)
```
AIC is **VERY** similar.  Merow chose quadratic.

```{r}
surv.models$predict[[1]]
```


plot the fits
```{r}
surv.predictions <- surv.models %>% dplyr::select(name, predict) %>% unnest(predict)

d94 %>% filter(!is.na(surv)) %>%
  ggplot(aes(x=size, y =surv)) +
  geom_point() +
  geom_smooth() + 
  geom_line(aes(color=name), data = surv.predictions) +
  scale_color_viridis_d(end = 0.9)
```


Growth

** Merrow **
```{r}
growthModelComp(dataf = d94, expVars = c(sizeNext~1, sizeNext~size,
                                         sizeNext~size + size2, sizeNext~size + size2 + size3), makePlot = TRUE,
                legendPos = "bottomright", mainTitle = "Growth")
abline(a = 0, b = 1, lty= 2, col = "gray", lwd=2)
```

Choose linear
```{r}
go94 <- makeGrowthObj(d94, sizeNext~size)
```

** Julin **

```{r}
growth.models <- tibble(name=c("0_intercept", "1_linear", "2_quadratic", "3_cubic"),
                        f = c("sizeNext ~ 1", 
                              "sizeNext ~ size", 
                              "sizeNext ~ size + I(size^2)", 
                              "sizeNext ~ size + I(size^2) + I(size^3)"))

growth.models <- growth.models %>%
  mutate(lm = map(f, ~ lm(as.formula(.), data = d94) ),
         predict = map(lm, ~ cbind(newdata, sizeNext=predict(., newdata))),
         glance = map(lm, glance))

growth.models %>% unnest(glance)
```
AIC is AIC is lowest for linear.

plot the fits
```{r}
growth.predictions <- growth.models %>% dplyr::select(name, predict) %>% unnest(predict)

d94 %>% filter(!is.na(sizeNext) & !is.na(size)) %>%
  ggplot(aes(x=size, y =sizeNext)) +
  geom_point() +
  geom_abline(slope = 1, intercept = 0, lty = 5, color = "grey") + 
  geom_line(aes(color=name), data = growth.predictions) +
  scale_color_viridis_d(end = 0.9)
```


Make the P matrix and plot it

** Merow **
```{r, fig.asp=1}
Pmatrix94 <- makeIPMPmatrix(survObj = so94, growObj = go94,
                            minSize = minSize, maxSize = maxSize,
                            nBigMatrix = 80, correction = "constant")
require(fields)
image.plot(Pmatrix94@meshpoints,
           Pmatrix94@meshpoints,
           t(Pmatrix94),
           main = "Pmatrix: survival and growth",
           xlab = "Size at t",
           ylab = "Size at t+1")
abline(a = 0, b = 1, lty= 2, col = "white", lwd=2)
```
evaluate P matrix
```{r, eval=FALSE}
diagnosticsPmatrix(Pmatrix94, growObj = go94, survObj = so94,
                   correction = "constant")
```

** Julin **

parameter data frame

```{r}
params=data.frame(
  surv.int=NA, # Intercept from logistic regression of survival
  surv.slope1=NA, # Slope from logistic regression of survival
  surv.slope2=NA, # Quadratic slope from logistic regression of survival
  growth.int=NA, # Intercept from linear regression of growth
  growth.slope=NA, # Slope from linear regression of growth
  growth.sd=NA # Residual sd from the linear regression of growth
)
```

Get the survival coefficients.  Using the quadratic model.
```{r}
# I am using broom:tidy to create a df with the coefficients.
surv.coefs <- surv.models %>% 
  filter(name == "2_quadratic") %>%
  mutate(coefs = map(glm, tidy)) %>% 
  dplyr::select(name, f, coefs) %>% 
  unnest(coefs) 

params$surv.int <- surv.coefs %>% filter(term == "(Intercept)") %>% pull(estimate)
params$surv.slope1 <- surv.coefs %>% filter(term == "size") %>% pull(estimate)
params$surv.slope2 <- surv.coefs %>% filter(term == "I(size^2)") %>% pull(estimate)
```

Get the growth coefficients.  Using the linear model.
```{r}
growth.coefs <- growth.models %>% 
  filter(name == "1_linear") %>%
  mutate(coefs = map(lm, tidy)) %>% 
  dplyr::select(name, f, coefs) %>% 
  unnest(coefs)

params$growth.int <- growth.coefs %>% filter(term=="(Intercept)") %>% pull(estimate)
params$growth.slope <- growth.coefs %>% filter(term=="size") %>% pull(estimate)

# I am pulling sigma which is the _modeled_ standard deviation of the residuals.  Merow uses observed.  
params$growth.sd <- growth.models %>% filter(name=="1_linear") %>% unnest(glance) %>% pull(sigma)
```

Set up the functions
```{r}
# 1. survival probability function
##This is inverse logit.  Provides survival probability based on size.
s.x=function(x,params) {
  u=exp(params$surv.int + params$surv.slope1*x + params$surv.slope2*(x^2))
  return(u/(1+u))
}

# 2. growth function
## Return a probability distribution of new sizes at t+1 (xp) at a given size x.  
g.yx=function(xp,x,params) {
  dnorm(xp,mean=(params$growth.int + params$growth.slope*x), sd=params$growth.sd)
}
```

Make a kernel
Define the parameters for integration
```{r}
# the sizes we are integrating over
min.size=.9*min(c(d94$size, d94$sizeNext),na.rm=T)
max.size=1.1*max(c(d94$size, d94$sizeNext),na.rm=T)

n=100 # dimensions of the matrix

b=min.size+c(0:n)*(max.size-min.size)/n # boundary points
y=0.5*(b[1:n]+b[2:(n+1)]) # mesh points
h=y[2]-y[1] # step size
```

Make the matrix
```{r}
G=h*outer(y,y,g.yx,params=params) # growth matrix
# OK for each value of y,y evaluate the growth function g.yx using params
# If I understand this correctly, the rows are x(t) and the columns are x(t+1)

S=s.x(y,params=params) # survival
# F=h*outer(y,y,f.yx,params=params) # reproduction matrix
P=G # placeholder; redefine P on the next line
for(i in 1:n) P[,i]=G[,i]*S[i] # growth/survival matrix
# K=P+F # full matrix
```

Correct the matrix using the Merow et al "constant" correction code

```{r} 
Pc <- P # a new, corrected matrix
nvals <- colSums(Pc, na.rm = TRUE)
loc0 <- which(nvals == 0, arr.ind = TRUE)

if (length(loc0) > 0) {
  print("warnings - columns that sum to 0 or that have NAs - assuming survival is along the diagonal; plot your Pmatrix to check it")
  Pc[, loc0] <- 0
  Pc[cbind(loc0, loc0)] <- s.x(y[loc0], params=params)
}
nvals <- colSums(Pc, na.rm = TRUE)
Pc <- t((t(Pc)/nvals) * s.x(y, params=params))
```

Test matrix division

```{r}
m <- matrix(1:9, nrow=3)
nvals <- colSums(m)
m
nvals
t(m) / nvals
```


Plot the matrix
```{r}
P %>% as_tibble() %>%
  set_colnames(y) %>% 
  mutate(size.t1=y) %>%
  pivot_longer(-size.t1, names_to = "size.t", names_transform = as.numeric) %>%
  ggplot(aes(x=size.t, y = size.t1)) +
  geom_raster(aes(fill = value)) +
  geom_contour(aes(z = value),lwd=.25) +
  geom_abline(intercept=0, slope = 1, color="gray90", lty=5) +
  scale_fill_viridis_c(option = "plasma") +
  labs(x = "Size (t)", y = "Size (t + 1)", title = "P Matrix: Size and Growth") +
  coord_equal() +
  theme_bw()
```

Plot the CORRECTED matrix
```{r}
Pc %>% as_tibble() %>%
  set_colnames(y) %>% 
  mutate(size.t1=y) %>%
  pivot_longer(-size.t1, names_to = "size.t", names_transform = as.numeric) %>%
  ggplot(aes(x=size.t, y = size.t1)) +
  geom_raster(aes(fill = value)) +
  geom_contour(aes(z = value),lwd=.25) +
  geom_abline(intercept=0, slope = 1, color="gray90", lty=5) +
  scale_fill_viridis_c(option = "plasma") +
  labs(x = "Size (t)", y = "Size (t + 1)", title = "P Matrix (corrected): Size and Growth") +
  coord_equal() +
  theme_bw()
```

evaluate P matrix -- Need a non-IPMpack way of doing this.
Merow (appendix C) says:

Calling the function PMatrixDiagnostics yields a figure
illustrating whether survival, life expectancy, and populations structure change with an increase
in the number of bins or an increase in the size range. 

To make this work I think we need a function that makes the Matrix for us based on some parameters (bin size, etc).  Then this can be called repeatedly.
```{r, eval=FALSE}
diagnosticsPmatrix(Pmatrix94, growObj = go94, survObj = so94,
                   correction = "constant")
```


#### Fecundity Kernel

Compare different models for flowering probability:
getting an error
```{r, eval=FALSE}
fec0.0_94 <- makeFecObj(d94, Formula = fec0~1, Family = "binomial")
fec0.1_94 <- makeFecObj(d94, Formula = fec0~size, Family = "binomial")
fec0.2_94 <- makeFecObj(d94, Formula = fec0~size+size2, Family = "binomial")
fec0.3_94 <- makeFecObj(d94, Formula = fec0~size+size2+size3, Family = "binomial")
```

```{r}
newdata = tibble(size=seq(min(d94$size, na.rm = TRUE), max(d94$size, na.rm = TRUE), length.out=101))

fec0.models <- tibble(name=c("0_intercept", "1_linear", "2_quadratic", "3_cubic"),
                      f = c("fec0 ~ 1", 
                            "fec0 ~ size", 
                            "fec0 ~ size + I(size^2)", 
                            "fec0 ~ size + I(size^2) + I(size^3)"))

fec0.models <- fec0.models %>%
  mutate(glm = map(f, ~ glm(as.formula(.), data = d94, family = "binomial") ),
         predict = map(glm, ~ cbind(newdata, fec0=predict(., newdata, type = "response"))),
         glance = map(glm, glance))

fec0.models %>% unnest(glance)

# AIC lowest in cubic, although not by a ton.
```

plot predictions:
```{r}
fec0.predictions <- fec0.models %>% dplyr::select(name, predict) %>% unnest(predict)

d94 %>% filter(!is.na(fec0)) %>%
  ggplot(aes(x=size, y =fec0)) +
  geom_point() +
  geom_line(aes(color=name), data = fec0.predictions) +
  scale_color_viridis_d(end = 0.9)
```


Compare different models for number of fruits
getting an error
```{r, eval=FALSE}
fec1.0_94 <- makeFecObj(d94, Formula = fec1~1, Family = "poisson")
fec1.1_94 <- makeFecObj(d94, Formula = fec1~size, Family = "poisson")
fec1.2_94 <- makeFecObj(d94, Formula = fec1~size+size2, Family = "poisson")
fec1.3_94 <- makeFecObj(d94, Formula = fec1~size+size2+size3, Family = "poisson")
```

```{r}
fec1.models <- tibble(name=c("0_intercept", "1_linear", "2_quadratic", "3_cubic"),
                      f = c("fec1 ~ 1", 
                            "fec1 ~ size", 
                            "fec1 ~ size + I(size^2)", 
                            "fec1 ~ size + I(size^2) + I(size^3)"))

fec1.models <- fec1.models %>%
  mutate(glm = map(f, ~ glm(as.formula(.), data = d94, family = "poisson") ),
         predict = map(glm, ~ cbind(newdata, fec1=predict(., newdata, type = "response"))),
         glance = map(glm, glance))

fec1.models %>% unnest(glance)

# AIC clearly lowest in cubic
```

plot predictions:
```{r}
fec1.predictions <- fec1.models %>% dplyr::select(name, predict) %>% unnest(predict)

d94 %>% filter(!is.na(fec1)) %>%
  ggplot(aes(x=size, y =fec1)) +
  geom_point() +
  geom_line(aes(color=name), data = fec1.predictions) +
  scale_color_viridis_d(end=.9)
```

put it together:
```{r, eval = FALSE}
d94$stage = factor(d94$stage)
fo94 <- makeFecObj(d94, Formula=c(fec0~size+size2, fec1~size+size2+size3),
                   Family=c("binomial", "poisson"),
                   Transform=c("none", -1),
                   meanOffspringSize=mean(d94[is.na(d1$size)==TRUE &
                                                is.na(d94$sizeNext)==FALSE,"sizeNext"]),
                   sdOffspringSize=sd(d94[is.na(d1$size)==TRUE &
                                            is.na(d94$sizeNext)==FALSE,"sizeNext"]),
                   fecConstants=data.frame(fec2=fec2,fec3=fec3,fec4=fec4),
                   offspringSplitter=data.frame(seedbank=goSB,
                                                continuous=(1-goSB)),
                   vitalRatesPerOffspringType=data.frame(
                     seedbank=c(1,1,1,0,0),
                     continuous=c(1,1,1,1,1),
                     row.names=c("fec0","fec1",
                                 "fec2","fec3","fec4")))
```
```{r}
vitalRatesPerOffspringType=data.frame(seedbank=c(1,1,1,0,0),
                                      continuous=c(1,1,1,1,1),
                                      row.names=c("fec0","fec1","fec2","fec3","fec4"))
vitalRatesPerOffspringType
```

Two typos in this sentence? or in the code? I think in the description.  "Closer examination of the data frame will help make it evident that fec0, fec1 and fec3
aﬀect all propagules, whereas fec3 and fec4 only aﬀect seeds in the seed bank."


Make the F Matrix:

```{r}
Fmatrix94 <- makeIPMFmatrix(fecObj=fo94, minSize=minSize, maxSize=maxSize,
nBigMatrix=80, correction="constant")
```

plot it:
```{r,fig.asp=1}
require(fields)
image.plot(Fmatrix94@meshpoints,
Fmatrix94@meshpoints,
t(Fmatrix94[2:maxSize,2:maxSize]),
main = "Fmatrix: fecundity",
xlab = "Size at t",
ylab = "Size at t+1")
```
```{r}
dto94 <- makeDiscreteTrans(d94)
dummy94 <- as.matrix(fo94@offspringRel$coefficients[1])
dimnames(dummy94) <- list(1,"seedbank")
dto94@meanToCont <- as.matrix(dummy94,dimnames=c(1,"seedbank"))
dummy94 <- as.matrix(fo94@sdOffspringSize)
dimnames(dummy94) <- list(1,"seedbank")
dto94@sdToCont <- as.matrix(dummy94,dimnames=c(1,"seedbank"))
dto94@discreteTrans[1,1] <- staySB + (1 - staySB)*fec3*fec4
```

Now recreate the P matrix, considering the discrete stages

```{r}
Pmatrix94 <- makeIPMPmatrix(growObj=go94, survObj=so94, discreteTrans=dto94,
minSize=minSize, maxSize=maxSize, nBigMatrix=80,correction="constant")
library(fields)
image.plot(c(Pmatrix94@meshpoints), c(Pmatrix94@meshpoints),
(log10(t(Pmatrix94[1:80,1:80]))), main = "Pmatrix: survival/growth",
xlab = "Size at t", ylab = "Size at t+1")
#abline(a = 0, b = 1, lty= 2, col = "white", lwd=2)
```

```{r}
IPM94 <- Pmatrix94 + Fmatrix94
IPM94 %>% as_tibble() %>%
  set_colnames(1:81) %>% 
  mutate(size.t1=1:81) %>%
  pivot_longer(-size.t1, names_to = "size.t", names_transform = as.numeric) %>%
  ggplot(aes(x=size.t, y = size.t1)) +
  geom_raster(aes(fill = value)) +
  geom_contour(aes(z = value),lwd=.25) +
  #geom_abline(intercept=0, slope = 1, color="gray90", lty=5) +
  scale_fill_viridis_c(option = "plasma") +
  labs(x = "Size (t)", y = "Size (t + 1)", title = "IPM Matrix") +
  coord_equal() +
  theme_bw()
```
### Julin F-Matrix

Based on the text and my above modeling

fec0 (flowering probability) is modeled as a binomal, with a cubic predictor on size
fec1 (number of fruits) is modeled as Poisson, with a cubic predictor on size

other parameters
fec2: number of seed per fruit
fec3: probability of germination (within the year of seed production)
fec4: probability of seedling survival (within the year of seedling production)

goSB: probability of a seed entering the seed bank
staySB: probability of a seed staying in the seed bank

For seed produced, need to model proportion that germinate vs going into the seed bank

The  fecundity kernel should return the number of new plants produced of various sizes based on the sizes of the parent plant.

The PDF defines the kernel as:

fec0(z) * fec1(z)  * fec2 * (1−goSB) * fec3 * fec4 * fec5(z′) +
fec0(z) * fec1(z) * fec2 * goSB * (1−staySB) * fec3 * fec4 * fec5(z′)

The first line is seedlings produced from seed that don't go to the seed bank

The second line is seedlings produced from seed that go to the seedbank but don't stay in the seed bank

We aren't dealing with seeds previously in the seedbank, I assume we will later...

```{r}

fec1.coefs <- fec1.models %>% 
  filter(name == "3_cubic") %>%
  mutate(coefs = map(glm, tidy)) %>% 
  dplyr::select(name, f, coefs) %>% 
  unnest(coefs) 

params$fec1.int <- fec1.coefs %>% filter(term=="(Intercept)") %>%
  pull(estimate)
params$fec1.slope1 <- fec1.coefs %>% filter(term=="size") %>%
  pull(estimate)
params$fec1.slope2 <- fec1.coefs %>% filter(term=="I(size^2)") %>%
  pull(estimate)
params$fec1.slope3 <- fec1.coefs %>% filter(term=="I(size^3)") %>%
  pull(estimate)
params

params$recruit.size.mean=mean(d94$sizeNext[is.na(d$size)], na.rm = TRUE)
params$recruit.size.sd=sd(d94$sizeNext[is.na(d$size)], na.rm = TRUE)

fruit.n <- function(x, params) {
  exp(params$fec1.int + 
        x*params$fec1.slope1 + 
        x^2*params$fec1.slope2 + 
        x^3*params$fec1.slope3)
}

# reality check (compare to plot...looks good)
fruit.n(40, params)
fruit.n(60, params)
fruit.n(80, params)


f.yx=function(xp,x,params) {
# For seed produced this cycle and not going in to the seed bank
  # flowering probability
    predict(fec0.models$glm[[3]], data.frame(size=x), type="response") * 
  # fruit production
    fruit.n(x, params) *
  # number of seed per fruit 
    fec2 *
  # probability of seed germinating at this time
    (1-goSB) *
  # probability of germination (is this not related to seed bank??)
    fec3 *
  # probability of seedling survival
    fec4 *
  # distribution of new sizes
    dnorm(xp,mean=params$recruit.size.mean,sd=params$recruit.size.sd)
  
# For seed produced this cycle and then entering and exiting the seed bank
  # For seed produced this cycle
  # flowering probability
    predict(fec0.models$glm[[3]], data.frame(size=x), type="response") * 
  # fruit production
    fruit.n(x, params) *
  # number of seed per fruit 
    fec2 *
  # probability of seed germinating at this time
    goSB * (1-staySB) *
    # probability of germination (is this not related to seed bank??)
    fec3 *
  # probability of seedling survival
    fec4 *
  # distribution of new sizes
    dnorm(xp,mean=params$recruit.size.mean,sd=params$recruit.size.sd)
}
```

```{r}
F=h*outer(y,y,f.yx,params=params) # reproduction matrix

F %>% as_tibble() %>%
  set_colnames(y) %>% 
  mutate(size.t1=y) %>%
  pivot_longer(-size.t1, names_to = "size.t", names_transform = as.numeric) %>%
  ggplot(aes(x=size.t, y = size.t1)) +
  geom_raster(aes(fill = value)) +
  #geom_contour(aes(z = value),lwd=.25) +
  scale_fill_viridis_c(option = "plasma") +
  labs(x = "Size (t)", y = "Size (t + 1)", title = "F Matrix") +
  coord_equal() +
  theme_bw()
```






