---
title: "Merow Appendix C"
author: "Julin Maloof"
date: "2024-12-04"
output:
  html_document:
    keep_md: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(broom)
#library(patchwork)
#library(ggbeeswarm)
library(magrittr)
#library(nlme)
library(IPMpack)
source("makeDiscreteTrans.R", local=as.environment("package:IPMpack"))
#conflicted::conflict_prefer("select", "dplyr")
```

```{r}
data("dataIPMpackHypericumCov")

d <- dataIPMpackHypericumCov
head(d)
```

```{r}
skimr::skim(d)
```

```{r}
d1 <- subset(d,is.na(d$size)==FALSE | d$ontogenyNext==1)
d1 <- subset(d1,d1$year==1994 | d1$year==1996)
```

some additional data:
```{r}
fec2 <- 13.78 # number of seed per fruit
fec3 <- 0.001336 # probability of germination the year that the seed was prodiced
fec4 <- 0.14 # probability of seedling survival from germination to time of next census
goSB <- 0.08234528 # probability of a seed entering the seed bank
staySB <- 0.671 # probability of a seed staying in the seed bank
```

keep columns that we are going to use
```{r}
d1 <- d1 |>
  dplyr::select(year, size, surv, sizeNext, fec0, fec1)
```

Add a variable to indicate the number of individuals changing size.  This will be be used later.  Set up stage and stage next

```{r}
d1$number <- 1
d1$stageNext <- d1$stage <- "continuous"
d1$stage[is.na(d1$size)] <- NA
d1$stageNext[d1$surv==0] <- "dead"
```

Set up a seedbank data frame
```{r}
seedbank <- data.frame(year="All",size=NA,surv=1,sizeNext=NA,fec0=NA,fec1=NA,
                       stage=c("seedbank","seedbank","continuous"),
                       stageNext=c("seedbank","continuous","seedbank"),
                       number=c(staySB,(1-staySB)*fec3*fec4,1))
seedbank
```
combine data frames
```{r}
d1 <- rbind(d1, seedbank)
```

## Analysis 

### 94 data

```{r}
d94 <- subset(d1, d1$year == "1994" | d1$year == "All")
minSize<-min(d1$size,na.rm=T)
maxSize<-max(d1$size,na.rm=T)
x<-seq(from=minSize,to=maxSize,length=1001)
x0<-data.frame(size=x,size2=x^2,size3=x^3) # for later use
```

#### Survival and Growth Kernel

** Merow Version **

Survival

```{r}
survModelComp(dataf = d94[!is.na(d$size),],
              expVars = c(surv~1, surv~size, surv~size + size2,
                          surv~size + size2 + size3),
              makePlot = TRUE, legendPos = "bottomleft", mainTitle = "Survival")
```
choose quadratic
```{r}
so94 <- makeSurvObj(d94, surv~size+I(size^2))
```


** JM Version **

```{r}
newdata = tibble(size=seq(min(d94$size, na.rm = TRUE), max(d94$size, na.rm = TRUE), length.out=101))

surv.models <- tibble(name=c("0_intercept", "1_linear", "2_quadratic", "3_cubic"),
                      f = c("surv ~ 1", 
                            "surv ~ size", 
                            "surv ~ size + I(size^2)", 
                            "surv ~ size + I(size^2) + I(size^3)"))

surv.models <- surv.models %>%
  mutate(glm = map(f, ~ glm(as.formula(.), data = d94, family = "binomial") ),
         predict = map(glm, ~ cbind(newdata, surv=predict(., newdata, type = "response"))),
         glance = map(glm, glance))

surv.models %>% unnest(glance)
```
AIC is **VERY** similar.  Merow chose quadratic.

```{r}
surv.models$predict[[1]]
```


plot the fits
```{r}
surv.predictions <- surv.models %>% dplyr::select(name, predict) %>% unnest(predict)

d94 %>% filter(!is.na(surv)) %>%
  ggplot(aes(x=size, y =surv)) +
  geom_point() +
  geom_smooth() + 
  geom_line(aes(color=name), data = surv.predictions) +
  scale_color_viridis_d(end = 0.9)
```


Growth

** Merrow **
```{r}
growthModelComp(dataf = d94, expVars = c(sizeNext~1, sizeNext~size,
                                         sizeNext~size + size2, sizeNext~size + size2 + size3), makePlot = TRUE,
                legendPos = "bottomright", mainTitle = "Growth")
abline(a = 0, b = 1, lty= 2, col = "gray", lwd=2)
```

Choose linear
```{r}
go94 <- makeGrowthObj(d94, sizeNext~size)
```

** Julin **

```{r}
growth.models <- tibble(name=c("0_intercept", "1_linear", "2_quadratic", "3_cubic"),
                        f = c("sizeNext ~ 1", 
                              "sizeNext ~ size", 
                              "sizeNext ~ size + I(size^2)", 
                              "sizeNext ~ size + I(size^2) + I(size^3)"))

growth.models <- growth.models %>%
  mutate(lm = map(f, ~ lm(as.formula(.), data = d94) ),
         predict = map(lm, ~ cbind(newdata, sizeNext=predict(., newdata))),
         glance = map(lm, glance))

growth.models %>% unnest(glance)
```
AIC is AIC is lowest for linear.

plot the fits
```{r}
growth.predictions <- growth.models %>% dplyr::select(name, predict) %>% unnest(predict)

d94 %>% filter(!is.na(sizeNext) & !is.na(size)) %>%
  ggplot(aes(x=size, y =sizeNext)) +
  geom_point() +
  geom_abline(slope = 1, intercept = 0, lty = 5, color = "grey") + 
  geom_line(aes(color=name), data = growth.predictions) +
  scale_color_viridis_d(end = 0.9)
```


Make the P matrix and plot it

** Merow **
```{r, fig.asp=1}
Pmatrix94 <- makeIPMPmatrix(survObj = so94, growObj = go94,
                            minSize = minSize, maxSize = maxSize,
                            nBigMatrix = 80, correction = "constant")
require(fields)
image.plot(Pmatrix94@meshpoints,
           Pmatrix94@meshpoints,
           t(Pmatrix94),
           main = "Pmatrix: survival and growth",
           xlab = "Size at t",
           ylab = "Size at t+1")
abline(a = 0, b = 1, lty= 2, col = "white", lwd=2)
```
evaluate P matrix
```{r, eval=FALSE}
diagnosticsPmatrix(Pmatrix94, growObj = go94, survObj = so94,
                   correction = "constant")
```

** Julin **

parameter data frame

```{r}
params=data.frame(
  surv.int=NA, # Intercept from logistic regression of survival
  surv.slope1=NA, # Slope from logistic regression of survival
  surv.slope2=NA, # Quadratic slope from logistic regression of survival
  growth.int=NA, # Intercept from linear regression of growth
  growth.slope=NA, # Slope from linear regression of growth
  growth.sd=NA # Residual sd from the linear regression of growth
)
```

Get the survival coefficients.  Using the quadratic model.
```{r}
# I am using broom:tidy to create a df with the coefficients.
surv.coefs <- surv.models %>% 
  filter(name == "2_quadratic") %>%
  mutate(coefs = map(glm, tidy)) %>% 
  dplyr::select(name, f, coefs) %>% 
  unnest(coefs) 

params$surv.int <- surv.coefs %>% filter(term == "(Intercept)") %>% pull(estimate)
params$surv.slope1 <- surv.coefs %>% filter(term == "size") %>% pull(estimate)
params$surv.slope2 <- surv.coefs %>% filter(term == "I(size^2)") %>% pull(estimate)
```

Get the growth coefficients.  Using the linear model.
```{r}
growth.coefs <- growth.models %>% 
  filter(name == "1_linear") %>%
  mutate(coefs = map(lm, tidy)) %>% 
  dplyr::select(name, f, coefs) %>% 
  unnest(coefs)

params$growth.int <- growth.coefs %>% filter(term=="(Intercept)") %>% pull(estimate)
params$growth.slope <- growth.coefs %>% filter(term=="size") %>% pull(estimate)

# I am pulling sigma which is the _modeled_ standard deviation of the residuals.  Merow uses observed.  
params$growth.sd <- growth.models %>% filter(name=="1_linear") %>% unnest(glance) %>% pull(sigma)

params$recruit.size.mean=mean(d94$sizeNext[is.na(d$size)], na.rm = TRUE)
params$recruit.size.sd=sd(d94$sizeNext[is.na(d$size)], na.rm = TRUE)
```

Set up the functions
```{r}
# 1. survival probability function
##This is inverse logit.  Provides survival probability based on size.
s.x=function(x,params) {
  u=exp(params$surv.int + params$surv.slope1*x + params$surv.slope2*(x^2))
  return(u/(1+u))
}

# 2. growth function
## Return a probability distribution of new sizes at t+1 (xp) at a given size x.  
g.yx=function(xp,x,params) {
  dnorm(xp,mean=(params$growth.int + params$growth.slope*x), sd=params$growth.sd)
}
```

Make a kernel
Define the parameters for integration
```{r}
nBigMatrix <- 80 # matrix dimensions
  b <- minSize + c(0:nBigMatrix) * (maxSize - minSize)/nBigMatrix # boundary points
  y <- 0.5 * (b[1:nBigMatrix] + b[2:(nBigMatrix + 1)]) # mesh points
  h <- y[2] - y[1] # step size

```

Make the matrix
```{r}
G=h*outer(y,y,g.yx,params=params) # growth matrix
# OK for each value of y,y evaluate the growth function g.yx using params
# If I understand this correctly, the rows are x(t) and the columns are x(t+1)

S=s.x(y,params=params) # survival
# F=h*outer(y,y,f.yx,params=params) # reproduction matrix
P=G # placeholder; redefine P on the next line
for(i in 1:n) P[,i]=G[,i]*S[i] # growth/survival matrix
# K=P+F # full matrix
```

Correct the matrix using the Merow et al "constant" correction code

```{r} 
Pc <- P # a new, corrected matrix
nvals <- colSums(Pc, na.rm = TRUE)
loc0 <- which(nvals == 0, arr.ind = TRUE)

if (length(loc0) > 0) {
  print("warnings - columns that sum to 0 or that have NAs - assuming survival is along the diagonal; plot your Pmatrix to check it")
  Pc[, loc0] <- 0
  Pc[cbind(loc0, loc0)] <- s.x(y[loc0], params=params)
}
nvals <- colSums(Pc, na.rm = TRUE)
Pc <- t((t(Pc)/nvals) * s.x(y, params=params))
```

Test matrix division

```{r}
m <- matrix(1:9, nrow=3)
nvals <- colSums(m)
m
nvals
t(m) / nvals
```


Plot the matrix
```{r}
P %>% as_tibble() %>%
  set_colnames(y) %>% 
  mutate(size.t1=y) %>%
  pivot_longer(-size.t1, names_to = "size.t", names_transform = as.numeric) %>%
  ggplot(aes(x=size.t, y = size.t1)) +
  geom_raster(aes(fill = value)) +
  geom_contour(aes(z = value),lwd=.25) +
  geom_abline(intercept=0, slope = 1, color="gray90", lty=5) +
  scale_fill_viridis_c(option = "plasma") +
  labs(x = "Size (t)", y = "Size (t + 1)", title = "P Matrix: Size and Growth") +
  coord_equal() +
  theme_bw()
```

Plot the CORRECTED matrix
```{r}
Pc %>% as_tibble() %>%
  set_colnames(y) %>% 
  mutate(size.t1=y) %>%
  pivot_longer(-size.t1, names_to = "size.t", names_transform = as.numeric) %>%
  ggplot(aes(x=size.t, y = size.t1)) +
  geom_raster(aes(fill = value)) +
  geom_contour(aes(z = value),lwd=.25) +
  geom_abline(intercept=0, slope = 1, color="gray90", lty=5) +
  scale_fill_viridis_c(option = "plasma") +
  labs(x = "Size (t)", y = "Size (t + 1)", title = "P Matrix (corrected): Size and Growth") +
  coord_equal() +
  theme_bw()
```

evaluate P matrix -- Need a non-IPMpack way of doing this.
Merow (appendix C) says:

Calling the function PMatrixDiagnostics yields a figure
illustrating whether survival, life expectancy, and populations structure change with an increase
in the number of bins or an increase in the size range. 

To make this work I think we need a function that makes the Matrix for us based on some parameters (bin size, etc).  Then this can be called repeatedly.
```{r, eval=FALSE}
diagnosticsPmatrix(Pmatrix94, growObj = go94, survObj = so94,
                   correction = "constant")
```


#### Fecundity Kernel

Compare different models for flowering probability:
getting an error
```{r, eval=FALSE}
fec0.0_94 <- makeFecObj(d94, Formula = fec0~1, Family = "binomial")
fec0.1_94 <- makeFecObj(d94, Formula = fec0~size, Family = "binomial")
fec0.2_94 <- makeFecObj(d94, Formula = fec0~size+size2, Family = "binomial")
fec0.3_94 <- makeFecObj(d94, Formula = fec0~size+size2+size3, Family = "binomial")
```

```{r}
newdata = tibble(size=seq(min(d94$size, na.rm = TRUE), max(d94$size, na.rm = TRUE), length.out=101))

fec0.models <- tibble(name=c("0_intercept", "1_linear", "2_quadratic", "3_cubic"),
                      f = c("fec0 ~ 1", 
                            "fec0 ~ size", 
                            "fec0 ~ size + I(size^2)", 
                            "fec0 ~ size + I(size^2) + I(size^3)"))

fec0.models <- fec0.models %>%
  mutate(glm = map(f, ~ glm(as.formula(.), data = d94, family = "binomial") ),
         predict = map(glm, ~ cbind(newdata, fec0=predict(., newdata, type = "response"))),
         glance = map(glm, glance))

fec0.models %>% unnest(glance)

# AIC lowest in cubic, although not by a ton.
```

plot predictions:
```{r}
fec0.predictions <- fec0.models %>% dplyr::select(name, predict) %>% unnest(predict)

d94 %>% filter(!is.na(fec0)) %>%
  ggplot(aes(x=size, y =fec0)) +
  geom_point() +
  geom_line(aes(color=name), data = fec0.predictions) +
  scale_color_viridis_d(end = 0.9)
```


Compare different models for number of fruits
getting an error
```{r, eval=FALSE}
fec1.0_94 <- makeFecObj(d94, Formula = fec1~1, Family = "poisson")
fec1.1_94 <- makeFecObj(d94, Formula = fec1~size, Family = "poisson")
fec1.2_94 <- makeFecObj(d94, Formula = fec1~size+size2, Family = "poisson")
fec1.3_94 <- makeFecObj(d94, Formula = fec1~size+size2+size3, Family = "poisson")
```

```{r}
fec1.models <- tibble(name=c("0_intercept", "1_linear", "2_quadratic", "3_cubic"),
                      f = c("fec1 ~ 1", 
                            "fec1 ~ size", 
                            "fec1 ~ size + I(size^2)", 
                            "fec1 ~ size + I(size^2) + I(size^3)"))

fec1.models <- fec1.models %>%
  mutate(glm = map(f, ~ glm(as.formula(.), data = d94, family = "poisson") ),
         predict = map(glm, ~ cbind(newdata, fec1=predict(., newdata, type = "response"))),
         glance = map(glm, glance))

fec1.models %>% unnest(glance)

# AIC clearly lowest in cubic
```

plot predictions:
```{r}
fec1.predictions <- fec1.models %>% dplyr::select(name, predict) %>% unnest(predict)

d94 %>% filter(!is.na(fec1)) %>%
  ggplot(aes(x=size, y =fec1)) +
  geom_point() +
  geom_line(aes(color=name), data = fec1.predictions) +
  scale_color_viridis_d(end=.9)
```

put it together:
```{r, eval = TRUE}
d94$stage = factor(d94$stage)
fo94 <- makeFecObj(d94, Formula=c(fec0~size+size2, fec1~size+size2+size3),
                   Family=c("binomial", "poisson"),
                   Transform=c("none", -1),
                   meanOffspringSize=mean(d94[is.na(d1$size)==TRUE &
                                                is.na(d94$sizeNext)==FALSE,"sizeNext"]),
                   sdOffspringSize=sd(d94[is.na(d1$size)==TRUE &
                                            is.na(d94$sizeNext)==FALSE,"sizeNext"]),
                   fecConstants=data.frame(fec2=fec2,fec3=fec3,fec4=fec4),
                   offspringSplitter=data.frame(seedbank=goSB,
                                                continuous=(1-goSB)),
                   vitalRatesPerOffspringType=data.frame(
                     seedbank=c(1,1,1,0,0),
                     continuous=c(1,1,1,1,1),
                     row.names=c("fec0","fec1",
                                 "fec2","fec3","fec4")))
```

```{r}
vitalRatesPerOffspringType=data.frame(seedbank=c(1,1,1,0,0),
                                      continuous=c(1,1,1,1,1),
                                      row.names=c("fec0","fec1","fec2","fec3","fec4"))
vitalRatesPerOffspringType
```

Two typos in this sentence? or in the code? I think in the description.  "Closer examination of the data frame will help make it evident that fec0, fec1 and fec3
aﬀect all propagules, whereas fec3 and fec4 only aﬀect seeds in the seed bank."


Make the F Matrix:

```{r}
#Fmatrix94 <- makeIPMFmatrix(fecObj=fo94, minSize=minSize, maxSize=maxSize,
#                            nBigMatrix=80, correction="constant")

# Go with no correction for the moment, to try to reconcile with my code:

Fmatrix94 <- makeIPMFmatrix(fecObj=fo94, minSize=minSize, maxSize=maxSize,
                            nBigMatrix=80, correction="none")
```

plot it:
```{r,fig.asp=1}
require(fields)
image.plot(Fmatrix94@meshpoints,
           Fmatrix94@meshpoints,
           t(Fmatrix94[2:maxSize,2:maxSize]),
           main = "Fmatrix: fecundity",
           xlab = "Size at t",
           ylab = "Size at t+1")
```

```{r}
dto94 <- makeDiscreteTrans(d94)
dummy94 <- as.matrix(fo94@offspringRel$coefficients[1])
dimnames(dummy94) <- list(1,"seedbank")
dto94@meanToCont <- as.matrix(dummy94,dimnames=c(1,"seedbank"))
dummy94 <- as.matrix(fo94@sdOffspringSize)
dimnames(dummy94) <- list(1,"seedbank")
dto94@sdToCont <- as.matrix(dummy94,dimnames=c(1,"seedbank"))
dto94@discreteTrans[1,1] <- staySB + (1 - staySB)*fec3*fec4 ## I think this is incorrect

# Correct code should be:
# dto94@discreteTrans[1:2,1] <- c(staySB, (1 - staySB)*fec3*fec4)
# commented out for now, so I can match appendix C


```

Now recreate the P matrix, considering the discrete stages

```{r}
Pmatrix94.old <- Pmatrix94 # for comparison, so I can understand what changed

Pmatrix94 <- makeIPMPmatrix(growObj=go94, survObj=so94, discreteTrans=dto94,
                            minSize=minSize, maxSize=maxSize, nBigMatrix=80,correction="constant")
library(fields)
image.plot(c(Pmatrix94@meshpoints), c(Pmatrix94@meshpoints),
           (log10(t(Pmatrix94[1:80,1:80]))), main = "Pmatrix: survival/growth",
           xlab = "Size at t", ylab = "Size at t+1")
#abline(a = 0, b = 1, lty= 2, col = "white", lwd=2)
```
```{r}
dim(Pmatrix94.old)
dim(Pmatrix94)

Pmatrix94[1:5, 1:5]
Pmatrix94[2:6, 2:6]
Pmatrix94.old[1:5, 1:5]

Pmatrix94[77:81, 77:81]
Pmatrix94.old[76:80, 76:80]

all.equal(Pmatrix94[-1, -1], Pmatrix94.old[,])
```
So the first row and column are new, everything else is the same.

Now size "1" (or the first row and column) represent seed, or seedbank

The first column represents the size of new recruits from the seed bank.

plot the first column

```{r}
plot(Pmatrix94[,1], type="l")
plot(Pmatrix94[2:20,1], type="l")
```

After removing the first cell, this is peaking near 4, which is consistent with the average size of new recruits (4.36)

What is the first cell? This is seed that stay in the seed bank.
```{r}
Pmatrix94[1,1]
staySB
```

What about the first row?

```{r}
plot(Pmatrix94[1,], type = "l")
plot(Pmatrix94[1,-1], type = "l")

```
I would have thought that this is seed going to the seedbank from parent plants of different sizes but it doesn't make any sense for it to be peaking at 25, fecundity is ~ 0 at this stage.  Also seems like maybe this should just be handled by the F-matrix and this could be set to zero?

```{r}
IPM94 <- Pmatrix94 + Fmatrix94
IPM94 %>% as_tibble() %>%
  set_colnames(1:81) %>% 
  mutate(size.t1=1:81) %>%
  pivot_longer(-size.t1, names_to = "size.t", names_transform = as.numeric) %>%
  ggplot(aes(x=size.t, y = size.t1)) +
  geom_raster(aes(fill = value)) +
  geom_contour(aes(z = value),lwd=.25) +
  #geom_abline(intercept=0, slope = 1, color="gray90", lty=5) +
  scale_fill_viridis_c(option = "plasma") +
  labs(x = "Size (t)", y = "Size (t + 1)", title = "IPM Matrix") +
  coord_equal() +
  theme_bw()
```

### Julin Update P-Matrix to include seed from seedbank

```{r}

correction <- "constant"

toContinuousP <- dnorm(y, params$recruit.size.mean, params$recruit.size.sd) * h

if (correction == "constant") 
        toContinuousP <- toContinuousP/sum(toContinuousP)

toContinuousP[which(is.na(toContinuousP))] <- 0

toContinuousP <- toContinuousP * 9.169973e-05 # This is the value in dto94, calculated from observations in the `makeDiscreteTrans` function

# Alternative, and I think this is what they meant to do (but had a typo)
# toContinuousP <- toContinuousP * (1 - staySB)*fec3*fec4 

toDiscreteP <- c(staySB, rep(0,ncol(Pc))) # in our case, continuous plants do not transition to the discrete class.  Seed going to seedbank is handled by F, see below.

Pc <- rbind(toDiscreteP, cbind(toContinuousP, Pc))
```

compare it
```{r}
range(Pc)
range(Pmatrix94)

Pc[1:5, 1:5]
Pmatrix94[1:5, 1:5]
```


### Julin F-Matrix

I SHOULD UPDATE THIS TO INCLUDE THE "CORRECTION" CODE, LIKE I DID FOR THE P MATRIX

Based on the text and my above modeling

fec0 (flowering probability) is modeled as a binomal, with a cubic predictor on size
fec1 (number of fruits) is modeled as Poisson, with a cubic predictor on size

other parameters
fec2: number of seed per fruit
fec3: probability of germination (within the year of seed production)
fec4: probability of seedling survival (within the year of seedling production)

goSB: probability of a seed entering the seed bank
staySB: probability of a seed staying in the seed bank

For seed produced, need to model proportion that germinate vs going into the seed bank

The  fecundity kernel should return the number of new plants produced of various sizes based on the sizes of the parent plant.

The PDF defines the kernel as:

fec0(z) * fec1(z)  * fec2 * (1−goSB) * fec3 * fec4 * fec5(z′) +
fec0(z) * fec1(z) * fec2 * goSB * (1−staySB) * fec3 * fec4 * fec5(z′)

The first line is seedlings produced from seed that don't go to the seed bank

The second line is seedlings produced from seed that go to the seedbank but don't stay in the seed bank

We aren't dealing with seeds previously in the seedbank, I assume we will later...

```{r}

fec1.coefs <- fec1.models %>% 
  filter(name == "3_cubic") %>%
  mutate(coefs = map(glm, tidy)) %>% 
  dplyr::select(name, f, coefs) %>% 
  unnest(coefs) 

params$fec1.int <- fec1.coefs %>% filter(term=="(Intercept)") %>%
  pull(estimate)
params$fec1.slope1 <- fec1.coefs %>% filter(term=="size") %>%
  pull(estimate)
params$fec1.slope2 <- fec1.coefs %>% filter(term=="I(size^2)") %>%
  pull(estimate)
params$fec1.slope3 <- fec1.coefs %>% filter(term=="I(size^3)") %>%
  pull(estimate)
params

fruit.n <- function(x, params) {
  exp(params$fec1.int + 
        x*params$fec1.slope1 + 
        x^2*params$fec1.slope2 + 
        x^3*params$fec1.slope3)
}

# reality check (compare to plot...looks good)
fruit.n(40, params)
fruit.n(60, params)
fruit.n(80, params)


f.yx=function(xp,x,params) {
  # For seed produced this cycle and not going in to the seed bank
  # flowering probability
  predict(fec0.models$glm[[3]], data.frame(size=x), type="response") * 
    # fruit production
    fruit.n(x, params) *
    # number of seed per fruit 
    fec2 *
    # probability of seed not going to seed bank
    (1-goSB) *
    # probability of germination (is this not related to seed bank??)
    fec3 *
    # probability of seedling survival
    fec4 *
    # distribution of new sizes
    dnorm(xp,mean=params$recruit.size.mean,sd=params$recruit.size.sd)  # + 
  
  # note that the code below is commented out because it is not included in makeIPMFmatrix function
  # So to get my numbers to match this has to be removed
    
    # # For seed produced this cycle and then entering and exiting the seed bank
    # # For seed produced this cycle
    # # flowering probability
    # predict(fec0.models$glm[[3]], data.frame(size=x), type="response") * 
    # # fruit production
    # fruit.n(x, params) *
    # # number of seed per fruit 
    # fec2 *
    # # probability of seed germinating at this time
    # goSB * (1-staySB) *
    # # probability of germination (is this not related to seed bank??)
    # fec3 *
    # # probability of seedling survival
    # fec4 *
    # # distribution of new sizes
    # dnorm(xp,mean=params$recruit.size.mean,sd=params$recruit.size.sd)
}

```

```{r}
# define the constants as done in makeIPMFmatrix
nBigMatrix <- 80
  b <- minSize + c(0:nBigMatrix) * (maxSize - minSize)/nBigMatrix
  y <- 0.5 * (b[1:nBigMatrix] + b[2:(nBigMatrix + 1)])
  h <- y[2] - y[1]
  
toContinuousF=h*outer(y,y,f.yx,params=params) # reproduction matrix for seed that germinate in this cycle
```

Update the matrix for the discrete classes
```{r}
fToSeedbank.x=function(x,params) {
  # For seed produced this cycle and going to the seed bank
  # flowering probability
  predict(fec0.models$glm[[3]], data.frame(size=x), type="response") * 
    # fruit production
    fruit.n(x, params) *
    # number of seed per fruit 
    fec2 *
    # probability of seed going to seedbank
    (goSB)
}

toDiscreteF <- fToSeedbank.x(y, params = params)

fromDiscreteF <- rep(0, nrow(toContinuousF) + 1) # 0 because the discrete class (seedbank) is not contributing seed
                                    # length is nrow + 1 because the matrix is being expanded for the discrete class
                                    # This would need to be a matrix instead of a vector if there were more than 1 discrete class

F <- cbind(fromDiscreteF, rbind(toDiscreteF, toContinuousF))
```

Plot it
```{r}
F %>% as_tibble() %>%
  set_colnames(c(0.5,y)) %>% 
  mutate(size.t1=c(0.5,y)) %>%
  pivot_longer(-size.t1, names_to = "size.t", names_transform = as.numeric) %>%
  ggplot(aes(x=size.t, y = size.t1)) +
  geom_raster(aes(fill = value)) +
  #geom_contour(aes(z = value),lwd=.25) +
  scale_fill_viridis_c(option = "plasma") +
  labs(x = "Size (t)", y = "Size (t + 1)", title = "F Matrix") +
  coord_equal() +
  theme_bw()
```
```{r}
F[2:maxSize,2:maxSize] %>% as_tibble() %>%
  set_colnames(y) %>% 
  mutate(size.t1=y) %>%
  pivot_longer(-size.t1, names_to = "size.t", names_transform = as.numeric) %>%
  ggplot(aes(x=size.t, y = size.t1)) +
  geom_raster(aes(fill = value)) +
  #geom_contour(aes(z = value),lwd=.25) +
  scale_fill_viridis_c(option = "plasma") +
  labs(x = "Size (t)", y = "Size (t + 1)", title = "F Matrix") +
  coord_equal() +
  theme_bw()
```
Compare to IPM F-matrix
First compare the toDiscrete row:
```{r}
plot(F[1,], Fmatrix94[1,])
abline(1,1)
```

```{r}
range(F)
range(Fmatrix94)
range(F[-1,])
range(Fmatrix94[-1,])
```

### MAke the IPMmatrix

```{r}
I <- Pc + F
```


## Dissecting the IPMpack functions

The IPMpack function `makeIPMFmatrix` ultimately calls `.fecPreCensus` which in turn calls `.fecRaw`.  I paste the code from this below to try to understand what is going on.

Note that makeIPMFmatrix is using `outer()` to call `.fecPreCenesus` thereby generating a matrix.  In the code below I am just running it once (no outer).

### .fecRaw
```{r}
# this would get run for each pairwise combination of sizes in the matrix

cov = data.frame(covariate = 1)

fecObj <- fo94

x <- y

newd <- data.frame(cbind(cov, size = x), stringsAsFactors = FALSE)
newd$size2 <- x^2
newd$size3 <- x^3
newd$expsize <- exp(x)
fecObj@fecConstants[is.na(fecObj@fecConstants)] <- 1

# Matrix to hold the new values
fecValues <- matrix(c(rep(1, length(fecObj@fitFec)), unlist(fecObj@fecConstants)), 
                    ncol = length(x), nrow = length(fecObj@fitFec) + length(fecObj@fecConstants))

dim(fecValues) # 5 rows (1 for each of fec0, fec1, etc)
# 1 columns, one for each value of x

fecValues
```


```{r}
# Now for each vital rate model do the predictions
# We have two prediction functions, so these will replace the "1"s in the first two rows of fecValues
for (i in 1:length(fecObj@fitFec)) fecValues[i, ] <- predict(fecObj@fitFec[[i]], 
                                                             newd, type = "response")

# the updated values
fecValues
```


```{r}
# this section will transform data in fecValues on a per-row basis, depending on the values in the @Transform slot
if (length(grep("log", fecObj@Transform)) > 0) 
  for (i in grep("log", fecObj@Transform)) fecValues[i, ] <- exp(fecValues[i, ])
if (length(grep("exp", fecObj@Transform)) > 0) 
  for (i in grep("exp", fecObj@Transform)) fecValues[i, ] <- log(fecValues[i, ])
if (length(grep("sqrt", fecObj@Transform)) > 0) 
  for (i in grep("sqrt", fecObj@Transform)) fecValues[i, ] <- (fecValues[i, ])^2
if (length(grep("-1", fecObj@Transform)) > 0) 
  for (i in grep("-1", fecObj@Transform)) fecValues[i, ] <- fecValues[i, ] + 1
## WHY IS THE TRANSFORM FUNCTION ABOVE +1?  I GUESS THESE ARE BACK TRANFORMATIONS FROM A PREVIOUS TRANSFORMATION.
fecValues
```


```{r}
# Now we apply the offspring splitter
if (length(which(fecObj@vitalRatesPerOffspringType[, "continuous"] == 1)) > 1) {
  
  prodFecValues <- 
    # subset the fecValue matrix to keep rows where "continuous" applies (in our case, all rows)
    # then, calculate the product of the rates in each column
    # this returns a vector of length y (80 in our case; aka x)
    apply(
      fecValues[which(fecObj@vitalRatesPerOffspringType[, "continuous"] == 1), ] , 2, prod) *
    
    # now multiply that vector of seed production by the offspring spliter value for "continuous" so that the rates reflect seed that are NOT going into the seedbank (1-goSB)
    unlist(fecObj@offspringSplitter["continuous"])
} else {
  # this is if there is only 1 vital rate contributing to "continuous" in which case no multiplication is needed
  # does not apply to us
  prodFecValues <- fecValues[which(fecObj@vitalRatesPerOffspringType[, 
                                                                     "continuous"] == 1), ] * unlist(fecObj@offspringSplitter["continuous"])
}

#     return(list(prodFecValues, fecValues))

# The end result of this (prodFecValues) is a vector of 80 rates, that represent seed production for each of the 80 size classes.
# This will eventually be one row in the fecundity matrix, after we do additional calculations
# To get the F Matrix from this, we need to calculate the size classes from new germinants.  That code is in `.fecPreCensus`
```

### .fecPreCensus

For a single value of SizeNext
```{r}
# these arguments are set in the function call.  the code below sets it up the way it is being run for Appendix C
cov = data.frame(covariate = 1)
offspringObj = NULL

# calculate this for a single value of SizeNext:
this.y <- y[2]

newd <- data.frame(cbind(cov, size = x), stringsAsFactors = FALSE)
newd$size2 <- x^2
newd$size3 <- x^3


if (is.null(offspringObj)) { # TRUE for us
  if (length(grep("expsize", fecObj@offspringRel$formula)) > 
      0) { # FALSE
    newd$expsize <- exp(x)
  }
  if (length(grep("logsize", fecObj@offspringRel$formula)) > 
      0) { # FALSE
    newd$logsize <- log(x)
  }
  
  # the line below is getting called.  I am commenting it out and rewriting because it calls `.fecRaw` and I already calculated that above  
  #   u <- .fecRaw(x = x, cov = cov, fecObj = fecObj)[[1]] * 
  #   dnorm(y, predict(fecObj@offspringRel, newdata = newd, 
  #                       type = "response"), fecObj@sdOffspringSize)
  u <- prodFecValues*
    dnorm(this.y, predict(fecObj@offspringRel, newdata = newd, 
                          type = "response"), fecObj@sdOffspringSize)
} else {
  u <- .fecRaw(x = x, cov = cov, fecObj = fecObj)[[1]] * 
    growth(y, y, newd, offspringObj)
}
#  return(u)

u

# u is a vector of 80 that will be one row in the F matrix.
# The way I set this up it is the row that represents the second smallest size class
# It gives the production rates of new seedlings of different sizes based on parent size classes
# It still needs to be corrected for grid size, etc.

```

```{r, fig.asp=1}
tmp2 <- u*h # correct for cell size
plot(tmp2, Fmatrix94[3, -1]) # Compare to IPMpack F Matrix
abline(0,1)
plot(tmp2, F[2,]) # Compare to JNM F Matrix
abline(0,1)

```
Matched makeIPMFmatrix, and very, very close to my calculations.  Slightly more off if I include the second part of my fecundity kernel (see commented out code above)


### makeIPMFmatix

The code below is a subset of what is in makeIPMFmatrix. This is the section that deals with discrete classes.
I am commenting out code that is not applicable to us, just for readability

First generate the basic F matrix (seed being produced from current plants...this is what was picked apart above):

```{r}
chosenCov = data.frame(covariate = 1)

 tmp <- t(outer(X = y, Y = y, .fecPreCensus, cov = chosenCov, 
                     fecObj = fecObj, offspringObj = offspringObj)) * 
        h
```


```{r}
nBigMatrix <- 80
chosenCov = data.frame(covariate = 1)
nEnvClass = 1

get.matrix <- to.cont <- tmp

nDisc <- length(fecObj@offspringSplitter) - 1
namesDiscrete <- "NA"
if (nDisc > 0) { # TRUE for us
  namesDiscrete <- colnames(fecObj@offspringSplitter[1:nDisc])
  to.discrete <- matrix(0, nrow = nDisc, ncol = nBigMatrix)
  for (i in 1:nDisc) {
    # This section is going to calculate the number of seed going to the discrete class (i.e. the seedbank) for each size class
    if (length(which(fecObj@vitalRatesPerOffspringType[, namesDiscrete[i]] == 1)) > 1) {
      to.discrete[i, ] <- 
        apply(
          # get / calculate the fecundity for each size class
          .fecRaw(x = y, 
                  cov = chosenCov, 
                  fecObj = fecObj)[[2]][ # subsetting to get get fecValues, not fecValuesprod
                    which(fecObj@vitalRatesPerOffspringType[, namesDiscrete[i]] == 1), ], # subset the fecValues to get those that apply to the discrete class 
          2, 
          prod) * # get the product of the applicable rates (in this case fec0, 1, 2)
                  # then multiply this fec0,1,2 product by the proportion of seed going to the seedbank
        unlist(fecObj@offspringSplitter[namesDiscrete[i]])
    }
    # else {
    #   to.discrete[i, ] <- .fecRaw(x = y, cov = chosenCov, 
    #                               fecObj = fecObj)[[2]][which(fecObj@vitalRatesPerOffspringType[, 
    #                                                                                             namesDiscrete[i]] == 1), ] * unlist(fecObj@offspringSplitter[namesDiscrete[i]])
    # }
  }
  
  # This section calculates the number of offspring from the discrete class
  # this is Zero, I guess because seed are not producing more seed
  # Seed exiting the seedbank are dealt with in the size/growth matrix
  
  from.discrete <- matrix(0, ncol = nDisc, nrow = nDisc + nBigMatrix)
  
  # if (names(fecObj@fecByDiscrete)[1] != "NA.") {
  #   if (sum(names(fecObj@fecByDiscrete) != namesDiscrete) > 
  #       0) 
  #     stop("Error - the names of the discrete classes as you provided for the data.frame fecByDiscrete are not 100% the same discrete class names in your data.frame offspringSplitter. They should also be in alphabetical order.")
  #   for (j in 1:nDisc) {
  #     for (i in 1:nDisc) {
  #       from.discrete[i, j] <- unlist(fecObj@offspringSplitter[namesDiscrete[i]] * 
  #                                       fecObj@fecByDiscrete[namesDiscrete[j]])
  #     }
  #   }
  #   if (sum(fecObj@fecByDiscrete) > 0 & fecObj@offspringSplitter["continuous"] > 
  #       0) {
  #     print("WARNING - number and sizes of offspring produced by individuals in discrete classes cannot be calculated yet. The Fmatrix contains zeros instead. Only solution at this point: change the F matrix yourself afterwards.")
  #   }
  # }
  
  
   get.matrix <- cbind(from.discrete, rbind(to.discrete, to.cont))

  
}
if (min(get.matrix) < 0) {
  print("Warning: fertility values < 0 exist in matrix, consider transforms. Negative values set to zero")
  get.matrix[get.matrix < 0] <- 0
}
rc <- new("IPMmatrix", nDiscrete = nDisc, nEnvClass = 1, 
          nBigMatrix = nBigMatrix, nrow = 1 * nBigMatrix + nDisc, 
          ncol = 1 * nBigMatrix + nDisc, meshpoints = y, env.index = rep(1:nEnvClass, 
                                                                         each = nBigMatrix), names.discrete = namesDiscrete)
rc[, ] <- get.matrix
#return(rc)
```


```{r}
range(rc[-1,]) # The matrix from dissecting IPMpack code
range(F[-1,]) # My hand calculated one
range(Fmatrix94[-1,]) # The matrix from using IPMpack functions
```

Ignoring the first row (seed going to the seedbank), these are all very similar!
