---
title: "Merow Appendix A"
author: "Julin Maloof"
date: "2024-11-15"
output: 
  html_document: 
    keep_md: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(patchwork)
library(ggbeeswarm)
library(magrittr)
library(nlme)
library(IPMpack)
```


Get the data

```{r}
d <- read.csv("../input/Intro_to_IPMs_Exercises_Data.csv")
head(d)
tail(d)
```

Make some plots

```{r}

p1 <- d %>%  filter(!is.na(surv)) %>%
  ggplot(aes(x=size, y = surv)) +
  geom_point(position=position_jitter(width=0, height = .05), size = .5) +
  labs(y = "Survival to t+1", x = "Size (t)")

p1.alt <- d %>% 
  filter(!is.na(surv)) %>%
  ggplot(aes(as.logical(surv), size)) +
  geom_violin() +
  geom_quasirandom(bandwidth = 1, width = .35, alpha=.5) +
  labs(x = "Survival to t+1", y = "Size (t)")

p2 <- d %>%
  ggplot(aes(x=size, y=sizeNext)) +
  geom_point() +
  labs(x = "Size (t)", y = "Size (t + 1)")

p3 <- d %>%
  ggplot(aes(x=size, y = fec.seed)) +
  geom_point() +
  labs(x = "Size (t)", y = "Seed Number")

p4 <- d %>%
  filter(is.na(size)) %>%
  ggplot(aes(x=sizeNext)) +
  geom_histogram(aes(y=after_stat(density)), bins=15, fill = "lightblue", color="grey20", lwd=.25) + 
  xlab("Recruit Size")

(p1 + p2) / (p3 + p4)

(p1.alt + p2) / (p3 + p4)

```

## Build Regressions for Vital Rate functions

Start with an empty data frame
```{r}
params=data.frame(
  surv.int=NA, # Intercept from logistic regression of survival
  surv.slope=NA, # Slope from logistic regression of survival
  growth.int=NA, # Intercept from linear regression of growth
  growth.slope=NA, # Slope from linear regression of growth
  growth.sd=NA, # Residual sd from the linear regression of growth
  seed.int=NA, # Intercept from Poisson regression of seed number
  seed.slope=NA, # Slope from Poisson regression of seed number
  recruit.size.mean=NA, # Mean recruit size
  recruit.size.sd=NA, # Standard deviation of recruit size
  establishment.prob=NA # Probability of establishment
)
```

### Survival

```{r}
surv.reg=glm(surv ~ size, data=d, family=binomial())
summary(surv.reg)

params$surv.int=coefficients(surv.reg)[1]
params$surv.slope=coefficients(surv.reg)[2]
```

### Growth

```{r}
growth.reg=lm(sizeNext ~ size, data=d)
summary(growth.reg)

params$growth.int=coefficients(growth.reg)[1]
params$growth.slope=coefficients(growth.reg)[2]
params$growth.sd=sd(resid(growth.reg))
```

### Seed production

```{r}
seed.reg=glm(fec.seed ~ size, data=d, family=poisson())
summary(seed.reg)
params$seed.int=coefficients(seed.reg)[1]
params$seed.slope=coefficients(seed.reg)[2]
```

### Size of recruits

```{r}
params$recruit.size.mean=mean(d$sizeNext[is.na(d$size)])
params$recruit.size.sd=sd(d$sizeNext[is.na(d$size)])
```

### Establishment

```{r}
params$establishment.prob=sum(is.na(d$size))/sum(d$fec.seed,na.rm=TRUE)
```

## Now check models to original data

THIS IS CRITICAL!


```{r}
p1.fit <- p1 + geom_function(fun = ~ predict(surv.reg,
                                             data.frame(size=.x), 
                                             type = "response"),
                             color = "red", lwd=1)

p2.fit <- p2  + geom_function(fun = ~ predict(growth.reg,
                                              data.frame(size=.x)),
                              color = "red", lwd = 1)

p3.fit <- p3 + geom_function(fun = ~ predict(seed.reg,
                                             data.frame(size=.x),
                                             type="response"),
                             color = "red",lwd=1)

p4.fit <- p4 + geom_function(fun = ~ dnorm(.x,params$recruit.size.mean, params$recruit.size.sd),
                             color = "red", lwd=1)

(p1.fit + p2.fit) / (p3.fit + p4.fit)

```

## Define functions to describe life history

unannotated
```{r, eval=FALSE}
# 1. survival probability function
s.x=function(x,params) {
  u=exp(params$surv.int+params$surv.slope*x)
  return(u/(1+u))
}

# 2. growth function
g.yx=function(xp,x,params) {
  dnorm(xp,mean=params$growth.int+params$growth.slope*x,sd=params$growth.sd)
}

# 3. reproduction function "expected offspring"
f.yx=function(xp,x,params) {
  params$establishment.prob*
    dnorm(xp,mean=params$recruit.size.mean,sd=params$recruit.size.sd)*
    exp(params$seed.int+params$seed.slope*x) # why exp()?
}
```

```{r}
# 1. survival probability function
## JM. This is inverse logit.  Provides survival probability based on size.
s.x=function(x,params) {
  u=exp(params$surv.int+params$surv.slope*x)
  return(u/(1+u))
}

# 2. growth function
## JM.  return a probablity distribution of new sizes at t+1 (xp) at a given size x.  
g.yx=function(xp,x,params) {
  dnorm(xp,mean=params$growth.int+params$growth.slope*x,sd=params$growth.sd)
}

# 3. reproduction function
## JM.  exp becauase the original was Poisson and that is fit with a log link.
f.yx=function(xp,x,params) {
  params$establishment.prob*
    dnorm(xp,mean=params$recruit.size.mean,sd=params$recruit.size.sd)*
    exp(params$seed.int+params$seed.slope*x) # why exp()?
}
```

## Make a kernel

Define the parameters for integration
```{r}
# the sizes we are integrating over
min.size=.9*min(c(d$size,d$sizeNext),na.rm=T)
max.size=1.1*max(c(d$size,d$sizeNext),na.rm=T)

n=100 # dimensions of the matrix

b=min.size+c(0:n)*(max.size-min.size)/n # boundary points
y=0.5*(b[1:n]+b[2:(n+1)]) # mesh points
h=y[2]-y[1] # step size
```

Illustration of `outer`
```{r}
x.test <- 1:5 %>% set_names(1:5)
y.test <- 11:15 %>% set_names(11:15)
outer(x.test, y.test, FUN = `+`)
```
```{r}
head(y)
range(y)
h
```


Make the matrix
```{r}
G=h*outer(y,y,g.yx,params=params) # growth matrix
# OK for each value of y,y evaluate the growth function g.yx using params
# If I understand this correctly, the rows are x(t) and the columns are x(t+1)

S=s.x(y,params=params) # survival
F=h*outer(y,y,f.yx,params=params) # reproduction matrix
P=G # placeholder; redefine P on the next line
for(i in 1:n) P[,i]=G[,i]*S[i] # growth/survival matrix
K=P+F # full matrix
```

### Basic Analyses

```{r}
(lam <- Re(eigen(K)$values[1])) #dominant eigen value, lam = asympototic growth rate

w.eigen <- Re(eigen(K)$vectors[,1]) # right eigen vector, stable stage (e.g. size) distribution
stable.dist <- w.eigen/sum(w.eigen) # normalized it

v.eigen <- Re(eigen(t(K))$vectors[,1]) # left eigen vector, stable reproductive value
repro.val <- v.eigen/v.eigen[1] #I don't understand this normalization

# The eigen-things can be combined to obtain the sensitivity and elasticity matrices.
v.dot.w=sum(stable.dist*repro.val)*h
sens=outer(repro.val,stable.dist)/v.dot.w
elas=matrix(as.vector(sens)*as.vector(K)/lam,nrow=n)
```

Plot it
```{r, fig.height=10}

pl11 <- K %>% as_tibble() %>%
  set_colnames(y) %>% 
  mutate(size.t1=y) %>%
  pivot_longer(-size.t1, names_to = "size.t", names_transform = as.numeric) %>%
  ggplot(aes(x=size.t, y = size.t1)) +
  geom_raster(aes(fill = value)) +
  geom_contour(aes(z = value),lwd=.25) +
  scale_fill_viridis_c(option = "plasma") +
  labs(x = "Size (t)", y = "Size (t + 1)", title = "IPM Matrix") +
  coord_equal() +
  theme_bw()

pl12 <- tibble(Size = y, Density = stable.dist) %>% 
  ggplot(aes(x=Size, y = Density)) +
  geom_line() +
  labs(x = "Size", y = "Density", title = "Stable size distribution")

pl13 <- tibble(Size = y, repro.val) %>%
  ggplot(aes(x=Size, y = repro.val)) +
  geom_line() +
  labs(y = "Reproductive Values", title = "Reproductive Values")

pl14 <- elas %>% as_tibble() %>%
  set_colnames(y) %>% 
  mutate(size.t1=y) %>%
  pivot_longer(-size.t1, names_to = "size.t", names_transform = as.numeric) %>%
  ggplot(aes(x=size.t, y = size.t1)) +
  geom_raster(aes(fill = value)) +
  scale_fill_viridis_c(option = "plasma") +
  labs(x = "Size (t)", y = "Size (t + 1)", title = "Elasticity") +
  coord_equal() +
  theme_bw()

pl15 <- sens %>% as_tibble() %>%
  set_colnames(y) %>% 
  mutate(size.t1=y) %>%
  pivot_longer(-size.t1, names_to = "size.t", names_transform = as.numeric) %>%
  ggplot(aes(x=size.t, y = size.t1)) +
  geom_raster(aes(fill = value)) +
  scale_fill_viridis_c(option = "plasma") +
  labs(x = "Size (t)", y = "Size (t + 1)", title = "Sensitivity") +
  coord_equal() +
  theme_bw()

(pl11 + pl14 + pl15)  / (pl12 + pl13 + plot_spacer()) 
```


## Improving the model

### Model variance in size as a function of size

Because variance isn't constant.

This is illustrated by plotting residuals vs size.

```{r}
plot(growth.reg$model$size,abs(resid(growth.reg)),
     xlab="size",ylab="residual")
```

Use `nlme::gls()` to model variance as a function of size

```{r}
growth.reg=gls(sizeNext ~ size, weights = varExp(), na.action = na.omit, data = d)
summary(growth.reg)
```
plot the model
```{r}
p2  + geom_function(fun = ~ predict(growth.reg,
                                    data.frame(size=.x)),
                    color = "red", lwd = 1)
```

Now update the parameters, growth function, and then recreate the IPM

```{r}
params$growth.int=coefficients(growth.reg)[1]
params$growth.slope=coefficients(growth.reg)[2]
params$growth.sigma2=summary(growth.reg)$sigma^2
params$growth.sigma2.exp=as.numeric(growth.reg$modelStruct$varStruct)

g.yx=function(xp,x,params) {
  dnorm(xp,mean=params$growth.int+params$growth.slope*x,
        sd=sqrt(params$growth.sigma2*exp(2*params$growth.sigma2.exp*x)))
}
```

Make the matrix
```{r}
G=h*outer(y,y,g.yx,params=params) # growth matrix
# OK for each value of y,y evaluate the growth function g.yx using params
# If I understand this correctly, the rows are x(t) and the columns are x(t+1)

S=s.x(y,params=params) # survival
F=h*outer(y,y,f.yx,params=params) # reproduction matrix
P=G # placeholder; redefine P on the next line
for(i in 1:n) P[,i]=G[,i]*S[i] # growth/survival matrix
K=P+F # full matrix
```

### Basic Analyses

```{r}
(lam <- Re(eigen(K)$values[1])) #dominant eigen value, lam = asympototic growth rate

w.eigen <- Re(eigen(K)$vectors[,1]) # right eigen vector, stable stage (e.g. size) distribution
stable.dist <- w.eigen/sum(w.eigen) # normalized it

v.eigen <- Re(eigen(t(K))$vectors[,1]) # left eigen vector, stable reproductive value
repro.val <- v.eigen/v.eigen[1] #I don't understand this normalization

# The eigen-things can be combined to obtain the sensitivity and elasticity matrices.
v.dot.w=sum(stable.dist*repro.val)*h
sens=outer(repro.val,stable.dist)/v.dot.w
elas=matrix(as.vector(sens)*as.vector(K)/lam,nrow=n)
```

Plot it
```{r, fig.height=10}

pl11 <- K %>% as_tibble() %>%
  set_colnames(y) %>% 
  mutate(size.t1=y) %>%
  pivot_longer(-size.t1, names_to = "size.t", names_transform = as.numeric) %>%
  ggplot(aes(x=size.t, y = size.t1)) +
  geom_raster(aes(fill = value)) +
  geom_contour(aes(z = value),lwd=.25) +
  scale_fill_viridis_c(option = "plasma") +
  labs(x = "Size (t)", y = "Size (t + 1)", title = "IPM Matrix") +
  coord_equal() +
  theme_bw()

pl12 <- tibble(Size = y, Density = stable.dist) %>% 
  ggplot(aes(x=Size, y = Density)) +
  geom_line() +
  labs(x = "Size", y = "Density", title = "Stable size distribution")

pl13 <- tibble(Size = y, repro.val) %>%
  ggplot(aes(x=Size, y = repro.val)) +
  geom_line() +
  labs(y = "Reproductive Values", title = "Reproductive Values")

pl14 <- elas %>% as_tibble() %>%
  set_colnames(y) %>% 
  mutate(size.t1=y) %>%
  pivot_longer(-size.t1, names_to = "size.t", names_transform = as.numeric) %>%
  ggplot(aes(x=size.t, y = size.t1)) +
  geom_raster(aes(fill = value)) +
  scale_fill_viridis_c(option = "plasma") +
  labs(x = "Size (t)", y = "Size (t + 1)", title = "Elasticity") +
  coord_equal() +
  theme_bw()

pl15 <- sens %>% as_tibble() %>%
  set_colnames(y) %>% 
  mutate(size.t1=y) %>%
  pivot_longer(-size.t1, names_to = "size.t", names_transform = as.numeric) %>%
  ggplot(aes(x=size.t, y = size.t1)) +
  geom_raster(aes(fill = value)) +
  scale_fill_viridis_c(option = "plasma") +
  labs(x = "Size (t)", y = "Size (t + 1)", title = "Sensitivity") +
  coord_equal() +
  theme_bw()

(pl11 + pl14 + pl15)  / (pl12 + pl13 + plot_spacer()) 
```

## Eviction

Evicition occurs when some individuals grow larger than the model's parameter space allows.

Check for eviction:

```{r}
plot(y,s.x(y,params),xlab="Size",type="l", ylab="Survival Probability",lwd=2)
points(y,apply(P,2,sum),col="red",lwd=3,cex=.1,pch=19)
```

### Recreate the matrices while capturing evicted individuals

```{r}
G=h*outer(y,y,g.yx,params=params) # growth matrix
S=s.x(y,params=params)
P=G
# fix eviction of offspring
for(i in 1:(n/2)) {
  G[1,i]<-G[1,i]+1-sum(G[,i])
  P[,i]<-G[,i]*S[i]
}
# fix eviction of large adults
for(i in (n/2+1):n) {
  G[n,i]<-G[n,i]+1-sum(G[,i])
  P[,i]<-G[,i]*S[i]
}
F=h*outer(y,y,f.yx,params=params) # reproduction matrix
K=P+F # full matrix
(lam=Re(eigen(K)$values[1])) # new population growth rate
```

And now plot again
```{r}
plot(y,s.x(y,params),xlab="Size",type="l", ylab="Survival Probability",lwd=2)
points(y,apply(P,2,sum),col="red",lwd=3,cex=.1,pch=19)
```

### Quadratic growth term

```{r}
growth.reg=gls(sizeNext~size+I(size^2),weights=varExp(), na.action=na.omit, data=d)
summary(growth.reg) 
```

```{r}
params$growth.int=coefficients(growth.reg)[1]
params$growth.slope=coefficients(growth.reg)[2]
params$growth.sqrd=coefficients(growth.reg)[3]
params$growth.sigma2=summary(growth.reg)$sigma^2
params$growth.sigma2.exp=as.numeric(growth.reg$modelStruct$varStruct)
g.yx=function(xp,x,params) {
  dnorm(xp,
        mean=params$growth.int+params$growth.slope*x+params$growth.sqrd*x^2,
        sd=sqrt(params$growth.sigma2*exp(2*params$growth.sigma2.exp*x)))
}

G=h*outer(y,y,g.yx,params=params) # growth matrix
S=s.x(y,params=params) # survival
P=G # placeholder; redefine P on the next line
# fix eviction of offspring
for(i in 1:(n/2)) {
  G[1,i]<-G[1,i]+1-sum(G[,i])
  P[,i]<-G[,i]*S[i]
}
# fix eviction of large adults
for(i in (n/2+1):n) {
  G[n,i]<-G[n,i]+1-sum(G[,i])
  P[,i]<-G[,i]*S[i]
}
F=h*outer(y,y,f.yx,params=params) # reproduction matrix
K=P+F # full matrix
(lam=Re(eigen(K)$values[1])) # new population growth rate
w.eigen=Re(eigen(K)$vectors[,1])
stable.dist=w.eigen/sum(w.eigen)
v.eigen=Re(eigen(t(K))$vectors[,1])
repro.val=v.eigen/v.eigen[1]
v.dot.w=sum(stable.dist*repro.val)*h
sens=outer(repro.val,stable.dist)/v.dot.w
elas=matrix(as.vector(sens)*as.vector(K)/lam,nrow=n)
```

```{r, fig.height=10}

pl11 <- K %>% as_tibble() %>%
  set_colnames(y) %>% 
  mutate(size.t1=y) %>%
  pivot_longer(-size.t1, names_to = "size.t", names_transform = as.numeric) %>%
  ggplot(aes(x=size.t, y = size.t1)) +
  geom_raster(aes(fill = value)) +
  geom_contour(aes(z = value),lwd=.25) +
  scale_fill_viridis_c(option = "plasma") +
  labs(x = "Size (t)", y = "Size (t + 1)", title = "IPM Matrix") +
  coord_equal() +
  theme_bw()

pl12 <- tibble(Size = y, Density = stable.dist) %>% 
  ggplot(aes(x=Size, y = Density)) +
  geom_line() +
  labs(x = "Size", y = "Density", title = "Stable size distribution")

pl13 <- tibble(Size = y, repro.val) %>%
  ggplot(aes(x=Size, y = repro.val)) +
  geom_line() +
  labs(y = "Reproductive Values", title = "Reproductive Values")

pl14 <- elas %>% as_tibble() %>%
  set_colnames(y) %>% 
  mutate(size.t1=y) %>%
  pivot_longer(-size.t1, names_to = "size.t", names_transform = as.numeric) %>%
  ggplot(aes(x=size.t, y = size.t1)) +
  geom_raster(aes(fill = value)) +
  scale_fill_viridis_c(option = "plasma") +
  labs(x = "Size (t)", y = "Size (t + 1)", title = "Elasticity") +
  coord_equal() +
  theme_bw()

pl15 <- sens %>% as_tibble() %>%
  set_colnames(y) %>% 
  mutate(size.t1=y) %>%
  pivot_longer(-size.t1, names_to = "size.t", names_transform = as.numeric) %>%
  ggplot(aes(x=size.t, y = size.t1)) +
  geom_raster(aes(fill = value)) +
  scale_fill_viridis_c(option = "plasma") +
  labs(x = "Size (t)", y = "Size (t + 1)", title = "Sensitivity") +
  coord_equal() +
  theme_bw()

(pl11 + pl14 + pl15)  / (pl12 + pl13 + plot_spacer()) 
```

## Flowering probability

Update to include a probability of flowering based on size, instead of assuming that all plants flower.

```{r}
# flowering probability function

p.flower.x=function(x,params) {
  u=exp(params$flower.int+params$flower.slope*x)
  return(u/(1+u))
}

# Modified reproduction function
f.yx=function(xp,x,params) {
  p.flower.x(x,params) * 
    params$establishment.prob * 
    dnorm(xp,mean=params$recruit.size.mean,sd=params$recruit.size.sd) * 
    exp(params$seed.int+params$seed.slope*x)
}

# Fit the flowering regression model
flower.reg=glm(fec.flower~size,data=d,family=binomial())
summary(flower.reg)

# Get the flowering parameters

params$flower.int=coefficients(flower.reg)[1]
params$flower.slope=coefficients(flower.reg)[2]

```
update seed regression to only fit plants that flowered and get the parameters

```{r}
seed.reg=glm(fec.seed~size,data=d[d$fec.flower==1,],family=poisson())
summary(seed.reg)

params$seed.int=coefficients(seed.reg)[1]
params$seed.slope=coefficients(seed.reg)[2]
```
Finally recalculate the matrices and plot

```{r}
G=h*outer(y,y,g.yx,params=params) # growth matrix
S=s.x(y,params=params) # survival
P=G # placeholder; redefine P on the next line
# fix eviction of offspring
for(i in 1:(n/2)) {
  G[1,i]<-G[1,i]+1-sum(G[,i])
  P[,i]<-G[,i]*S[i]
}
# fix eviction of large adults
for(i in (n/2+1):n) {
  G[n,i]<-G[n,i]+1-sum(G[,i])
  P[,i]<-G[,i]*S[i]
}
# for(i in 1:n) P[,i]=G[,i]*S[i] # growth/survival matrix

F=h*outer(y,y,f.yx,params=params) # reproduction matrix

K=P+F # full matrix
(lam=Re(eigen(K)$values[1])) # new population growth rate
w.eigen=Re(eigen(K)$vectors[,1])
stable.dist=w.eigen/sum(w.eigen)
v.eigen=Re(eigen(t(K))$vectors[,1])
repro.val=v.eigen/v.eigen[1]
v.dot.w=sum(stable.dist*repro.val)*h
sens=outer(repro.val,stable.dist)/v.dot.w
elas=matrix(as.vector(sens)*as.vector(K)/lam,nrow=n)
```

```{r, fig.height=10}

pl11 <- K %>% as_tibble() %>%
  set_colnames(y) %>% 
  mutate(size.t1=y) %>%
  pivot_longer(-size.t1, names_to = "size.t", names_transform = as.numeric) %>%
  ggplot(aes(x=size.t, y = size.t1)) +
  geom_raster(aes(fill = value)) +
  geom_contour(aes(z = value),lwd=.25) +
  scale_fill_viridis_c(option = "plasma") +
  labs(x = "Size (t)", y = "Size (t + 1)", title = "IPM Matrix") +
  coord_equal() +
  theme_bw()

pl12 <- tibble(Size = y, Density = stable.dist) %>% 
  ggplot(aes(x=Size, y = Density)) +
  geom_line() +
  labs(x = "Size", y = "Density", title = "Stable size distribution")

pl13 <- tibble(Size = y, repro.val) %>%
  ggplot(aes(x=Size, y = repro.val)) +
  geom_line() +
  labs(y = "Reproductive Values", title = "Reproductive Values")

pl14 <- elas %>% as_tibble() %>%
  set_colnames(y) %>% 
  mutate(size.t1=y) %>%
  pivot_longer(-size.t1, names_to = "size.t", names_transform = as.numeric) %>%
  ggplot(aes(x=size.t, y = size.t1)) +
  geom_raster(aes(fill = value)) +
  scale_fill_viridis_c(option = "plasma") +
  labs(x = "Size (t)", y = "Size (t + 1)", title = "Elasticity") +
  coord_equal() +
  theme_bw()

pl15 <- sens %>% as_tibble() %>%
  set_colnames(y) %>% 
  mutate(size.t1=y) %>%
  pivot_longer(-size.t1, names_to = "size.t", names_transform = as.numeric) %>%
  ggplot(aes(x=size.t, y = size.t1)) +
  geom_raster(aes(fill = value)) +
  scale_fill_viridis_c(option = "plasma") +
  labs(x = "Size (t)", y = "Size (t + 1)", title = "Sensitivity") +
  coord_equal() +
  theme_bw()

p16 <- tibble(size=y) %>%
  mutate(flowering.probability = predict(flower.reg, newdata = ., type = "response")) %>%
  ggplot(aes(x=size, y = flowering.probability)) +
  geom_line() +
  labs(x="Size (t)", y = "Flowering probability")

(pl11 + pl14 + pl15)  / (pl12 + pl13 + p16) 
```

## Diagnostics

### passage times.
Start with life expectancy

```{r}
Pmat = new("IPMmatrix", nDiscrete = 0, nEnvClass = 0, nBigMatrix = n, nrow = n, ncol = n, meshpoints = y, env.index = 0, names.discrete = "")
Pmat[, ] = P
str(Pmat)
```
```{r}
(mle=meanLifeExpect(Pmat))
```

plot the results

```{r}
plot(y,meanLifeExpect(Pmat), xlab="Size (t)",ylab="Time")
```

### Maximum size

```{r fig.asp=2}
par(mfrow=c(2,1),mar=c(4,5,2,2))
image(y,y,t(P), xlab="Size (t)",ylab="Size (t+1)",
           col=topo.colors(100), main="IPM matrix")
contour(y,y,t(P), add = TRUE, drawlabels = TRUE)
abline(0,1,lwd=3,lty=2)
plot(density(d$sizeNext[!is.na(d$sizeNext)]),xlab="Size(t+1)",
     main="Observed distribution of sizes")
```
### Transient dynamics

IF the population is not at equilibrium what happens...

Asymptotic growth rate
```{r}
(lam=Re(eigen(K)$values[1])) # asymptotic growth rate
```

Damping ratio (ratio of dominant eigen value to the second largest one)
(what does the second largest eigen value represent?)
```{r}
(damp=Re(eigen(K)$values[1])/Re(eigen(K)$values[2])) # damping ratio
```

Create a random starting structure and run it forward for 20 years to see when it achieves the asymptotic growth rate.
(Seems like you would want to do this a bunch of times if you cared)
```{r}
initial.pop=runif(100) #random starting population structure
initial.pop=initial.pop/sum(initial.pop)
nyears=20
size.dist=matrix(NA,n,nyears)
lambda=rep(NA,nyears)
xold=initial.pop
for(i in 1:nyears){
  xnew=K%*%xold
  lambda[i]=sum(xnew)/sum(xold)
  size.dist[,i]=xnew
  xold=xnew
}
lambda
```

