---
title: "WL2023_Size_Growth"
author: "Julin Maloof"
date: "2025-01-24"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Intitial attempt at making a growth/survival matrix from WL2 2023, now using random effects

```{r}
library(tidyverse)
library(tidymodels)
library(multilevelmod)
library(lmerTest)
library(broom.mixed)
library(doMC)
library(furrr)
library(modelr)
registerDoMC(cores = 7)
conflicted::conflict_prefer("select", "dplyr")
conflicted::conflict_prefer("filter", "dplyr")
conflicted::conflict_prefer("lag", "dplyr")
```

## Import the data

```{r}
growth <- read_csv("../input/WL2-2023_Size_Combined.csv")
growth
```

```{r}
survival <- read_csv("../input/WL2_Mortality.csv")
survival
```

## Wrangle Growth

We want to predict size next from size.  In the Merow examples there was a column "size" and "size next" for each time step.  I guess we are going to need to do that here.

We also should compute the interval length 

```{r}
growth2 <- growth %>% arrange(Genotype, survey_date) %>%
  select(-survey.notes, -long.leaf.cm) %>%
  group_by(Genotype) %>%
  mutate(height.cm.next = lead(height.cm),
         elapsed_days= lead(survey_date) - survey_date,
         elapsed_weeks = as.integer(elapsed_days)/7,
         survey_date_char = as.character(survey_date)) %>%
  ungroup() %>%
  drop_na(block, height.cm, height.cm.next)

growth2 
```

For easy visualization lets subset just to elapsed_days of 7

```{r}
growth2 %>%
  filter(elapsed_days==7) %>%
  ggplot(aes(x=height.cm, y=height.cm.next)) +
  geom_point(aes(color=parent.pop), alpha = 0.3) +
  geom_abline() +
  geom_smooth(method="lm")
```

Do some model comparison

Define the formula


```{r}
minSize <- min(growth2$height.cm, na.rm = TRUE)
maxSize <- max(growth2$height.cm, na.rm = TRUE)
# newdata = tibble(height.cm=seq(minSize, maxSize, length.out=101), elapsed_weeks = 1)
```


In order to be able to use the tidymodel tools on mixed effect models, it works best to add variables and formulas in the following way.  This is a bit cumbersome, but it will help later.

CURRENTLY NONE OF THE GROWTH MODELS ARE TESTING TIME AS A FACTOR (SURV_DATE_CHAR).  IT COULD BE WORTH GOING BACK AND TESTING THAT.
```{r}
lmer.spec <- linear_reg() %>%
  set_engine("lmer")

growth.wflow <- workflow() %>%
  add_variables(outcomes = height.cm.next, predictors = c(height.cm, elapsed_weeks, parent.pop, mf, block)) 

growth.models <- tibble(wflow=list(
  m1_block = {growth.wflow %>%
      add_model(lmer.spec,
                formula = height.cm.next ~ (1|block)) },
  
  m2_parent = { growth.wflow %>% 
      add_model(lmer.spec,
                formula = height.cm.next ~ (1|parent.pop) )},
  
  m3_block.parent = { growth.wflow %>% 
      add_model(lmer.spec,
                formula = height.cm.next ~ (1|parent.pop) + (1|block) )},
  
  m4_block.parent.mf = { growth.wflow %>% 
      add_model(lmer.spec,
                formula = height.cm.next ~ (1|parent.pop/mf) + (1|block) )},
  
  m5_linear.parent.mf.block = { growth.wflow %>% 
      add_model(lmer.spec,
                formula = height.cm.next ~ height.cm + (1|parent.pop/mf) + (1|block) )},
  
  m6_linear.parent.mf.block.slope = { growth.wflow %>% 
      add_model(lmer.spec,
                formula = height.cm.next ~ height.cm + (height.cm|parent.pop/mf) + (1|block) )},
  
  m7_linear.parent.mf.block.slope.slope = { growth.wflow %>% 
      add_model(lmer.spec,
                formula = height.cm.next ~ height.cm + (height.cm|parent.pop/mf) + (height.cm|block) )},
  
  # adding this after looking at m7...intercept and slope for mf have a correlation of -1.00, so remove slope from mf.
  m7_simple.linear.parent.mf.block.slope.slope = { growth.wflow %>% 
      add_model(lmer.spec,
                formula = height.cm.next ~ height.cm + (height.cm|parent.pop) + (height.cm|block) + (1|parent.pop/mf) )},
  
  m8_linear_with_weeks_int.parent = { growth.wflow %>% 
      add_model(lmer.spec,
                formula = height.cm.next ~ height.cm*elapsed_weeks + (1|parent.pop/mf) + (1|block) )},
  
  m9_linear_with_weeks_int.parent.all.slope = { growth.wflow %>% 
      add_model(lmer.spec,
                formula = height.cm.next ~ height.cm*elapsed_weeks + (height.cm + elapsed_weeks|parent.pop/mf) + (height.cm + elapsed_weeks|block) )},
  
  m10_quadratic.parent.mf.block = { growth.wflow %>% 
      add_model(lmer.spec,
                formula = height.cm.next ~ height.cm + I(height.cm^2) + (1|parent.pop/mf) + (1|block) )},
  
  m11_cubic.parent.mf.block = { growth.wflow %>% 
      add_model(lmer.spec,
                formula = height.cm.next ~ height.cm + I(height.cm^2) + I(height.cm^3) + (1|parent.pop/mf) + (1|block) )},
  
  m12_quadratic_weeks.parent.height.slope = { growth.wflow %>% 
      add_model(lmer.spec,
                formula = height.cm.next ~ height.cm*elapsed_weeks*parent.pop + I(height.cm^2) + (height.cm + I(height.cm^2) | parent.pop / mf) + (height.cm + I(height.cm^2) | block) )}, 
  
  m13_cubic_weeks.parent.height.slope = { growth.wflow %>% 
      add_model(lmer.spec,
                formula = height.cm.next ~ height.cm*elapsed_weeks*parent.pop + I(height.cm^2) + I(height.cm^3) + (height.cm + I(height.cm^2) | parent.pop) + (height.cm + I(height.cm^2) | block) )}
),
name = names(wflow)
)
```

Fit the models

```{r}
system.time( { # 55 seconds
  growth.models <- growth.models %>%
    mutate(fit = map(wflow, fit, data = growth2),
           glance = map(fit, glance)
    ) 
})

growth.models %>% select(-wflow, -fit) %>% unnest(glance) %>% arrange(BIC)
```

```{r}
growth.models %>% filter(name=="m7_linear.parent.mf.block.slope.slope") %>%
  pull(fit) %>%
  magrittr::extract2(1) %>%
  tidy()
```

```{r}
growth.models %>% filter(name=="m7_simple.linear.parent.mf.block.slope.slope") %>%
  pull(fit) %>%
  magrittr::extract2(1) %>%
  tidy()
```

```{r}
growth.models %>% filter(name=="m9_linear_with_weeks_int.parent.all.slope") %>%
  pull(fit) %>%
  magrittr::extract2(1) %>%
  tidy()
```

Concentrate on the five best models and do some cross-validation to better assess predictive power
```{r}
growth.models.best <- growth.models %>%
  unnest(glance) %>%
  slice_min(AIC, n=5) 

set.seed(1001)
growth_folds <- vfold_cv(growth2, v = 10)

growth.models.best <- growth.models.best %>%
  mutate(resamples = map(wflow, fit_resamples, resamples = growth_folds, control = control_resamples(save_pred = TRUE)))
```

```{r}
growth.models.best %>% 
  mutate(metrics = map(resamples, collect_metrics, type = "wide")) %>%
  select(name, metrics) %>%
  unnest(metrics) %>%
  select(-.config) %>%
  arrange(rmse)

```

```{r}
growth.models.best %>% 
  mutate(metrics = map(resamples, collect_metrics, type = "long")) %>%
  select(name, metrics) %>%
  unnest(metrics) %>%
  arrange(.metric, mean) %>%
  select(name, .metric, mean, std_err)

```

```{r}
growth.models.best %>% 
  mutate(metrics = map(resamples, collect_metrics, type = "long")) %>%
  select(name, metrics) %>%
  unnest(metrics) %>%
  arrange(.metric, mean) %>%
  select(name, .metric, mean, std_err) %>%
  ggplot(aes(x=name, y = mean, ymin=mean-std_err, ymax=mean+std_err, fill = name)) +
  geom_col() +
  geom_errorbar(width = 0.5) +
  facet_wrap(~.metric, scales = "free_y") +
  theme(axis.text.x = element_blank(), axis.title.x = element_blank()) +
  ggtitle("10-fold cross validation") +
  scale_fill_viridis_d()
```

So, we go with simplified m7 (or possibly m9 if we decide we need weeks).
```{r}
growth.model.final <- growth.models.best %>% filter(name == "m7_simple.linear.parent.mf.block.slope.slope") %>% pull(fit) %>% magrittr::extract2(1)

growth.model.final
```

For plotting the prediction curve, it looks like I can't combine `geom_function` and `facet_wrap` so I will compute the prediction curves myself and then combine in the plots

```{r}
newdata.growth <- growth2 %>%
  group_by(parent.pop) %>%
  summarize(minSize = min(height.cm, na.rm = TRUE),
            maxSize = max(height.cm, na.rm = TRUE)) %>%
  
  # generate a "newdata" data frame for each parent.pop
  mutate(newdata = pmap(list(parent.pop, minSize, maxSize), \(p, x, y) tibble(parent.pop = p,
                                                                              elapsed_weeks = 1,
                                                                              height.cm = seq(x, y, length.out = 101)))) %>%
  select(newdata) %>%
  unnest(newdata)

growth.predictions <- extract_fit_engine(growth.model.final) %>%
  predict(newdata.growth, re.form = ~ (height.cm | parent.pop)) %>%
  cbind(newdata.growth, height.cm.next=.)
```

Plot it.  Points are actual data (height.next vs height).  The line is the model prediction.
```{r, fig.asp=1}
growth2 %>%
  ggplot(aes(x=height.cm, y = height.cm.next)) +
  geom_point(alpha=.25) +
  geom_line(data = growth.predictions, color = "blue") +
  facet_wrap(~parent.pop, scales = "free")
```



## On to survival

Since we want to predict survival as a function of size, we need to combine the data sets

```{r}
survival2 <- survival %>% select(Genotype, death.date) %>%
  right_join(growth) %>%
  mutate(ppmf = str_c(parent.pop, "_", mf), # allows me to specify RE for pp and mf separately.
         death.date = lubridate::mdy(death.date),
         surv = ifelse(is.na(death.date), 1, death.date > survey_date),
         survey_date_char = as.character(survey_date),
         elapsed_days= lead(survey_date) - survey_date,
         elapsed_weeks = as.integer(elapsed_days)/7,) %>%
  
  # because we don't have height measurement on the day death is observed we need a new column, height previous
  arrange(Genotype, survey_date) %>% 
  group_by(Genotype) %>%
  mutate(height.cm.prev = lag(height.cm)) %>%
  drop_na(surv, parent.pop, height.cm.prev)

# check it
survival2 %>% select(Genotype, death.date, survey_date, surv, height.cm, height.cm.prev)
```

```{r}
surv.spec <- linear_reg() %>%
  set_engine("glmer", family = "binomial")

surv.wflow <- workflow() %>%
  add_variables(outcomes = surv, predictors = c(height.cm.prev, survey_date_char, elapsed_weeks, parent.pop, mf, ppmf, block)) 

surv.models <- tibble(wflow=list(
  m1_RE = {surv.wflow %>%
      add_model(surv.spec,
                formula = surv ~ (1|parent.pop/mf) + (1|block)) },
  
  m1b_RE2 = {surv.wflow %>% # reality check; this should be the same as M1
      add_model(surv.spec,
                formula = surv ~ (1|parent.pop) + (1|ppmf) + (1|block)) },
  
  m2_linear = {surv.wflow %>%
      add_model(surv.spec,
                formula = surv ~ height.cm.prev + (1|parent.pop/mf) + (1|block)) },
  
  m2d_linear.date = {surv.wflow %>%
      add_model(surv.spec,
                formula = surv ~ height.cm.prev + survey_date_char + (1|parent.pop/mf) + (1|block)) },
  
  m2e_linear.weeks = {surv.wflow %>%
      add_model(surv.spec,
                formula = surv ~ height.cm.prev + elapsed_weeks + (1|parent.pop/mf) + (1|block)) },
  
  m3_linear.slope.parent.pop = {surv.wflow %>%
      add_model(surv.spec,
                formula = surv ~ height.cm.prev + (height.cm.prev|parent.pop) + (1|ppmf) + (1|block)) },
  
  m5_linear.slope.parent.pop.mf = {surv.wflow %>%
      add_model(surv.spec,
                formula = surv ~ height.cm.prev + (height.cm.prev|parent.pop/mf) + (1|block)) },
  
  m6_linear.slope.parent.pop.mf.block = {surv.wflow %>%
      add_model(surv.spec,
                formula = surv ~ height.cm.prev + (height.cm.prev|parent.pop/mf) + (height.cm.prev|block)) },
  
  #Too Slow when including date as a slope for RE, so remove that
  # Still does not converge
  # m6d_linear.date.slope.parent.pop.mf.block = {surv.wflow %>%
  #     add_model(surv.spec,
  #               formula = surv ~ height.cm.prev + survey_date_char + (height.cm.prev|parent.pop/mf) + (height.cm.prev|block)) },
  # 
  # m6e_linear.weeks.slope.parent.pop.mf.block = {surv.wflow %>%
  #     add_model(surv.spec,
  #               formula = surv ~ height.cm.prev + elapsed_weeks + (elapsed_weeks + height.cm.prev|parent.pop/mf) + (height.cm.prev|block)) },
  
  # Does not converge
  #   m7_quad = {surv.wflow %>%
  #     add_model(surv.spec,
  #               formula = surv ~ height.cm.prev + I(height.cm.prev^2) + (1|parent.pop/mf) + (1|block)) },
  # 
  # m7d_quad.date = {surv.wflow %>%
  #     add_model(surv.spec,
  #               formula = surv ~ height.cm.prev + survey_date_char + I(height.cm.prev^2) + (1|parent.pop/mf) + (1|block)) },
  # 
  # m7e_quad.weeks = {surv.wflow %>%
  #     add_model(surv.spec,
  #               formula = surv ~ height.cm.prev + elapsed_weeks + I(height.cm.prev^2) + (1|parent.pop/mf) + (1|block)) },
  
  m8_quad.slope.parent.pop = {surv.wflow %>%
      add_model(surv.spec,
                formula = surv ~ height.cm.prev + I(height.cm.prev^2) + (height.cm.prev + I(height.cm.prev^2)|parent.pop) + (1|ppmf) + (1|block)) },
  
  # Fails to converge
  # m9_quad.slope.parent.pop.mf = {surv.wflow %>%
  #     add_model(surv.spec,
  #               formula = surv ~ height.cm.prev + I(height.cm.prev^2) + (height.cm.prev + I(height.cm.prev^2)|parent.pop/mf) + (1|block)) },
  
  m10_quad.slope.parent.pop.mf.blck = {surv.wflow %>%
      add_model(surv.spec,
                formula = surv ~ height.cm.prev + I(height.cm.prev^2) + (height.cm.prev + I(height.cm.prev^2)|parent.pop/mf) + (height.cm.prev + I(height.cm.prev^2)|block)) },
  
  # Keep date out of RE so it will fit
  # Still fails to converge
  # m10d_quad.date.slope.parent.pop.mf.blck = {surv.wflow %>%
  #     add_model(surv.spec,
  #               formula = surv ~ height.cm.prev + survey_date_char + I(height.cm.prev^2) + (height.cm.prev + I(height.cm.prev^2)|parent.pop/mf) + (height.cm.prev + I(height.cm.prev^2)|block)) },
  # 
  # m10e_quad.weeks.slope.parent.pop.mf.blck = {surv.wflow %>%
  #     add_model(surv.spec,
  #               formula = surv ~ height.cm.prev + elapsed_weeks + I(height.cm.prev^2) + (height.cm.prev + I(height.cm.prev^2) + elapsed_weeks|parent.pop/mf) + (height.cm.prev + I(height.cm.prev^2) + elapsed_weeks|block)) },
  # 
  
  m11_cubic = {surv.wflow %>%
      add_model(surv.spec,
                formula = surv ~ height.cm.prev + I(height.cm.prev^2) + I(height.cm.prev^3) + (1|parent.pop/mf) + (1|block)) },
  
  m11d_cubic.date = {surv.wflow %>%
      add_model(surv.spec,
                formula = surv ~ height.cm.prev + survey_date_char + I(height.cm.prev^2) + I(height.cm.prev^3) + (1|parent.pop/mf) + (1|block)) },
  
  m11e_cubic.weeks = {surv.wflow %>%
      add_model(surv.spec,
                formula = surv ~ height.cm.prev + elapsed_weeks + I(height.cm.prev^2) + I(height.cm.prev^3) + (1|parent.pop/mf) + (1|block)) },
  
  m12_cubic.slope.parent.pop = {surv.wflow %>%
      add_model(surv.spec,
                formula = surv ~ height.cm.prev + I(height.cm.prev^2) + I(height.cm.prev^3) + (height.cm.prev + I(height.cm.prev^2) + I(height.cm.prev^3)|parent.pop) + (1|ppmf) + (1|block)) },
  
  m13_cubic.slope.parent.pop.mf = {surv.wflow %>%
      add_model(surv.spec,
                formula = surv ~ height.cm.prev + I(height.cm.prev^2) + I(height.cm.prev^3) + (height.cm.prev + I(height.cm.prev^2) + I(height.cm.prev^3)|parent.pop/mf) + (1|block)) },
  
  m14_cubic.slope.parent.pop.mf.blck = {surv.wflow %>%
      add_model(surv.spec,
                formula = surv ~ height.cm.prev + I(height.cm.prev^2) + I(height.cm.prev^3) + (height.cm.prev + I(height.cm.prev^2) + I(height.cm.prev^3)|parent.pop/mf) + (height.cm.prev + I(height.cm.prev^2) + I(height.cm.prev^3)|block)) },
  
  # keep date out of RE
  m14d_cubic.date.slope.parent.pop.mf.blck = {surv.wflow %>%
      add_model(surv.spec,
                formula = surv ~ height.cm.prev + survey_date_char + I(height.cm.prev^2) + I(height.cm.prev^3) + (height.cm.prev + I(height.cm.prev^2) + I(height.cm.prev^3)|parent.pop/mf) + (height.cm.prev + I(height.cm.prev^2) + I(height.cm.prev^3)|block)) },
  
  m14e_cubic.weeks.slope.parent.pop.mf.blck = {surv.wflow %>%
      add_model(surv.spec,
                formula = surv ~ height.cm.prev + elapsed_weeks + I(height.cm.prev^2) + I(height.cm.prev^3) + (height.cm.prev + I(height.cm.prev^2) + I(height.cm.prev^3) + elapsed_weeks|parent.pop/mf) + (height.cm.prev + I(height.cm.prev^2) + I(height.cm.prev^3) + elapsed_weeks|block)) }
),
name = names(wflow)
)

```

Fit the models

```{r}
#plan(multisession, workers = 7)
system.time( { # 3386.480  seconds (Vs 2540 if done with future_map)
  surv.models <- surv.models %>%
    #    mutate(fit = future_map(wflow, fit, data = survival2, .progress = TRUE, .options = furrr_options(packages = c("workflows", "multilevelmod"))),
    mutate(fit = map(wflow, fit, data = survival2, .progress = TRUE),
           glance = map(fit, glance)
    ) 
})

surv.models %>% select(-wflow, -fit) %>% unnest(glance) %>% arrange(BIC)
#plan(sequential)
```
```{r}
save.image(file="../output/WL2_2023_growth_survival_models.Rdata")
```

Take a look at some of the best models...

Concentrate on the five best models and do some cross-validation to better assess predictive power
```{r}
surv.models.best <- surv.models %>%
  unnest(glance) %>%
  slice_min(BIC, n=5)
```

```{r}
surv.models.best %>%
  pull(fit) %>%
  walk(print(.))

```



```{r}
set.seed(1001)
surv_folds <- vfold_cv(survival2, v = 10)

system.time( { #100 seconds
  surv.models.best <- surv.models.best %>%
    mutate(resamples = map(wflow, fit_resamples, resamples = surv_folds, control = control_resamples(save_pred = TRUE)))
})
```

```{r}
surv.models.best %>% 
  mutate(metrics = map(resamples, collect_metrics, type = "wide")) %>%
  select(name, metrics) %>%
  unnest(metrics) %>%
  select(-.config) %>%
  arrange(rmse)

```

```{r}
surv.models.best %>% 
  mutate(metrics = map(resamples, collect_metrics, type = "long")) %>%
  select(name, metrics) %>%
  unnest(metrics) %>%
  arrange(.metric, mean) %>%
  select(name, .metric, mean, std_err)

```

```{r}
surv.models.best %>% 
  mutate(metrics = map(resamples, collect_metrics, type = "long")) %>%
  select(name, metrics) %>%
  unnest(metrics) %>%
  arrange(.metric, mean) %>%
  select(name, .metric, mean, std_err) %>%
  ggplot(aes(x=name, y = mean, ymin=mean-std_err, ymax=mean+std_err, fill = name)) +
  geom_col() +
  geom_errorbar(width = 0.5) +
  facet_wrap(~.metric, scales = "free_y") +
  theme(axis.text.x = element_blank(), axis.title.x = element_blank()) +
  ggtitle("10-fold cross validation") +
  scale_fill_viridis_d()
```

So, we go with either m2d or m11d.  Given the overlapping error bars I lean towards the simpler one, but maybe plot them both first.

```{r}
surv.models.best2 <- surv.models.best %>% slice_min(AIC, n = 2)
```

For plotting the prediction curve, it looks like I can't combine `geom_function` and `facet_wrap` so I will compute the prediction curves myself and then combine in the plots

```{r}
survival2 %>% ungroup() %>% select(parent.pop, survey_date_char) %>% table()
newdata %>% ungroup() %>% select(parent.pop, survey_date_char) %>% table()

```


```{r}
surv.predict.func <- function(m, newdata=newdata.survival) {
  extract_fit_engine(m) %>%
  predict(newdata, type = "response", re.form = ~ (1 | parent.pop)) %>%
  cbind(newdata, surv=.)
}

newdata.survival <- survival2 %>%
  group_by(parent.pop, survey_date_char) %>%
  reframe(height.cm.prev = seq(min(height.cm.prev, na.rm = TRUE),
                               max(height.cm.prev, na.rm = TRUE),
                               length.out = 101)) %>%
  filter(parent.pop != "WV", survey_date_char != "2023-07-03")

surv.predictions <- surv.models.best2 %>%
  mutate(predict = map(fit, surv.predict.func)) %>%
  select(name, predict) %>%
  unnest(predict)

surv.predictions %>%
  ungroup() %>%
  nest(.by = parent.pop) %>% 
  mutate(plot = map2(data, parent.pop, \(x, pp) {
  ggplot(x, aes(x=height.cm.prev, y = surv)) +
      geom_line(aes(color=name)) +
      facet_wrap(~survey_date_char) +
      geom_point(alpha = 0.3, data = {survival2 %>% filter(parent.pop==pp)}) + 
      geom_smooth(alpha = 0.3, se = FALSE, data = {survival2 %>% filter(parent.pop==pp)}) + 
      ggtitle(pp)
  })) %>%
  pull(plot)
```

The cubic or linear size models both seem like reasonable options.
