---
title: "WL2023_Size_Growth"
author: "Julin Maloof"
date: "2025-01-24"
output: 
  html_document: 
    keep_md: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Testing WL2 2023 growth and survival models that include climate

```{r}
library(tidyverse)
library(tidymodels)
library(multilevelmod)
library(lmerTest)
library(broom.mixed)
library(furrr)
library(modelr)
library(zoo)
library(future)
plan(multisession, workers = 7)
conflicted::conflict_prefer("select", "dplyr")
conflicted::conflict_prefer("filter", "dplyr")
conflicted::conflict_prefer("lag", "dplyr")
```

## Import

### Climate data



## Read in soil temp

```{r}
temp <- read_csv("../input/WL2_2022_2023_iButton_Data_Corrected.csv") %>%
  select(-`...3`) %>%
  mutate(Date_Time = mdy_hm(Date_Time)) %>%
  filter(Date_Time > ymd("2023-07-06"))
head(temp)
skimr::skim(temp)
```

```{r}
temp %>% ggplot(aes(x=Date_Time, y = SoilTemp, color= Bed)) + 
  geom_line()

#Take a closer look
temp %>% mutate(month = month(Date_Time)) %>%
  filter(Date_Time > ymd("2023-07-17"), Date_Time <= ymd("2023-07-24")) %>%
  ggplot(aes(x=Date_Time, y = SoilTemp, color= Bed, lty = Bed)) + 
  geom_line() +
  scale_color_brewer(type="qual", palette = "Accent")

temp %>% mutate(month = month(Date_Time)) %>%
  filter(month==7) %>%
  ggplot(aes(x=Date_Time, y = SoilTemp)) + 
  geom_line() +
  scale_color_brewer(type="qual", palette = "Accent") +
  facet_wrap(~Bed)


```
Not sure how to deal with the differences here since we don't have something in every bed or block. For now I am going to take A_2

```{r}
temp_summary <- temp %>%
  filter(Bed=="A_2") %>%
  mutate(Date=as.Date(Date_Time)) %>%
  filter(Date != min(Date), Date != max(Date)) %>% # trim ragged ends
  group_by(Date) %>%
  summarize(
    min_temp_d = min(SoilTemp),
    max_temp_d = max(SoilTemp),
    mean_temp_d = mean(SoilTemp)
  ) %>%
  mutate(
    across(ends_with("temp_d"), \(x) rollmean(x, k = 7, align = "right", fill = NA), .names="{.col}1_7"),
    across(ends_with("temp_d"), \(x) rollmean(x, k = 13, align = "right", fill = NA), .names="{.col}1_13"), # 13 so I can get the first survey date
    across(ends_with("1_7"), \(x) dplyr::lag(x, 6), .names = "{.col}_13") 
  ) %>%
  rename_with(\(x) str_replace(x, "1_7_13", "6_13"))

temp_summary
```

## Read in soil moisture

```{r}
moisture <- read_csv("../input/WL2_2023_Bed_C_Soil_Moisture_Corrected.csv") %>%
  mutate(Date_Time = mdy_hm(Date_Time)) 
head(moisture)
skimr::skim(moisture)
```
```{r}
moisture %>% pivot_longer(-Date_Time) %>%
  ggplot(aes(x=Date_Time, y = value, color= name)) + 
  geom_line()
```
Will average across the ports

```{r}
moisture_summary <- moisture %>%
  rowwise() %>%
  mutate(s_moisture = mean(c_across(-Date_Time)) ) %>%
  select(Date_Time, s_moisture) %>%
  mutate(Date=as.Date(Date_Time)) %>%
  group_by(Date) %>%
  summarize(
    mean_moisture_d = mean(s_moisture)
  ) %>%
  mutate(
    s_moisture_1_7 = rollmean(mean_moisture_d, k = 7, align = "right", fill = "extend"),
    s_moisture_1_14 = rollmean(mean_moisture_d, k = 14, align = "right", fill = "extend"),
    s_moisture_7_14 = dplyr::lag(s_moisture_1_7, 7)
  )

moisture_summary
```


### Growth and survival data

```{r}
growth <- read_csv("../input/WL2-2023_Size_Combined.csv")
growth
```

```{r}
survival <- read_csv("../input/WL2_Mortality.csv")
survival
```

## Wrangle Growth

We want to predict size next from size.  In the Merow examples there was a column "size" and "size next" for each time step.  I guess we are going to need to do that here.

We also should compute the interval length 

```{r}
growth2 <- growth %>% arrange(Genotype, survey_date) %>%
  select(-survey.notes, -long.leaf.cm) %>%
  group_by(Genotype) %>%
  mutate(height.cm.next = lead(height.cm),
         elapsed_days= lead(survey_date) - survey_date,
         elapsed_weeks = as.integer(elapsed_days)/7) %>%
  ungroup() %>%
  mutate(week = as.numeric(survey_date - ymd("2023-07-03")) / 7, # 7/03 = pre-transplant survey date
         week_char = as.character(week)) %>% 
  drop_na(block, height.cm, height.cm.next)

growth2 
```

Add climate

```{r}
growth2 <- growth2 %>% 
  left_join(temp_summary, by = c("survey_date" = "Date")) %>%
  left_join(moisture_summary, by = c("survey_date" = "Date"))
```

For easy visualization lets subset just to elapsed_days of 7

```{r}
growth2 %>%
  filter(elapsed_days==7) %>%
  ggplot(aes(x=height.cm, y=height.cm.next)) +
  geom_point(aes(color=parent.pop), alpha = 0.3) +
  geom_abline() +
  geom_smooth(method="lm")
```

Do some model comparison

Define the formula


```{r}
minSize <- min(growth2$height.cm, na.rm = TRUE)
maxSize <- max(growth2$height.cm, na.rm = TRUE)
# newdata = tibble(height.cm=seq(minSize, maxSize, length.out=101), elapsed_weeks = 1)
```

In order to be able to use the tidymodel tools on mixed effect models, it works best to add variables and formulas in the following way.  This is a bit cumbersome, but it will help later.

Start with the preferred model from the test of models without climate.
```{r}
lmer.spec <- linear_reg() %>%
  set_engine("lmer")

growth.rec <- growth2 %>%
  select(height.cm.next, height.cm, elapsed_weeks, parent.pop, mf, block, contains("temp_d"), contains("s_moisture")) %>%
  recipe() %>%
  update_role(height.cm.next, new_role = "outcome") %>% 
  update_role(c(height.cm, elapsed_weeks, parent.pop, mf, block, contains("temp_d"), contains("s_moisture")), new_role = "predictor")

growth.rec.poly <- growth.rec %>% 
  step_poly(height.cm, degree = 3, keep_original_cols = TRUE)  %>%
  step_rename_at(contains("poly"), fn = \(x) str_replace(x, "_poly_", ".p"))

growth.wflow <- workflow() %>%
  add_recipe(growth.rec)

growth.wflow.poly <- workflow() %>% add_recipe(growth.rec.poly)

growth.models <- tibble(wflow=list(
  
  m1_cubic.parent.block = { growth.wflow.poly %>% 
      add_model(lmer.spec,
                formula = height.cm.next ~ height.cm.p1 + height.cm.p2 + height.cm.p3 + (1|parent.pop) + (1|block) )},
  
  m2_min_temp1_7 = { growth.wflow.poly %>% 
      add_model(lmer.spec,
                formula = height.cm.next ~ height.cm.p1 + height.cm.p2 + height.cm.p3 + min_temp_d1_7 + (1|parent.pop) + (1|block) )},
  
  m3_min_temp6_13 = { growth.wflow.poly %>% 
      add_model(lmer.spec,
                formula = height.cm.next ~ height.cm.p1 + height.cm.p2 + height.cm.p3 + min_temp_d6_13  + (1|parent.pop) + (1|block) )},
  
  m4_min_temp1_13 = { growth.wflow.poly %>% 
      add_model(lmer.spec,
                formula = height.cm.next ~ height.cm.p1 + height.cm.p2 + height.cm.p3 + min_temp_d1_13 + (1|parent.pop) + (1|block) )},
  
  m5_max_temp1_7 = { growth.wflow.poly %>% 
      add_model(lmer.spec,
                formula = height.cm.next ~ height.cm.p1 + height.cm.p2 + height.cm.p3 + max_temp_d1_7 + (1|parent.pop) + (1|block) )},
  
  # m5re_max_temp1_7 = { growth.wflow.poly %>% 
  #     add_model(lmer.spec,
  #               formula = height.cm.next ~ height.cm.p1 + height.cm.p2 + height.cm.p3 + max_temp_d1_7 + (max_temp_d1_7|parent.pop) + (1|block) )},
  
  m6_max_temp6_13 = { growth.wflow.poly %>% 
      add_model(lmer.spec,
                formula = height.cm.next ~ height.cm.p1 + height.cm.p2 + height.cm.p3 + max_temp_d6_13 + (1|parent.pop) + (1|block) )},
  
  m7_max_temp1_13 = { growth.wflow.poly %>% 
      add_model(lmer.spec,
                formula = height.cm.next ~ height.cm.p1 + height.cm.p2 + height.cm.p3 +  max_temp_d1_13 + (1|parent.pop) + (1|block) )},
  
  # convergence issue on cross-validation
  # m7re_max_temp1_13_reslope = { growth.wflow.poly %>% 
  #     add_model(lmer.spec,
  #               formula = height.cm.next ~ height.cm.p1 + height.cm.p2 + height.cm.p3 +  max_temp_d1_13 + (max_temp_d1_13|parent.pop) + (1|block) )},
  
  m8_mean_temp1_7 = { growth.wflow.poly %>% 
      add_model(lmer.spec,
                formula = height.cm.next ~ height.cm.p1 + height.cm.p2 + height.cm.p3  + mean_temp_d1_7 + (1|parent.pop) + (1|block) )},
  
  m9_mean_temp6_13 = { growth.wflow.poly %>% 
      add_model(lmer.spec,
                formula = height.cm.next ~ height.cm.p1 + height.cm.p2 + height.cm.p3 + mean_temp_d6_13 + (1|parent.pop) + (1|block) )},
  
  m10_mean_temp1_13 = { growth.wflow.poly %>% 
      add_model(lmer.spec,
                formula = height.cm.next ~ height.cm.p1 + height.cm.p2 + height.cm.p3 + mean_temp_d1_13 + (1|parent.pop) + (1|block) )},
  
  m11_mois1_7 = { growth.wflow.poly %>% 
      add_model(lmer.spec,
                formula = height.cm.next ~ height.cm.p1 + height.cm.p2 + height.cm.p3 + s_moisture_1_7 + (1|parent.pop) + (1|block) )},
  
  m12_mois7_14 = { growth.wflow.poly %>% 
      add_model(lmer.spec,
                formula = height.cm.next ~ height.cm.p1 + height.cm.p2 + height.cm.p3 + s_moisture_7_14 + (1|parent.pop) + (1|block) )},
  
  m12re_mois7_14.reslope = { growth.wflow.poly %>% 
      add_model(lmer.spec,
                formula = height.cm.next ~ height.cm.p1 + height.cm.p2 + height.cm.p3 + s_moisture_7_14 + (s_moisture_7_14|parent.pop) + (1|block) )},
  
  m13_mois1_14 = { growth.wflow.poly %>% 
      add_model(lmer.spec,
                formula = height.cm.next ~ height.cm.p1 + height.cm.p2 + height.cm.p3 + s_moisture_1_14 + (1|parent.pop) + (1|block) )},
  
  m14_mois1_7_maxtemp1_7 = { growth.wflow.poly %>% 
      add_model(lmer.spec,
                formula = height.cm.next ~ height.cm.p1 + height.cm.p2 + height.cm.p3 + s_moisture_1_7 + max_temp_d1_7 + (1|parent.pop) + (1|block) )}
  
),
name = names(wflow)
)
```

Fit the models

BEWARE NOBS IS DIFFERENT!

```{r}
system.time( { # fast
  growth.models <- growth.models %>%
    mutate(fit = map(wflow, fit, data = drop_na(growth2, contains("temp_d"), contains("s_moisture") )),
           glance = map(fit, glance)
    ) 
})

growth.models %>% select(-wflow, -fit) %>% unnest(glance) %>% arrange(BIC)
```

Concentrate on the five best models and do some cross validation
```{r}
growth.models.best <- growth.models %>%
  unnest(glance) %>%
  slice_min(BIC, n=5) %>%
  bind_rows({growth.models %>% filter(name == "m1_cubic.parent.block") })

#growth.models.best$fit

set.seed(1001)
growth_folds <- vfold_cv(growth2, v = 10)

growth.models.best <- growth.models.best %>%
  mutate(resamples = map(wflow, fit_resamples, resamples = growth_folds, control = control_resamples(save_pred = TRUE, pkgs = "tidyverse"), .progress = TRUE))
```

```{r}
growth.models.best %>% 
  mutate(metrics = map(resamples, collect_metrics, type = "wide")) %>%
  select(name, metrics) %>%
  unnest(metrics) %>%
  select(-.config) %>%
  arrange(rmse)

```

Doesn't help predictions :-(


```{r}
growth.models.best %>% 
  mutate(metrics = map(resamples, collect_metrics, type = "long")) %>%
  select(name, metrics) %>%
  unnest(metrics) %>%
  arrange(.metric, mean) %>%
  select(name, .metric, mean, std_err)

```

```{r}
growth.models.best %>% 
  mutate(metrics = map(resamples, collect_metrics, type = "long")) %>%
  select(name, metrics) %>%
  unnest(metrics) %>%
  arrange(.metric, mean) %>%
  select(name, .metric, mean, std_err) %>%
  ggplot(aes(x=name, y = mean, ymin=mean-std_err, ymax=mean+std_err, fill = name)) +
  geom_col() +
  geom_errorbar(width = 0.5) +
  facet_wrap(~.metric, scales = "free_y") +
  theme(axis.text.x = element_blank(), axis.title.x = element_blank()) +
  ggtitle("10-fold cross validation") +
  scale_fill_viridis_d()
```

So, might as well stick with the base model
```{r}
growth.model.final <- growth.models.best %>% filter(name == "m1_cubic.parent.block") %>% pull(fit) %>% magrittr::extract2(1)

growth.model.final
```

For plotting the prediction curve, it looks like I can't combine `geom_function` and `facet_wrap` so I will compute the prediction curves myself and then combine in the plots

```{r, eval=FALSE}
newdata.growth <- growth2 %>%
  group_by(parent.pop) %>%
  summarize(minSize = min(height.cm, na.rm = TRUE),
            maxSize = max(height.cm, na.rm = TRUE)) %>%
  
  # generate a "newdata" data frame for each parent.pop
  mutate(newdata = pmap(list(parent.pop, minSize, maxSize), \(p, x, y) tibble(parent.pop = p,
                                                                              elapsed_weeks = 1,
                                                                              height.cm = seq(x, y, length.out = 101)))) %>%
  select(newdata) %>%
  unnest(newdata)

newdata.growth <- growth.rec.poly %>% # These steps are necessary to add the polynomials to the new data
  update_role(mf, block, new_role = "unused") %>%
  update_role_requirements(role="unused", bake = FALSE) %>%
  prep() %>% bake(newdata.growth)

growth.predictions <- extract_fit_engine(growth.model.final) %>%
  predict(newdata.growth, re.form = ~ (1 | parent.pop)) %>%
  cbind(newdata.growth, height.cm.next=.)
```

Plot it.  Points are actual data (height.next vs height).  The blue line is the model prediction, the red line is slope = 1.
```{r, fig.asp=1, eval=FALSE}
growth2 %>%
  ggplot(aes(x=height.cm, y = height.cm.next)) +
  geom_point(alpha=.25) +
  geom_line(data = growth.predictions, color = "blue") +
  geom_abline(intercept = 0, slope = 1, color = "red", lty=2 ) +
  #geom_smooth(color = "red") +
  facet_wrap(~parent.pop, scales = "free")
```

## On to survival

Since we want to predict survival as a function of size, we need to combine the data sets

```{r}
survival2 <- survival %>% select(Genotype, death.date) %>%
  right_join(growth) %>%
  arrange(Genotype, survey_date) %>%
  select(-survey.notes, -long.leaf.cm) %>%
  group_by(Genotype) %>%
  mutate(elapsed_days= lead(survey_date) - survey_date,
         elapsed_weeks = as.integer(elapsed_days)/7,
         block = ifelse(is.na(block), unique(na.omit(block)), block), # fill in missing block info 
         death.date = lubridate::mdy(death.date),
         surv.next = ifelse(is.na(death.date), 1, death.date > lead(survey_date))) %>% 
  ungroup() %>%
  mutate(week = as.numeric(survey_date - ymd("2023-07-03")) / 7, # 7/03 = pre-transplant survey date
         week_char = as.character(round(week, 2)),
         ppmf = str_c(parent.pop, "_", mf)) %>% # allows me to specify RE for pp and mf separately.
  drop_na(surv.next, parent.pop, height.cm, elapsed_days)

# check it
survival2 %>% select(Genotype, death.date, survey_date, surv.next, height.cm)
```



## add climate data

```{r}
survival2 <- survival2 %>% 
  left_join(temp_summary, by = c("survey_date" = "Date")) %>%
  left_join(moisture_summary, by = c("survey_date" = "Date"))
```

Try the best linear and cubic models
```{r}
surv.spec <- linear_reg() %>%
  set_engine("glmer", family = "binomial")

surv.rec <- survival2 %>%
  select(surv.next, height.cm, elapsed_weeks, week, parent.pop, block, contains("temp_d"), contains("s_moisture") ) %>%
  recipe() %>%
  update_role(surv.next, new_role = "outcome") %>% 
  update_role(c(height.cm, elapsed_weeks, week, parent.pop, block, contains("temp_d"), contains("s_moisture")), new_role = "predictor") 

surv.rec.poly <- surv.rec %>% 
  step_poly(height.cm, degree = 3, keep_original_cols = TRUE)  %>%
  step_poly(week, degree = 3, keep_original_cols = TRUE) %>%
  step_normalize(all_numeric_predictors()) %>%
  step_rename_at(contains("poly"), fn = \(x) str_replace(x, "_poly_", ".p"))

surv.wflow <- workflow() %>%
  add_recipe(surv.rec)

surv.wflow.poly <- workflow() %>% add_recipe(surv.rec.poly)

surv.models <- tibble(wflow=list(
  
  ml1_linear = {surv.wflow %>%
      add_model(surv.spec,
                formula = surv.next ~ height.cm + week + (1|parent.pop) + (1|block)) },
  
  ml2_min_temp_1_7 = {surv.wflow %>%
      add_model(surv.spec,
                formula = surv.next ~ height.cm + week + min_temp_d1_7 + (1|parent.pop) + (1|block)) },
  
  ml3_min_temp_1_13 = {surv.wflow %>%
      add_model(surv.spec,
                formula = surv.next ~ height.cm + week + min_temp_d1_13 +(1|parent.pop) + (1|block)) },
  
  ml4_min_temp_6_13 = {surv.wflow %>%
      add_model(surv.spec,
                formula = surv.next ~ height.cm + week + min_temp_d6_13 + (1|parent.pop) + (1|block)) },
  
  ml5_mean_temp_1_7 = {surv.wflow %>%
      add_model(surv.spec,
                formula = surv.next ~ height.cm + week + mean_temp_d1_7 + (1|parent.pop) + (1|block)) },
  
  ml1_mean_temp_1_13 = {surv.wflow %>%
      add_model(surv.spec,
                formula = surv.next ~ height.cm + week + mean_temp_d1_13 + (1|parent.pop) + (1|block)) },
  
  ml1_mean_temp_6_13 = {surv.wflow %>%
      add_model(surv.spec,
                formula = surv.next ~ height.cm + week + mean_temp_d6_13 +(1|parent.pop) + (1|block)) },
  
  ml1_max_temp_1_7 = {surv.wflow %>%
      add_model(surv.spec,
                formula = surv.next ~ height.cm + week + max_temp_d1_7 + (1|parent.pop) + (1|block)) },
  
  ml1_max_temp_1_13 = {surv.wflow %>%
      add_model(surv.spec,
                formula = surv.next ~ height.cm + week + max_temp_d1_13 + (1|parent.pop) + (1|block)) },
  
  ml1_max_temp_6_13 = {surv.wflow %>%
      add_model(surv.spec,
                formula = surv.next ~ height.cm + week + max_temp_d6_13 + (1|parent.pop) + (1|block)) },
  
  ml1_s_moisture_1_7 = {surv.wflow %>%
      add_model(surv.spec,
                formula = surv.next ~ height.cm + week + s_moisture_1_7 + (1|parent.pop) + (1|block)) },
  
  ml1_s_moisture_7_14 = {surv.wflow %>%
      add_model(surv.spec,
                formula = surv.next ~ height.cm + week + s_moisture_7_14 + (1|parent.pop) + (1|block)) },
  
  ml1_s_moisture_1_14 = {surv.wflow %>%
      add_model(surv.spec,
                formula = surv.next ~ height.cm + week + s_moisture_1_14 + (1|parent.pop) + (1|block)) },
  
  mc1_cubic = {surv.wflow.poly %>%
      add_model(surv.spec,
                formula = surv.next ~ height.cm.p1 + elapsed_weeks + week.p1 + week.p2 + week.p3 + height.cm.p2 + height.cm.p3 + (1|parent.pop) + (1|block)) },
  
  mc2_min_temp_1_7 = {surv.wflow.poly %>%
      add_model(surv.spec,
                formula = surv.next ~ height.cm.p1 + elapsed_weeks + week.p1 + week.p2 + week.p3 + height.cm.p2 + height.cm.p3 + min_temp_d1_7 + (1|parent.pop) + (1|block)) },
  
  mc3_min_temp_1_13 = {surv.wflow.poly %>%
      add_model(surv.spec,
                formula = surv.next ~ height.cm.p1 + elapsed_weeks + week.p1 + week.p2 + week.p3 + height.cm.p2 + height.cm.p3 + min_temp_d1_13 +(1|parent.pop) + (1|block)) },
  
  mc4_min_temp_6_13 = {surv.wflow.poly %>%
      add_model(surv.spec,
                formula = surv.next ~ height.cm.p1 + elapsed_weeks + week.p1 + week.p2 + week.p3 + height.cm.p2 + height.cm.p3 + min_temp_d6_13 + (1|parent.pop) + (1|block)) },
  
  mc5_mean_temp_1_7 = {surv.wflow.poly %>%
      add_model(surv.spec,
                formula = surv.next ~ height.cm.p1 + elapsed_weeks + week.p1 + week.p2 + week.p3 + height.cm.p2 + height.cm.p3 + mean_temp_d1_7 + (1|parent.pop) + (1|block)) },
  
  mc6_mean_temp_1_13 = {surv.wflow.poly %>%
      add_model(surv.spec,
                formula = surv.next ~ height.cm.p1 + elapsed_weeks + week.p1 + week.p2 + week.p3 + height.cm.p2 + height.cm.p3 + mean_temp_d1_13 + (1|parent.pop) + (1|block)) },
  
  mc7_mean_temp_6_13 = {surv.wflow.poly %>%
      add_model(surv.spec,
                formula = surv.next ~ height.cm.p1 + elapsed_weeks + week.p1 + week.p2 + week.p3 + height.cm.p2 + height.cm.p3 + mean_temp_d6_13 +(1|parent.pop) + (1|block)) },
  
  mc8_max_temp_1_7 = {surv.wflow.poly %>%
      add_model(surv.spec,
                formula = surv.next ~ height.cm.p1 + elapsed_weeks + week.p1 + week.p2 + week.p3 + height.cm.p2 + height.cm.p3 + max_temp_d1_7 + (1|parent.pop) + (1|block)) },
  
  mc9_max_temp_1_13 = {surv.wflow.poly %>%
      add_model(surv.spec,
                formula = surv.next ~ height.cm.p1 + elapsed_weeks + week.p1 + week.p2 + week.p3 + height.cm.p2 + height.cm.p3 + max_temp_d1_13 + (1|parent.pop) + (1|block)) },
  
  mc10_max_temp_6_13 = {surv.wflow.poly %>%
      add_model(surv.spec,
                formula = surv.next ~ height.cm.p1 + elapsed_weeks + week.p1 + week.p2 + week.p3 + height.cm.p2 + height.cm.p3 + max_temp_d6_13 + (1|parent.pop) + (1|block)) },
  
  mc11_moisture_1_7 = {surv.wflow.poly %>%
      add_model(surv.spec,
                formula = surv.next ~ height.cm.p1 + elapsed_weeks + week.p1 + week.p2 + week.p3 + height.cm.p2 + height.cm.p3 + s_moisture_1_7 + (1|parent.pop) + (1|block)) },
  
  mc12_moisture_7_14 = {surv.wflow.poly %>%
      add_model(surv.spec,
                formula = surv.next ~ height.cm.p1 + elapsed_weeks + week.p1 + week.p2 + week.p3 + height.cm.p2 + height.cm.p3 + s_moisture_7_14 + (1|parent.pop) + (1|block)) },
  
  mc13_moisture_1_14 = {surv.wflow.poly %>%
      add_model(surv.spec,
                formula = surv.next ~ height.cm.p1 + elapsed_weeks + week.p1 + week.p2 + week.p3 + height.cm.p2 + height.cm.p3 + s_moisture_1_14 + (1|parent.pop) + (1|block)) }
  
),
name = names(wflow)
)
```

Fit the models

```{r}
plan(multisession, workers = 7)

system.time( { # 110 seconds if using multisession; 
  surv.models <- surv.models %>%
    mutate(fit = future_map(wflow, fit, data = drop_na(survival2,contains("temp_d"), contains("s_moisture") ),  .progress = TRUE, .options = furrr_options(packages = c("workflows", "multilevelmod"))),
           #mutate(fit = map(wflow, fit, data = survival2, .progress = TRUE),
           glance = map(fit, glance)
    ) 
})

surv.models %>% select(-wflow, -fit) %>% unnest(glance) %>% arrange(BIC)
plan(sequential)
```
No joy

```{r}
surv.models <- surv.models %>% mutate(messages = map(fit, \(x) {x <- extract_fit_engine(x)
c(x@optinfo$message, unlist(x@optinfo$warnings), unlist(x@optinfo$conv$lme4$messages))
})
) 

surv.models %>% select(name, messages) %>% unnest(messages) %>% filter(str_detect(messages, "Model"))
```


Take a look at some of the best models...

Concentrate on the three best models
```{r}
surv.models.best <- surv.models %>%
  unnest(glance) %>%
  filter(rank(BIC) <= 5 | str_detect(name, "ml1_linear")) %>%
  arrange(BIC)

surv.models.best %>% select(-wflow, -fit)
```

```{r}
surv.models.best %>%
  pull(fit) 
```


```{r}
set.seed(1001)
surv_folds <- vfold_cv(survival2, v = 10)

system.time( { #127 seconds
  surv.models.best <- surv.models.best %>%
    mutate(resamples = map(wflow, fit_resamples, resamples = surv_folds, control = control_resamples(save_pred = TRUE)))
})
```

```{r}
surv.models.best %>% 
  mutate(metrics = map(resamples, collect_metrics, type = "wide")) %>%
  select(name, metrics) %>%
  unnest(metrics) %>%
  select(-.config) %>%
  arrange(desc(rsq))
```
cubic for both weeks and size is the best

```{r}
surv.models.best %>% 
  mutate(metrics = map(resamples, collect_metrics, type = "long")) %>%
  select(name, metrics) %>%
  unnest(metrics) %>%
  arrange(.metric, mean) %>%
  select(name, .metric, mean, std_err)

```

```{r}
surv.models.best %>% 
  mutate(metrics = map(resamples, collect_metrics, type = "long")) %>%
  select(name, metrics) %>%
  unnest(metrics) %>%
  arrange(.metric, mean) %>%
  select(name, .metric, mean, std_err) %>%
  ggplot(aes(x=reorder(name, mean), y = mean, ymin=mean-std_err, ymax=mean+std_err, fill = reorder(name, mean))) +
  geom_col() +
  geom_errorbar(width = 0.5) +
  facet_wrap(~.metric, scales = "free_y") +
  theme(axis.text.x = element_blank(), axis.title.x = element_blank()) +
  ggtitle("10-fold cross validation") +
  scale_fill_viridis_d()
```

```{r}
surv.model.final <- surv.models.best %>% slice_min(BIC) %>% pull(fit) %>% magrittr::extract2(1)
```

For plotting the prediction curve, it looks like I can't combine `geom_function` and `facet_wrap` so I will compute the prediction curves myself and then combine in the plots.

```{r}
newdata.survival <- survival2 %>%
  group_by(parent.pop, week) %>%
  reframe(height.cm = seq(min(height.cm, na.rm = TRUE),
                          max(height.cm, na.rm = TRUE),
                          length.out = 101),
          elapsed_weeks = 1) %>%
  filter(parent.pop != "WV")

surv.predictions <- surv.model.final %>% 
  extract_fit_engine() %>%
  predict(newdata.survival, type = "response", re.form = ~ (1 | parent.pop)) %>%
  cbind(newdata.survival, surv.next=.)

surv.predictions %>%
  ungroup() %>%
  nest(.by = parent.pop) %>% 
  mutate(plot = map2(data, parent.pop, \(x, pp) {
    ggplot(x, aes(x=height.cm, y = surv.next)) +
      geom_smooth(color = "grey50", lty=3, se = FALSE, method = "gam", data = {survival2 %>% filter(parent.pop==pp)}) + 
      geom_line() +
      facet_wrap(~week, scale = "free_x") +
      geom_point(alpha = 0.3, data = {survival2 %>% filter(parent.pop==pp)}) + 
      ggtitle(pp) +
      scale_color_brewer(type = "qual", palette = "Accent")
  })) %>%
  pull(plot) %>% walk(print)
```

```{r}
surv.model.final <- surv.model.final %>% extract_fit_engine()
growth.model.final <- growth.model.final %>% extract_fit_engine()
```

