---
title: "WL2023_Size_Growth"
author: "Julin Maloof"
date: "2025-01-24"
output: 
  html_document: 
    keep_md: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Intitial attempt at making a growth/survival matrix from WL2 2023, now using random effects

```{r}
library(tidyverse)
library(tidymodels)
library(multilevelmod)
library(lmerTest)
library(broom.mixed)
library(doMC)
library(furrr)
library(modelr)
registerDoMC(cores = 7)
conflicted::conflict_prefer("select", "dplyr")
conflicted::conflict_prefer("filter", "dplyr")
conflicted::conflict_prefer("lag", "dplyr")
```

## Import the data

```{r}
growth <- read_csv("../input/WL2-2023_Size_Combined.csv")
growth
```

```{r}
survival <- read_csv("../input/WL2_Mortality.csv")
survival
```

## Wrangle Growth

We want to predict size next from size.  In the Merow examples there was a column "size" and "size next" for each time step.  I guess we are going to need to do that here.

We also should compute the interval length 

```{r}
growth2 <- growth %>% arrange(Genotype, survey_date) %>%
  select(-survey.notes, -long.leaf.cm) %>%
  group_by(Genotype) %>%
  mutate(height.cm.next = lead(height.cm),
         elapsed_days= lead(survey_date) - survey_date,
         elapsed_weeks = as.integer(elapsed_days)/7) %>%
  ungroup() %>%
  mutate(week = as.numeric(survey_date - ymd("2023-07-03")) / 7, # 7/03 = pre-transplant survey date
         week_char = as.character(week)) %>% 
  drop_na(block, height.cm, height.cm.next)

growth2 
```

For easy visualization lets subset just to elapsed_days of 7

```{r}
growth2 %>%
  filter(elapsed_days==7) %>%
  ggplot(aes(x=height.cm, y=height.cm.next)) +
  geom_point(aes(color=parent.pop), alpha = 0.3) +
  geom_abline() +
  geom_smooth(method="lm")
```

Do some model comparison

Define the formula


```{r}
minSize <- min(growth2$height.cm, na.rm = TRUE)
maxSize <- max(growth2$height.cm, na.rm = TRUE)
# newdata = tibble(height.cm=seq(minSize, maxSize, length.out=101), elapsed_weeks = 1)
```


In order to be able to use the tidymodel tools on mixed effect models, it works best to add variables and formulas in the following way.  This is a bit cumbersome, but it will help later.

elapsed weeks is probably important, it is the time between survey dates, so if it has been 2 weeks instead of 1, we expect more growth
```{r}
lmer.spec <- linear_reg() %>%
  set_engine("lmer")

growth.rec <- growth2 %>%
  select(height.cm.next, height.cm, elapsed_weeks, parent.pop, mf, block) %>%
  recipe() %>%
  update_role(height.cm.next, new_role = "outcome") %>% 
  update_role(c(height.cm, elapsed_weeks, parent.pop, mf, block), new_role = "predictor")

growth.rec.poly <- growth.rec %>% 
  step_poly(height.cm, degree = 3, keep_original_cols = TRUE)  %>%
  step_rename_at(contains("poly"), fn = \(x) str_replace(x, "_poly_", ".p"))

growth.wflow <- workflow() %>%
  add_recipe(growth.rec)

growth.wflow.poly <- workflow() %>% add_recipe(growth.rec.poly)

growth.models <- tibble(wflow=list(
  
### Models with only Random Effects
  m1_block = {growth.wflow %>%
      add_model(lmer.spec,
                formula = height.cm.next ~ (1|block)) },
  
  m2_parent = { growth.wflow %>% 
      add_model(lmer.spec,
                formula = height.cm.next ~ (1|parent.pop) )},
  
  m3_block.parent = { growth.wflow %>% 
      add_model(lmer.spec,
                formula = height.cm.next ~ (1|parent.pop) + (1|block) )},
  
  m4_block.parent.mf = { growth.wflow %>% 
      add_model(lmer.spec,
                formula = height.cm.next ~ (1|parent.pop/mf) + (1|block) )},
  
### Models with a linear height component.  Compare also having random effects slope for height w.r.t. parent.pop and block.
  
  m5_linear.parent.mf.block = { growth.wflow %>% 
      add_model(lmer.spec,
                formula = height.cm.next ~ height.cm + (1|parent.pop/mf) + (1|block) )},
  
  m6_linear.parent.mf.block.slope = { growth.wflow %>% 
      add_model(lmer.spec,
                formula = height.cm.next ~ height.cm + (height.cm|parent.pop/mf) + (1|block) )},
  
  m7_linear.parent.mf.block.slope.slope = { growth.wflow %>% 
      add_model(lmer.spec,
                formula = height.cm.next ~ height.cm + (height.cm|parent.pop/mf) + (height.cm|block) )},
  
  # adding this after looking at m7...intercept and slope for mf have a correlation of -1.00.  Also, very little variation attributable to mf, so remove it all the way.
  m7s_linear.parent.block.slope.slope = { growth.wflow %>% 
      add_model(lmer.spec,
                formula = height.cm.next ~ height.cm + (height.cm|parent.pop) + (height.cm|block) )},

### Models with a linear height component and elapsed weeks.
  
  m8_linear_with_weeks_int.parent = { growth.wflow %>% 
      add_model(lmer.spec,
                formula = height.cm.next ~ height.cm*elapsed_weeks + (1|parent.pop/mf) + (1|block) )},
  
  m9_linear_with_weeks_int.parent.all.slope = { growth.wflow %>% 
      add_model(lmer.spec,
                formula = height.cm.next ~ height.cm*elapsed_weeks + (height.cm|parent.pop/mf) + (height.cm|block) )},

### Models with quadratic and cubic terms for height
  
  m10_quadratic.parent.mf.block = { growth.wflow.poly %>% 
      add_model(lmer.spec,
                formula = height.cm.next ~ height.cm.p1 + height.cm.p2 + (1|parent.pop/mf) + (1|block) )},
  
  m11_cubic.parent.mf.block = { growth.wflow.poly %>% 
      add_model(lmer.spec,
                formula = height.cm.next ~ height.cm.p1 + height.cm.p2 + height.cm.p3 + (1|parent.pop/mf) + (1|block) )},

  # also try m11 without mf because m11 fits well but has no variance associated with mf.

  m11s_cubic.parent.block = { growth.wflow.poly %>% 
      add_model(lmer.spec,
                formula = height.cm.next ~ height.cm.p1 + height.cm.p2 + height.cm.p3 + (1|parent.pop) + (1|block) )},

  m12_quadratic_weeks.parent.height.slope = { growth.wflow.poly %>% 
      add_model(lmer.spec,
                formula = height.cm.next ~ height.cm.p1*elapsed_weeks*parent.pop + height.cm.p2 + (height.cm.p1 + height.cm.p2 | parent.pop / mf) + (height.cm.p1 + height.cm.p2 | block) )}, 
  
  m13_cubic_weeks.parent.height.slope = { growth.wflow.poly %>% 
      add_model(lmer.spec,
                formula = height.cm.next ~ height.cm.p1*elapsed_weeks*parent.pop + height.cm.p2 + height.cm.p3 + (height.cm.p1 + height.cm.p2 | parent.pop) + (height.cm.p1 + height.cm.p2 | block) )}
),
name = names(wflow)
)
```

Fit the models

```{r}
system.time( { # 190 seconds
  growth.models <- growth.models %>%
    mutate(fit = map(wflow, fit, data = growth2),
           glance = map(fit, glance)
    ) 
})

growth.models %>% select(-wflow, -fit) %>% unnest(glance) %>% arrange(BIC)
```

Concentrate on the five best models.
```{r}
growth.models.best <- growth.models %>%
  unnest(glance) %>%
  slice_min(BIC, n=6)

growth.models.best$fit
```


Do some cross-validation to better assess predictive power
```{r}
set.seed(1001)
growth_folds <- vfold_cv(growth2, v = 10)

growth.models.best <- growth.models.best %>%
  mutate(resamples = map(wflow, fit_resamples, resamples = growth_folds, control = control_resamples(save_pred = TRUE), .progress = TRUE))
```

```{r}
growth.models.best %>% 
  mutate(metrics = map(resamples, collect_metrics, type = "wide")) %>%
  select(name, metrics) %>%
  unnest(metrics) %>%
  select(-.config) %>%
  arrange(rmse)

```
Interesting.  The rsq is very similar for all models, but m11 clearly wins on rmse.

```{r}
growth.models.best %>% 
  mutate(metrics = map(resamples, collect_metrics, type = "long")) %>%
  select(name, metrics) %>%
  unnest(metrics) %>%
  arrange(.metric, mean) %>%
  select(name, .metric, mean, std_err)

```

```{r}
growth.models.best %>% 
  mutate(metrics = map(resamples, collect_metrics, type = "long")) %>%
  select(name, metrics) %>%
  unnest(metrics) %>%
  arrange(.metric, mean) %>%
  select(name, .metric, mean, std_err) %>%
  ggplot(aes(x=name, y = mean, ymin=mean-std_err, ymax=mean+std_err, fill = name)) +
  geom_col() +
  geom_errorbar(width = 0.5) +
  facet_wrap(~.metric, scales = "free_y") +
  theme(axis.text.x = element_blank(), axis.title.x = element_blank()) +
  ggtitle("10-fold cross validation") +
  scale_fill_viridis_d()
```

So, we go with simplified m11 (or possibly m12 if we decide we need weeks).
```{r}
growth.model.final <- growth.models.best %>% filter(name == "m11s_cubic.parent.block") %>% pull(fit) %>% magrittr::extract2(1)

growth.model.final
```

For plotting the prediction curve, it looks like I can't combine `geom_function` and `facet_wrap` so I will compute the prediction curves myself and then combine in the plots

```{r}
newdata.growth <- growth2 %>%
  group_by(parent.pop) %>%
  summarize(minSize = min(height.cm, na.rm = TRUE),
            maxSize = max(height.cm, na.rm = TRUE)) %>%
  
  # generate a "newdata" data frame for each parent.pop
  mutate(newdata = pmap(list(parent.pop, minSize, maxSize), \(p, x, y) tibble(parent.pop = p,
                                                                              elapsed_weeks = 1,
                                                                              height.cm = seq(x, y, length.out = 101)))) %>%
  select(newdata) %>%
  unnest(newdata)

newdata.growth <- growth.rec.poly %>% # These steps are necessary to add the polynomials to the new data
  update_role(mf, block, new_role = "unused") %>%
  update_role_requirements(role="unused", bake = FALSE) %>%
  prep() %>% bake(newdata.growth)

growth.predictions <- extract_fit_engine(growth.model.final) %>%
  predict(newdata.growth, re.form = ~ (1 | parent.pop)) %>%
  cbind(newdata.growth, height.cm.next=.)
```

Plot it.  Points are actual data (height.next vs height).  The blue line is the model prediction, the red line is slope = 1.
```{r, fig.asp=1}
growth2 %>%
  ggplot(aes(x=height.cm, y = height.cm.next)) +
  geom_point(alpha=.25) +
  geom_line(data = growth.predictions, color = "blue") +
  geom_abline(intercept = 0, slope = 1, color = "red", lty=2 ) +
  #geom_smooth(color = "red") +
  facet_wrap(~parent.pop, scales = "free")
```

## On to survival

Since we want to predict survival as a function of size, we need to combine the data sets

```{r}
survival2 <- survival %>% select(Genotype, death.date) %>%
  right_join(growth) %>%
  arrange(Genotype, survey_date) %>%
  select(-survey.notes, -long.leaf.cm) %>%
  group_by(Genotype) %>%
  mutate(elapsed_days= lead(survey_date) - survey_date,
         elapsed_weeks = as.integer(elapsed_days)/7,
         block = ifelse(is.na(block), unique(na.omit(block)), block), # fill in missing block info 
         death.date = lubridate::mdy(death.date),
         surv.next = ifelse(is.na(death.date), 1, death.date > lead(survey_date))) %>% 
  ungroup() %>%
  mutate(week = as.numeric(survey_date - ymd("2023-07-03")) / 7, # 7/03 = pre-transplant survey date
         week_char = as.character(round(week, 2)),
         ppmf = str_c(parent.pop, "_", mf)) %>% # allows me to specify RE for pp and mf separately.
  drop_na(surv.next, parent.pop, height.cm, elapsed_days)

# check it
survival2 %>% select(Genotype, death.date, survey_date, surv.next, height.cm)
```
I SHOULD TRY DIFFERENT OPTIMIZERS TO SEE IF I CAN GET MORE OF THESE TO CONVERGE
```{r}
surv.spec <- linear_reg() %>%
  set_engine("glmer", family = "binomial")

surv.rec <- survival2 %>%
  select(surv.next, height.cm, elapsed_weeks, week, parent.pop, mf, ppmf, block) %>%
  recipe() %>%
  update_role(surv.next, new_role = "outcome") %>% 
  update_role(c(height.cm, elapsed_weeks, week, parent.pop, mf, ppmf, block), new_role = "predictor")

surv.rec.poly <- surv.rec %>% 
  step_poly(height.cm, degree = 3, keep_original_cols = TRUE)  %>%
  step_poly(week, degree = 3, keep_original_cols = TRUE) %>%
  step_rename_at(contains("poly"), fn = \(x) str_replace(x, "_poly_", ".p"))

surv.wflow <- workflow() %>%
  add_recipe(surv.rec)

surv.wflow.poly <- workflow() %>% add_recipe(surv.rec.poly)

surv.models <- tibble(wflow=list(
  m1_RE = {surv.wflow %>%
      add_model(surv.spec,
                formula = surv.next ~ (1|parent.pop/mf) + (1|block)) },
  
  m1b_RE2 = {surv.wflow %>% # reality check; this should be the same as M1
      add_model(surv.spec,
                formula = surv.next ~ (1|parent.pop) + (1|ppmf) + (1|block)) },
  
  m2_linear = {surv.wflow %>%
      add_model(surv.spec,
                formula = surv.next ~ height.cm + (1|parent.pop/mf) + (1|block)) },
  
  m2_linear.week_cont = {surv.wflow %>%
      add_model(surv.spec,
                formula = surv.next ~ height.cm + week + (1|parent.pop/mf) + (1|block)) },
  
  m2e_linear.elap_weeks = {surv.wflow %>%
      add_model(surv.spec,
                formula = surv.next ~ height.cm + elapsed_weeks + (1|parent.pop/mf) + (1|block)) },
  
  m2wne_linear.elap_weeks.week_cont = {surv.wflow %>%
      add_model(surv.spec,
                formula = surv.next ~ height.cm + week + elapsed_weeks + (1|parent.pop/mf) + (1|block)) },
  
  m3_linear.slope.parent.pop = {surv.wflow %>%
      add_model(surv.spec,
                formula = surv.next ~ height.cm + (height.cm|parent.pop) + (1|ppmf) + (1|block)) },
  
  m5_linear.slope.parent.pop.mf = {surv.wflow %>%
      add_model(surv.spec,
                formula = surv.next ~ height.cm + (height.cm|parent.pop/mf) + (1|block)) },
  
  m6_linear.slope.parent.pop.mf.block = {surv.wflow %>%
      add_model(surv.spec,
                formula = surv.next ~ height.cm + (height.cm|parent.pop/mf) + (height.cm|block)) },
  
  m7_quad = {surv.wflow.poly %>%
      add_model(surv.spec,
                formula = surv.next ~ height.cm + height.cm.p2 + (1|parent.pop/mf) + (1|block)) },
  
  m8_quad.slope.parent.pop = {surv.wflow.poly %>%
      add_model(surv.spec,
                formula = surv.next ~ height.cm.p1 + height.cm.p2 + (height.cm.p1 + height.cm.p2|parent.pop) + (1|ppmf) + (1|block)) }
  
),
name = names(wflow)
)

  
  #Too Slow when including date as a slope for RE, so remove that
  # Still does not converge
  # m6d_linear.date.slope.parent.pop.mf.block = {surv.wflow %>%
  #     add_model(surv.spec,
  #               formula = surv.next ~ height.cm + week_char + (height.cm|parent.pop/mf) + (height.cm|block)) },
  # 
  # m6e_linear.weeks.slope.parent.pop.mf.block = {surv.wflow %>%
  #     add_model(surv.spec,
  #               formula = surv.next ~ height.cm + elapsed_weeks + (elapsed_weeks + height.cm|parent.pop/mf) + (height.cm|block)) },
  
  # Does not converge
  #   m7_quad = {surv.wflow %>%
  #     add_model(surv.spec,
  #               formula = surv.next ~ height.cm + height.cm.p2 + (1|parent.pop/mf) + (1|block)) },
  # 
  # m7d_quad.date = {surv.wflow %>%
  #     add_model(surv.spec,
  #               formula = surv.next ~ height.cm + week_char + height.cm.p2 + (1|parent.pop/mf) + (1|block)) },
  # 
  # m7e_quad.weeks = {surv.wflow %>%
  #     add_model(surv.spec,
  #               formula = surv.next ~ height.cm + elapsed_weeks + height.cm.p2 + (1|parent.pop/mf) + (1|block)) },
  
 
  # Fails to converge
  # m9_quad.slope.parent.pop.mf = {surv.wflow %>%
  #     add_model(surv.spec,
  #               formula = surv.next ~ height.cm + height.cm.p2 + (height.cm + height.cm.p2|parent.pop/mf) + (1|block)) },
  

  
  # Keep date out of RE so it will fit
  # Still fails to converge
  # m10d_quad.date.slope.parent.pop.mf.blck = {surv.wflow %>%
  #     add_model(surv.spec,
  #               formula = surv.next ~ height.cm + week_char + height.cm.p2 + (height.cm + height.cm.p2|parent.pop/mf) + (height.cm + height.cm.p2|block)) },
  # 
  # m10e_quad.weeks.slope.parent.pop.mf.blck = {surv.wflow %>%
  #     add_model(surv.spec,
  #               formula = surv.next ~ height.cm + elapsed_weeks + height.cm.p2 + (height.cm + height.cm.p2 + elapsed_weeks|parent.pop/mf) + (height.cm + height.cm.p2 + elapsed_weeks|block)) },
  # 

 
  # m10_quad.slope.parent.pop.mf.blck = {surv.wflow.poly %>%
  #    add_model(surv.spec,
  #              formula = surv.next ~ height.cm.p1 + height.cm.p2 + (height.cm.p1 + height.cm.p2|parent.pop/mf) + (height.cm.p1 + height.cm.p2|block)) },
  
  # m11_cubic = {surv.wflow.poly %>%
  #     add_model(surv.spec,
  #               formula = surv.next ~ height.cm.p1 + height.cm.p2 + height.cm.p3 + (1|parent.pop/mf) + (1|block)) },
  # 
  # m11e_cubic.elap_weeks = {surv.wflow.poly %>%
  #     add_model(surv.spec,
  #               formula = surv.next ~ height.cm.p1 + elapsed_weeks + height.cm.p2 + height.cm.p3 + (1|parent.pop/mf) + (1|block)) },
  # 
  # m11wne_cubic.date.elap_weeks.weeks_cont = {surv.wflow.poly %>%
  #     add_model(surv.spec,
  #               formula = surv.next ~ height.cm.p1 + elapsed_weeks + week + height.cm.p2 + height.cm.p3 + (1|parent.pop/mf) + (1|block)) },
  # 
  # m11wnec_cubic.date.elap_weeks.cubic.weeks_cont = {surv.wflow.poly %>%
  #     add_model(surv.spec,
  #               formula = surv.next ~ height.cm.p1 + elapsed_weeks + week.p1 + week.p2 + week.p3 + height.cm.p2 + height.cm.p3 + (1|parent.pop/mf) + (1|block)) },
# 
#   m12_cubic.slope.parent.pop = {surv.wflow.poly %>%
#       add_model(surv.spec,
#                 formula = surv.next ~ height.cm.p1 + height.cm.p2 + height.cm.p3 + (height.cm.p1|parent.pop) + (1|ppmf) + (1|block)) },
#   
#   m12wn_cubic.week_cont.slope.parent.pop = {surv.wflow.poly %>%
#       add_model(surv.spec,
#                 formula = surv.next ~ height.cm.p1 + week + height.cm.p2 + height.cm.p3 + (height.cm.p1|parent.pop) + (1|ppmf) + (1|block)) },
#   
#   m12e_cubic.elap_weeks.slope.parent.pop = {surv.wflow.poly %>%
#       add_model(surv.spec,
#                 formula = surv.next ~ height.cm.p1 + elapsed_weeks + height.cm.p2 + height.cm.p3 + (height.cm.p1|parent.pop) + (1|ppmf) + (1|block)) },
#   
#   m12wne_cubic.date.elap_weeks.weeks_cont.slope.parent.pop = {surv.wflow.poly %>%
#       add_model(surv.spec,
#                 formula = surv.next ~ height.cm.p1 + elapsed_weeks + week + height.cm.p2 + height.cm.p3 + (height.cm.p1|parent.pop) + (1|ppmf) + (1|block)) },
#   
#   m13_cubic.slope.parent.pop.blk = {surv.wflow.poly %>%
#       add_model(surv.spec,
#                 formula = surv.next ~ height.cm.p1 + height.cm.p2 + height.cm.p3 + (height.cm.p1|parent.pop) + (1|ppmf) + (height.cm|block)) },
#   
#   m13wn_cubic.week_cont.slope.parent.pop.blk = {surv.wflow.poly %>%
#       add_model(surv.spec,
#                 formula = surv.next ~ height.cm.p1 + week + height.cm.p2 + height.cm.p3 + (height.cm.p1|parent.pop) + (1|ppmf) + (height.cm|block)) },
#   
#   m13e_cubic.elap_weeks.slope.parent.pop.blk = {surv.wflow.poly %>%
#       add_model(surv.spec,
#                 formula = surv.next ~ height.cm.p1 + elapsed_weeks + height.cm.p2 + height.cm.p3 + (height.cm.p1|parent.pop) + (1|ppmf) + (height.cm.p1|block)) },
#   
#   m13wne_cubic.date.elap_weeks.weeks_cont.slope.parent.pop.blk = {surv.wflow.poly %>%
#       add_model(surv.spec,
#                 formula = surv.next ~ height.cm.p1 + elapsed_weeks + week + height.cm.p2 + height.cm.p3 + (height.cm.p1|parent.pop) + (1|ppmf) + (height.cm.p1|block)) }

```

Fit the models

```{r}
plan(multisession, workers = 7)

system.time( { # 110 seconds if using multisession; 
  surv.models <- surv.models %>%
        mutate(fit = future_map(wflow, fit, data = survival2, .progress = TRUE, .options = furrr_options(packages = c("workflows", "multilevelmod"))),
    #mutate(fit = map(wflow, fit, data = survival2, .progress = TRUE),
           glance = map(fit, glance)
    ) 
})

surv.models %>% select(-wflow, -fit) %>% unnest(glance) %>% arrange(BIC)
plan(sequential)
```

```{r}
surv.models <- surv.models %>% mutate(messages = map(fit, \(x) {x <- extract_fit_engine(x)
c(x@optinfo$message, unlist(x@optinfo$warnings), unlist(x@optinfo$conv$lme4$messages))
})
) 

surv.models %>% select(name, messages) %>% unnest(messages) %>% filter(str_detect(messages, "Model"))
```



```{r}
save.image(file="../output/WL2_2023_growth_survival_models.Rdata")
```

Take a look at some of the best models...

Concentrate on the three best models
```{r}
surv.models.best <- surv.models %>%
  unnest(glance) %>%
  filter(rank(BIC) <= 5 ) %>%
  arrange(BIC)

surv.models.best
```

```{r}
surv.models.best %>%
  pull(fit) 
```


```{r}
set.seed(1001)
surv_folds <- vfold_cv(survival2, v = 10)

system.time( { #127 seconds
  surv.models.best <- surv.models.best %>%
    mutate(resamples = map(wflow, fit_resamples, resamples = surv_folds, control = control_resamples(save_pred = TRUE)))
})
```

```{r}
surv.models.best %>% 
  mutate(metrics = map(resamples, collect_metrics, type = "wide")) %>%
  select(name, metrics) %>%
  unnest(metrics) %>%
  select(-.config) %>%
  arrange(desc(rsq))
```
cubic for both weeks and size is the best

```{r}
surv.models.best %>% 
  mutate(metrics = map(resamples, collect_metrics, type = "long")) %>%
  select(name, metrics) %>%
  unnest(metrics) %>%
  arrange(.metric, mean) %>%
  select(name, .metric, mean, std_err)

```

```{r}
surv.models.best %>% 
  mutate(metrics = map(resamples, collect_metrics, type = "long")) %>%
  select(name, metrics) %>%
  unnest(metrics) %>%
  arrange(.metric, mean) %>%
  select(name, .metric, mean, std_err) %>%
  ggplot(aes(x=reorder(name, mean), y = mean, ymin=mean-std_err, ymax=mean+std_err, fill = reorder(name, mean))) +
  geom_col() +
  geom_errorbar(width = 0.5) +
  facet_wrap(~.metric, scales = "free_y") +
  theme(axis.text.x = element_blank(), axis.title.x = element_blank()) +
  ggtitle("10-fold cross validation") +
  scale_fill_viridis_d()
```

```{r}
surv.model.final <- surv.models.best %>% slice_min(BIC) %>% pull(fit) %>% magrittr::extract2(1)
```

For plotting the prediction curve, it looks like I can't combine `geom_function` and `facet_wrap` so I will compute the prediction curves myself and then combine in the plots.

```{r}
newdata.survival <- survival2 %>%
  group_by(parent.pop, week) %>%
  reframe(height.cm = seq(min(height.cm, na.rm = TRUE),
                               max(height.cm, na.rm = TRUE),
                               length.out = 101),
          elapsed_weeks = 1) %>%
  filter(parent.pop != "WV")

surv.predictions <- surv.model.final %>% 
  extract_fit_engine() %>%
  predict(newdata.survival, type = "response", re.form = ~ (1 | parent.pop)) %>%
  cbind(newdata.survival, surv.next=.)

surv.predictions %>%
  ungroup() %>%
  nest(.by = parent.pop) %>% 
  mutate(plot = map2(data, parent.pop, \(x, pp) {
  ggplot(x, aes(x=height.cm, y = surv.next)) +
      geom_smooth(color = "grey50", lty=3, se = FALSE, method = "gam", data = {survival2 %>% filter(parent.pop==pp)}) + 
      geom_line() +
      facet_wrap(~week, scale = "free_x") +
      geom_point(alpha = 0.3, data = {survival2 %>% filter(parent.pop==pp)}) + 
      ggtitle(pp) +
      scale_color_brewer(type = "qual", palette = "Accent")
  })) %>%
  pull(plot) %>% walk(print)
```

```{r}
surv.model.final <- surv.model.final %>% extract_fit_engine()
growth.model.final <- growth.model.final %>% extract_fit_engine()
```

```{r}
save.image(file="../output/WL2_2023_growth_survival_models.Rdata")
```

Save just the pertinent stuff
```{r}
save(surv.model.final, growth.model.final, survival2, growth2, growth.rec.poly, surv.rec.poly, file = "../output/WL2_2023_growth_survival_models_FINAL.Rdata")
```



