---
title: "WL2 Fecundity"
author: "Julin Maloof"
date: "`r Sys.Date()`"
output: 
  html_document: 
    keep_md: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, message=FALSE}
library(tidyverse)
library(patchwork)
library(tidymodels)
library(multilevelmod)
library(lmerTest)
library(broom.mixed)
library(modelr)
conflicted::conflict_prefer("select", "dplyr")
conflicted::conflict_prefer("filter", "dplyr")
conflicted::conflict_prefer("lag", "dplyr")
```


```{r}
fruit <- read_csv("../input/WL2_2023_Fruits_forIPMs.csv") %>%
  filter(BedLoc!="K_5_C") %>% #get rid of duplicate locations
  filter(BedLoc!="B_32_A") %>% #get rid of duplicate locations
  filter(!is.na(pop), !str_detect(Genotype, ".*buff*"))  #remove buffers
fruit
```


```{r}
pheno.dates <- read_csv("../input/WL2_2023_PhenologyDates.csv")
pheno.dates
```

```{r}
growth <- read_csv("../input/WL2-2023_Size_Combined.csv")
growth
```



```{r}
# I don't necessarily need all of these new variables but just keeping the code from the G+S models
growth2 <- growth %>% arrange(Genotype, survey_date) %>%
  select(-survey.notes, -long.leaf.cm) %>%
  group_by(Genotype) %>%
  mutate(l.height.cm = log(height.cm),
         l.height.cm.next = lead(l.height.cm),
         elapsed_days= lead(survey_date) - survey_date,
         elapsed_weeks = as.integer(elapsed_days)/7) %>%
  ungroup() %>%
  mutate(week = as.numeric(survey_date - ymd("2023-07-03")) / 7, # 7/03 = pre-transplant survey date
         week_char = as.character(week)) %>% 
  drop_na(block, l.height.cm, l.height.cm.next)

growth2 
```

load survival
```{r}
survival <- read_csv("../input/WL2_Mortality.csv")

survival2 <- survival %>% select(Genotype, death.date) %>%
  right_join(growth) %>%
  mutate(l.height.cm = log(height.cm)) %>%
  arrange(Genotype, survey_date) %>%
  select(-survey.notes, -long.leaf.cm) %>%
  group_by(Genotype) %>%
  mutate(elapsed_days= lead(survey_date) - survey_date,
         elapsed_weeks = as.integer(elapsed_days)/7,
         block = ifelse(is.na(block), unique(na.omit(block)), block), # fill in missing block info 
         death.date = lubridate::mdy(death.date),
         surv.next = ifelse(is.na(death.date), 1, death.date > lead(survey_date))) %>% 
  ungroup() %>%
  mutate(week = as.numeric(survey_date - ymd("2023-07-03")) / 7, # 7/03 = pre-transplant survey date
         week_char = as.character(round(week, 2)),
         ppmf = str_c(parent.pop, "_", mf)) %>% # allows me to specify RE for pp and mf separately.
  drop_na(surv.next, parent.pop, l.height.cm, elapsed_days)
```

Here we are just going to predict fruit from size 2 months after transplant

```{r}
fruit_growth <- growth2 %>% filter(survey_date=="2023-09-06") %>%
  select(Genotype, height.cm, l.height.cm) %>%
  right_join(fruit, by="Genotype")
```

```{r, fig.asp=1}
p1 <- fruit_growth %>%
  ggplot(aes(x=height.cm, y=fruits)) +
  geom_point() +
  geom_smooth(se = FALSE) +
  geom_smooth(method="lm", color="red", se = FALSE) 

p2 <- fruit_growth %>%
  ggplot(aes(x=l.height.cm, y=fruits)) +
  geom_point() +
  geom_smooth(se = FALSE) +
  geom_smooth(method="lm", color="red", se = FALSE) 

p3 <- fruit_growth %>%
  ggplot(aes(x=height.cm, y=log(fruits))) +
  geom_point() +
  geom_smooth(se = FALSE) +
  geom_smooth(method="lm", color="red", se = FALSE) 

p4 <- fruit_growth %>%
  ggplot(aes(x=l.height.cm, y=log(fruits))) +
  geom_point() +
  geom_smooth(se = FALSE) +
  geom_smooth(method="lm", color="red", se = FALSE) 

(p1 + p2) / (p3 + p4)
```

## Some models

```{r}
fruit.spec <- poisson_reg() %>%
  set_engine("glmer", family = "poisson")

fruit.recipe <- fruit_growth %>%
  select(fruits, height.cm, l.height.cm, block) %>%
  recipe() %>%
  update_role(fruits, new_role = "outcome") %>% 
  update_role(c(height.cm, l.height.cm, block), new_role = "predictor") %>%
  step_poly(c(height.cm, l.height.cm), degree = 3, keep_original_cols = TRUE) %>%
  step_rename_at(contains("poly"), fn = \(x) str_replace(x, "_poly_", ".p"))


fruit.wflow <- workflow() %>%
  add_recipe(fruit.recipe)

```


```{r}
fruit.models <- tibble(wflow=list(
  
  ml1_linear = {fruit.wflow %>%
      add_model(fruit.spec,
                formula = fruits ~ height.cm + (1|block)) },
  
  ml2_linear_log.height = {fruit.wflow %>%
      add_model(fruit.spec,
                formula = fruits ~ l.height.cm + (1|block)) },
  
  ml3_quad = {fruit.wflow %>%
      add_model(fruit.spec,
                formula = fruits ~ height.cm.p1 + height.cm.p2 + (1|block)) },
  
  ml4_quad_log.height = {fruit.wflow %>%
      add_model(fruit.spec,
                formula = fruits ~ l.height.cm.p1 + l.height.cm.p2 + (1|block)) },
  
  ml5_cubic = {fruit.wflow %>%
      add_model(fruit.spec,
                formula = fruits ~ height.cm.p1 + height.cm.p2 + height.cm.p3 + (1|block)) },
  
  ml6_cubic_log.height = {fruit.wflow %>%
      add_model(fruit.spec,
                formula = fruits ~ l.height.cm.p1 + l.height.cm.p2 + l.height.cm.p3  + (1|block)) }
), 
name = names(wflow))
```

Fit
```{r}
#plan(multisession, workers = 7)

system.time( { # 110 seconds if using multisession; 
  fruit.models <- fruit.models %>%
    #mutate(fit = future_map(wflow, fit, data = fruit_growth,  .progress = TRUE, .options = furrr_options(packages = c("workflows", "multilevelmod"))),
    mutate(fit = map(wflow, fit, data = fruit_growth, .progress = TRUE),
           glance = map(fit, glance)
    ) 
})

fruit.models %>% select(-wflow, -fit) %>% unnest(glance) %>% arrange(BIC)
#plan(sequential)
```

Linear is the best...
```{r}
fruit.model.best <- fruit.models %>% unnest(glance) %>% slice_min(BIC, n=1) %>% pull(fit) %>% magrittr::extract2(1) %>% extract_fit_engine
```


Plot it

```{r, fig.asp=1}
fruit_growth %>%
  ggplot(aes(x=height.cm, y=fruits)) +
  geom_point() +
  geom_function(fun = \(x) predict(fruit.model.best, newdata=tibble(height.cm=x), type = "response", re.form=NA), color="red")
```

## Establishment and recruit size...

Also need something about probability of germination...

calculate survival for first time period
```{r}
tm2.params <- list()

tm2.params$establishment.prob <- survival2 %>% 
  filter(parent.pop=="TM2") %>% 
  distinct(Genotype, death.date) %>%
  summarize(establisment = 
              sum(death.date > "2023-07-26" | is.na(death.date)) / n() ) %>%
  pull(establisment)
  

tm2.params <- growth %>% filter(survey_date=="2023-07-03", parent.pop=="TM2") %>%
  summarize(recruit.size=mean(height.cm),
            recruit.size.sd=sd(height.cm)) %>%
  append(tm2.params, .)

tm2.params 
```

reproduction function:
```{r}
f.xy <- function(xp, x, tm2.params) {
  tm2.params$establishment.prob *
    dnorm(xp, mean = tm2.params$recruit.size, sd = tm2.params$recruit.size.sd) *
    predict(fruit.model.best, newdata=tibble(height.cm=x), re.form = NA, type = "response")
}
```

make a matrix

```{r}
minSize <- min(growth2$height.cm, na.rm = TRUE)*.95
maxSize <- max(growth2$height.cm, na.rm = TRUE)*1.05

nBigMatrix <- 100 # matrix dimensions
  b <- minSize + c(0:nBigMatrix) * (maxSize - minSize)/nBigMatrix # boundary points
  y <- 0.5 * (b[1:nBigMatrix] + b[2:(nBigMatrix + 1)]) # mesh points
  h <- y[2] - y[1] # step size
```



Functions to create the matrices
```{r}
F <-  h*outer(y, y, f.xy, tm2.params=tm2.params)
```

Function to plot matrices
```{r}
plotMatrix <- function(m, title=NA, y) {
m %>% as_tibble() %>%
  magrittr::set_colnames(y) %>% 
  mutate(size.t1=y) %>%
  pivot_longer(-size.t1, names_to = "size.t", names_transform = as.numeric) %>%
  ggplot(aes(x=size.t, y = size.t1)) +
  geom_raster(aes(fill = value)) +
  geom_contour(aes(z = value),lwd=.25) +
  geom_abline(intercept=0, slope = 1, color="gray90", lty=5) +
  scale_fill_viridis_c(option = "plasma") +
  labs(x = "Size (t)", y = "Size (t + 1)", title = title) +
  coord_equal() +
  theme_bw()
}
```

Create a tibble to hold the matrices from each pop and fill it
```{r}
plotMatrix(F, y=y)
```

